---
title: "de_analysis.Rmd"
output: html_document
date: "2026-04-09"
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(results = "hide")


```


```{r packages}
install.packages("BiocManager")

# Install the necessary bioinformatics packages
# TCGAbiolinks: for data access
# GSVA: for the 94-gene signature scoring
# SummarizedExperiment: the standard data container for TCGA data
BiocManager::install(c("TCGAbiolinks", "GSVA", "SummarizedExperiment"))
BiocManager::install("IlluminaHumanMethylation450kanno.ilmn12.hg19")
BiocManager::install("sesameData")
BiocManager::install("sesame")
BiocManager::install("sesame")
BiocManager::install("edgeR")
BiocManager::install("org.Hs.eg.db")
BiocManager::install("DESeq2")
BiocManager::install("apeglm", update = FALSE, ask = FALSE)

BiocManager::install(c("BiocGenerics", "S4Vectors", "IRanges"), force = TRUE, update = TRUE, ask = FALSE)
BiocManager::install("DESeq2", force=TRUE, update = FALSE, ask = FALSE)

install.packages("nloptr")
install.packages("ggpubr")
install.packages("cowplot")
install.packages("gt")
install.packages("minifi")

# 4/15/26 trying to get DESeq2
# 1. Remove the mismatched packages from your local Linux library
remove.packages(c("BiocGenerics", "S4Vectors", "IRanges"))

# 2. Force BiocManager to use the correct repository for R 4.3
BiocManager::install(version = "3.18", ask = FALSE)

# 3. Install DESeq2 cleanly from the 3.18 repository
BiocManager::install("DESeq2", update = FALSE, ask = FALSE)

BiocManager::install(c("BiocGenerics", "IRanges", "S4Vectors", "DESeq2"), update = FALSE, ask = FALSE)

# 1. Delete the corrupted phantom folder directly from the file system
unlink("/home/nvsharm/R/x86_64-pc-linux-gnu-library/4.3/S4Vectors", recursive = TRUE)

# 2. Install S4Vectors by pointing explicitly to the Bioconductor 3.18 repository
install.packages("S4Vectors", repos = "https://bioconductor.org/packages/3.18/bioc")

install.packages("IRanges", repos = "https://bioconductor.org/packages/3.18/bioc")

# 3. Test DESeq2
library(DESeq2)




# Install clusterProfiler and msigdbr
BiocManager::install("clusterProfiler")
install.packages("msigdbr")

```


```{r load packages}
library(BiocManager)
library(TCGAbiolinks)
library(GSVA)
library(SummarizedExperiment)
library(tidyverse)
library(nloptr)
library(ggpubr)
library(cowplot)
library(gt)
library(webshot2)
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
library(sesameData)
library(sesame)
library(edgeR)
library(org.Hs.eg.db)
library(AnnotationDbi)
library(S4Vectors)
library(IRanges)
library(DESeq2)
library(apeglm)
library(clusterProfiler)
library(msigdbr)
```

```{r read objects}

#candidate_sample_groups <- readRDS("candidate_sample_groups.rds")
#pol3_loci_long <- readRDS("pol3_loci_long.rds")
#pol3_gr <- readRDS("pol3_gr.rds")

```

# Extract raw counts for de analysis
Note: HNSC has the strongest biological signal but the smallest ATAC-seq sample size in this dataset.
```{r}


candidates <- c("BRCA", "COAD", "KIRP", "STAD", "KIRC", "BLCA")



prepare_rnaseq <- function(cancer_type, matched_samples) {
  message("Processing ", cancer_type)
  
  query <- GDCquery(
    project = paste0("TCGA-", cancer_type),
    data.category = "Transcriptome Profiling",
    data.type = "Gene Expression Quantification",
    workflow.type = "STAR - Counts",
    sample.type = "Primary Tumor"
  )
  
  se <- GDCprepare(query)
  
  # Filter to ATAC-seq matched samples only
  matched_barcodes <- matched_samples %>%
    filter(cancer_type == .env$cancer_type) %>%
    pull(case_id)
  
  #se_matched <- se[, se$barcode %in% matched_barcodes]
  se_matched <- se[, substr(colData(se)$barcode, 1, 12) %in% substr(matched_barcodes, 1, 12)]
  se_matched <- se_matched[, !duplicated(substr(colData(se_matched)$barcode, 1, 12))]
  
  message("  Matched ", ncol(se_matched), " of ", length(matched_barcodes), " samples")
  
  list(
    counts = assay(se_matched, "unstranded"),
    se = se_matched,
    meta = colData(se_matched) %>%
      as.data.frame() %>%
      mutate(case_id = substr(barcode, 1, 12)) %>%
      left_join(
        matched_samples %>%
          filter(cancer_type == .env$cancer_type) %>%
          dplyr::select(case_id, engager_group),
        by = "case_id"
      )
  )
}
```

```{r}

prepare_and_cache_rnaseq <- function(cancer_type, matched_samples, cache_dir = "tcga_cache") {
  
  # 1. Setup cache directory
  if(!dir.exists(cache_dir)) dir.create(cache_dir)
  cache_file <- file.path(cache_dir, paste0(cancer_type, "_rnaseq_matched.rds"))
  
  # 2. Check for cached data
  if(file.exists(cache_file)) {
    message("\n>>> Loading CACHED RNA-seq data for ", cancer_type)
    return(readRDS(cache_file))
  }
  
  # 3. If no cache, perform the download and processing
  message("\n>>> Downloading NEW RNA-seq data for ", cancer_type)
  
  query <- GDCquery(
    project = paste0("TCGA-", cancer_type),
    data.category = "Transcriptome Profiling",
    data.type = "Gene Expression Quantification",
    workflow.type = "STAR - Counts",
    sample.type = "Primary Tumor"
  )
  
  GDCdownload(query = query)
  se <- GDCprepare(query)
  
  # Filter to ATAC-seq matched samples
  matched_barcodes <- matched_samples %>%
    filter(cancer_type == .env$cancer_type) %>%
    pull(case_id)
  
  se_matched <- se[, substr(colData(se)$barcode, 1, 12) %in% substr(matched_barcodes, 1, 12)]
  se_matched <- se_matched[, !duplicated(substr(colData(se_matched)$barcode, 1, 12))]
  
  message("  Matched ", ncol(se_matched), " of ", length(matched_barcodes), " samples")
  
  # Compile the final object
  final_data <- list(
    counts = assay(se_matched, "unstranded"),
    se = se_matched,
    meta = colData(se_matched) %>%
      as.data.frame() %>%
      mutate(case_id = substr(barcode, 1, 12)) %>%
      left_join(
        matched_samples %>%
          filter(cancer_type == .env$cancer_type) %>%
          dplyr::select(case_id, engager_group),
        by = "case_id"
      )
  )
  
  # 4. Save to cache
  saveRDS(final_data, cache_file)
  message("  Saved to cache: ", cache_file)
  
  return(final_data)
}

```


# Functions for DE analysis
1) run edgeR engagers vs non_engagers
2) get up and down genes
3) Are any up genes in IFN signature genes?
4) Do up genes have engager loci within them?

Then we can test in BRCA 
## edgeR
```{r}
run_de_analysis <- function(cancer_type, matched_samples, pol3_loci_long, ifn_genes) {
  
  message("Running DE for ", cancer_type)
  
  # Prepare RNA-seq data
  data <- prepare_rnaseq(cancer_type, matched_samples)
  
  # 1. EdgeR DE engager vs non_engager ####
  dge <- DGEList(counts = data$counts, group = data$meta$engager_group)
  keep <- filterByExpr(dge)
  dge <- dge[keep, , keep.lib.sizes = FALSE]
  dge <- calcNormFactors(dge)
  
  design <- model.matrix(~ 0 + engager_group, data = data$meta)
  colnames(design) <- levels(factor(data$meta$engager_group))
  dge <- estimateDisp(dge, design)
  fit <- glmQLFit(dge, design)
  
  con <- makeContrasts(engager - non_engager, levels = design)
  qlf <- glmQLFTest(fit, contrast = con)
  
  results <- topTags(qlf, n = Inf)$table
  results$ensembl_id <- rownames(results) %>% str_remove("\\..*")
  results$gene_name <- mapIds(org.Hs.eg.db,
                              keys = results$ensembl_id,
                              column = "SYMBOL",
                              keytype = "ENSEMBL",
                              multiVals = "first")
  
  # 2. Upregulated and downregulated genes ####
  up_genes <- results %>%
    filter(FDR < 0.05, !is.na(gene_name), logFC > 0) %>%
    pull(gene_name)
  
  down_genes <- results %>%
    filter(FDR < 0.05, !is.na(gene_name), logFC < 0) %>%
    pull(gene_name)
  
  # 3. IFN overlap ####
  ifn_overlap <- intersect(up_genes, ifn_genes)
  
  # 4. Engager loci overlap ####
  engager_loci <- pol3_loci_long %>%
    filter(cancer_type == .env$cancer_type,
           is_5ppp == 1,
           accessibility == 1) %>%
    distinct(V1, V2, V3)
  
  engager_gr <- GRanges(
    seqnames = engager_loci$V1,
    ranges = IRanges(start = as.integer(engager_loci$V2),
                     end = as.integer(engager_loci$V3))
  )
  
  up_gr <- rowRanges(data$se)[rowData(data$se)$gene_name %in% up_genes]
  overlaps <- findOverlaps(up_gr, engager_gr)
  genes_with_engager_loci <- names(up_gr)[queryHits(overlaps)]
  
  list(
    de_results = results,
    up_genes = up_genes,
    down_genes = down_genes,
    ifn_overlap = ifn_overlap,
    genes_with_engager_loci = genes_with_engager_loci,
    n_samples = nrow(data$meta),
    n_per_group = table(data$meta$engager_group)
  )
}
```

### For one cancer type
```{r}
brca_de <- run_de_analysis("BRCA", candidate_sample_groups, pol3_loci_long, ifn_genes)

# How many DE genes
length(brca_de$up_genes) #11
length(brca_de$down_genes) #20

# Any IFN overlap
brca_de$ifn_overlap

# Any genes with engager loci
brca_de$genes_with_engager_loci

# Sample counts
brca_de$n_per_group

# Are there any IFN genes that are even in the expression matrix and were filtered out?
intersect(ifn_genes, rownames(brca_de$de_results %>% filter(!is.na(gene_name)))) # 0 in expression matrix


```

### For all cancer types
```{r}
candidates <- c("COAD", "KIRP", "STAD", "KIRC", "BLCA") #"BRCA", 

de_results_all <- lapply(candidates, run_de_analysis,
                         matched_samples = candidate_sample_groups,
                         pol3_loci_long = pol3_loci_long,
                         ifn_genes = ifn_genes)
names(de_results_all) <- candidates

de_results_all[["BRCA"]] <- brca_de

# Rename in active environment b/c doing DESeq2 also
#de_results_all_edgeR <- de_results_all

# Save object to disk
#saveRDS(de_results_all_edgeR, "de_results_all_edgeR.rds")

```

### Dig into DE results for all cancer types
```{r}
lapply(names(de_results_all), function(ct) {
  cat("\n", ct, "\n")
  cat("N per group:", de_results_all[[ct]]$n_per_group, "\n")
  cat("N up genes FDR<0.05:", length(de_results_all[[ct]]$up_genes), "\n")
  cat("N down genes FDR<0.05:", length(de_results_all[[ct]]$down_genes), "\n")
  cat("IFN overlap:", paste(de_results_all[[ct]]$ifn_overlap, collapse=", "), "\n")
  cat("Genes with engager loci:", paste(de_results_all[[ct]]$genes_with_engager_loci, collapse=", "), "\n")
})


```

## DESeq2
### Module 1 Core DESeq2 Function
```{r}

# Function handles matrix setup, variance stabilization, and DE testing 
# returns a list containing the raw dds object, the full results df, and intersection logic
run_deseq2_core <- function(cancer_type, matched_samples, pol3_loci_long, ifn_genes) {
  
  message("\n>>> Running DESeq2 Core for ", cancer_type)
  data <- prepare_and_cache_rnaseq(cancer_type, matched_samples)
  
  # Ensure reference level
  data$meta$engager_group <- factor(data$meta$engager_group, levels = c("non_engager", "engager"))
  
  dds <- DESeqDataSetFromMatrix(countData = data$counts,
                                colData = data$meta,
                                design = ~ engager_group)
  
  keep <- rowSums(counts(dds)) >= 10
  dds <- dds[keep,]
  dds <- DESeq(dds)
  res <- lfcShrink(dds, coef="engager_group_engager_vs_non_engager", type="apeglm")
  
  res_df <- as.data.frame(res)
  res_df$ensembl_id <- rownames(res_df) %>% str_remove("\\..*")
  res_df$gene_name <- mapIds(org.Hs.eg.db,
                             keys = res_df$ensembl_id,
                             column = "SYMBOL",
                             keytype = "ENSEMBL",
                             multiVals = "first")
  
  up_genes <- res_df %>% filter(padj < 0.05, !is.na(gene_name), log2FoldChange > 0.5) %>% pull(gene_name)
  down_genes <- res_df %>% filter(padj < 0.05, !is.na(gene_name), log2FoldChange < -0.5) %>% pull(gene_name)
  
  # Engager loci overlap logic
  engager_loci <- pol3_loci_long %>%
    filter(cancer_type == .env$cancer_type, is_5ppp == 1, accessibility == 1) %>%
    distinct(V1, V2, V3)
  
  engager_gr <- GRanges(seqnames = engager_loci$V1, ranges = IRanges(start = as.integer(engager_loci$V2), end = as.integer(engager_loci$V3)))
  up_gr <- rowRanges(data$se)[rowData(data$se)$gene_name %in% up_genes]
  overlaps <- findOverlaps(up_gr, engager_gr)
  genes_with_engager_loci <- names(up_gr)[queryHits(overlaps)]
  
  return(list(
    cancer_type = cancer_type,
    dds = dds,
    res_df = res_df,
    up_genes = up_genes,
    down_genes = down_genes,
    ifn_overlap = intersect(up_genes, ifn_genes),
    genes_with_engager_loci = genes_with_engager_loci,
    n_samples = nrow(data$meta),
    n_per_group = table(data$meta$engager_group)
  ))
}

```




### Module 2 QC Plotting
```{r}
generate_deseq_qc_plots <- function(deseq_output) {
  ct <- deseq_output$cancer_type
  dds <- deseq_output$dds
  
  dir.create("DESeq_QC_Plots", showWarnings = FALSE)
  
  # Dispersion Plot
  png(paste0("DESeq_QC_Plots/", ct, "_dispersion.png"), width=600, height=500)
  plotDispEsts(dds, main=paste(ct, "Dispersion Estimates"))
  dev.off()
  
  # PCA Plot
  vsd <- vst(dds, blind=FALSE)
  pca_data <- plotPCA(vsd, intgroup="engager_group", returnData=TRUE)
  percentVar <- round(100 * attr(pca_data, "percentVar"))
  
  pca_plot <- ggplot(pca_data, aes(PC1, PC2, color=engager_group)) +
    geom_point(size=3, alpha=0.8) +
    theme_minimal() +
    labs(title=paste(ct, "PCA"),
         x=paste0("PC1: ", percentVar[1], "%"),
         y=paste0("PC2: ", percentVar[2], "%")) +
    scale_color_manual(values=c("non_engager"="#00BFC4", "engager"="#F8766D"))
  
  ggsave(paste0("DESeq_QC_Plots/", ct, "_PCA.png"), plot=pca_plot, width=5, height=4)
  message("Saved QC plots for ", ct)
}
```


### Module 3 GSVA sample-by-sample scoring
```{r}
calculate_sample_scores <- function(deseq_output, custom_index = NULL) {
  
  # Default to the cancer-specific up_genes if no index is provided
  if(is.null(custom_index)) {
    custom_index <- deseq_output$up_genes
  }
  
  if(length(custom_index) == 0) {
    message("No index genes available for ", deseq_output$cancer_type)
    return(NULL)
  }
  
  # Extract VST normalized counts
  vsd <- vst(deseq_output$dds, blind=FALSE)
  vst_mat <- assay(vsd)
  
  # Map rownames from Ensembl to Symbol for GSVA matching
  rownames(vst_mat) <- deseq_output$res_df$gene_name[match(rownames(vst_mat), rownames(deseq_output$res_df))]
  vst_mat <- vst_mat[!is.na(rownames(vst_mat)), ] # Drop any genes that didn't map correctly to a symbol (NAs)
  
  # Put the up genes (for brca = 133 up genes) into a list format for GSVA package
  index_list <- list(Target_Index = custom_index)
  
  # run GSVA to get one score per column (per sample)
  gsva_res <- gsva(expr = vst_mat, gset.idx.list = index_list, method = "gsva", verbose = FALSE)
  
  # format into a table that can be used to make a boxplot
  sample_scores <- data.frame(
    case_id = colnames(vst_mat),
    custom_metric = as.numeric(gsva_res["Target_Index", ]),
    group = deseq_output$dds$engager_group
  )
  
  return(sample_scores)
}
```

### BRCA 
```{r}
#### Module 1_____ ####
brca_deseq2_test <- run_deseq2_core("BRCA", candidate_sample_groups, pol3_loci_long, ifn_genes)
brca_deseq2_test$n_per_group
# non_engager     engager 
#          15          14
length(brca_deseq2_test$up_genes)
# [1] 169
brca_deseq2_test$res_df %>% View


#### Module 2______ ####
# Generate the PCA and Dispersion plots
generate_deseq_qc_plots(brca_deseq2_test)


#### Module 3______ ####
# Generate the custom sample-by-sample metric
brca_gsva_scores <- calculate_sample_scores(brca_deseq2_test)

# Visualize the custom 133 gene score
group_colors <- c("non_engager" = "#4575b4", "engager" = "#d73027")

ggplot(brca_gsva_scores, aes(x = group, y = custom_metric, fill = group)) +
  # Create the boxplot with a bit of transparency
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  # Add the sample points (dots) and color them by group
  geom_jitter(aes(color = group), width = 0.2, size = 2.5, alpha = 0.8) +
  # Use a clean, professional theme
  theme_minimal() +
  # Map the colors we defined
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  # Refine labels and titles
  labs(
    title = "BRCA: Sample-Level Gene Set Scoring",
    subtitle = paste0("Based on ", length(brca_deseq2_test$up_genes), "-gene Signature"),
    y = "GSVA Enrichment Score",
    x = NULL # Removing 'group' label as the axis categories make it obvious
  ) +
  # Remove the legend since the x-axis already labels the groups
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 11, color = "black"),
    panel.grid.minor = element_blank()
  ) +
  stat_compare_means(method = "wilcox.test", label.x = 1.5, label.y = 0.6)




ggplot(brca_gsva_scores, aes(x = group, y = custom_metric, fill = group)) +
  geom_boxplot() +
  geom_jitter(width = 0.2) +
  theme_minimal() +
  labs(title = "BRCA: Sample-by-Sample Custom Metric (133-Gene Index)", y = "GSVA Score")


#### biological proof: enrichment of immunostimulatory pathways______ ####
# 2. Pull the Hallmark Interferon Alpha list
# ifn_alpha <- msigdbr(species = "human", category = "H") %>% 
#   filter(gs_name == "HALLMARK_INTERFERON_ALPHA_RESPONSE") %>% 
#   pull(gene_symbol)

# 2. See which of 33 genes overlap with ifn_genes
brca_deseq2_ifn_overlap <- intersect(brca_deseq2_test$up_genes, ifn_genes)

# 3. Check the results
print(paste("Number of overlapping IFN genes:", length(brca_deseq2_ifn_overlap)))
print(brca_deseq2_ifn_overlap)

# Check overlap with edgeR genes
intersect(brca_deseq2_test$up_genes, brca_de$up_genes)


#### MSigDB interrogation of up genes______####

# 1. Pull the official Hallmark immune sets
msigdb_hallmark_sets <- msigdbr(species = "human", category = "H")
ifn_alpha_list <- msigdb_hallmark_sets[msigdb_hallmark_sets$gs_name == "HALLMARK_INTERFERON_ALPHA_RESPONSE", ]$gene_symbol
ifn_gamma_list <- msigdb_hallmark_sets[msigdb_hallmark_sets$gs_name == "HALLMARK_INTERFERON_GAMMA_RESPONSE", ]$gene_symbol
inflam_list <- msigdb_hallmark_sets[msigdb_hallmark_sets$gs_name == "HALLMARK_INFLAMMATORY_RESPONSE", ]$gene_symbol

# Create a list of these specific pathways
target_pathways <- list(
  ifn_alpha = ifn_alpha_list,
  ifn_gamma = ifn_gamma_list,
  inflammatory = inflam_list
)

# 2. Extract VST normalized counts and convert Ensembl IDs to Gene Symbols
vst_brca <- vst(brca_deseq2_test$dds, blind=FALSE)
brca_vst_mat <- assay(vst_brca)
rownames(brca_vst_mat) <- brca_deseq2_test$res_df$gene_name[match(rownames(brca_vst_mat), rownames(brca_deseq2_test$res_df))]
brca_vst_mat <- brca_vst_mat[!is.na(rownames(brca_vst_mat)), ]

# 3. Score the BRCA samples against ONLY these immune pathways
brca_immune_gsva <- gsva(expr = brca_vst_mat, 
                         gset.idx.list = target_pathways, 
                         method = "gsva", 
                         verbose = FALSE)


# 4. Format all three pathways into a dataframe
brca_immune_scores <- data.frame(
  case_id = colnames(brca_vst_mat),
  ifn_alpha = as.numeric(brca_immune_gsva["ifn_alpha", ]),
  ifn_gamma = as.numeric(brca_immune_gsva["ifn_gamma", ]), 
  inflammatory = as.numeric(brca_immune_gsva["inflammatory", ]),
  group = brca_deseq2_test$dds$engager_group
)

# Reshape data for ggplot faceting
brca_immune_long <- brca_immune_scores %>%
  pivot_longer(cols = c(ifn_alpha, ifn_gamma, inflammatory),
               names_to = "pathway",
               values_to = "score")

# 5. Plot the result for IFN Alpha
# 5. Plot all three pathways side-by-side
ggplot(brca_immune_long, aes(x = group, y = score, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  geom_jitter(aes(color = group), width = 0.2, size = 2, alpha = 0.8) +
  facet_wrap(~ pathway) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(
    title = "BRCA: Targeted Immune Pathway GSVA Scores", 
    y = "GSVA Enrichment Score", 
    x = NULL
  ) +
  theme(
    legend.position = "none", 
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 11, color = "black"),
    strip.text = element_text(size = 12, face = "bold"),
    panel.grid.minor = element_blank()
  ) +
  stat_compare_means(method = "wilcox.test", label = "p.format", label.x = 1.3)
```


### COAD 
```{r}
#### Module 1 ####
coad_deseq2_test <- run_deseq2_core("COAD", candidate_sample_groups, pol3_loci_long, ifn_genes)
coad_deseq2_test$n_per_group
# non_engager     engager 
#          9          8
length(coad_deseq2_test$up_genes)
# [1] 26
coad_deseq2_test$res_df %>% View


#### Module 2 ####
# Generate the PCA and Dispersion plots
generate_deseq_qc_plots(coad_deseq2_test)


#### Module 3 ####
# Generate the custom sample-by-sample metric
coad_gsva_scores <- calculate_sample_scores(coad_deseq2_test)


# Visualize the custom 133 gene score
group_colors <- c("non_engager" = "#4575b4", "engager" = "#d73027")

ggplot(coad_gsva_scores, aes(x = group, y = custom_metric, fill = group)) +
  # Create the boxplot with a bit of transparency
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  # Add the sample points (dots) and color them by group
  geom_jitter(aes(color = group), width = 0.2, size = 2.5, alpha = 0.8) +
  # Use a clean, professional theme
  theme_minimal() +
  # Map the colors we defined
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  # Refine labels and titles
  labs(
    title = "COAD: Sample-Level Gene Set Scoring",
    subtitle = paste0("Based on ", length(brca_deseq2_test$up_genes), "-gene Signature"),
    y = "GSVA Enrichment Score",
    x = NULL # Removing 'group' label as the axis categories make it obvious
  ) +
  # Remove the legend since the x-axis already labels the groups
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 11, color = "black"),
    panel.grid.minor = element_blank()
  ) +
  stat_compare_means(method = "wilcox.test", label.x = 1.5, label.y = 0.6)

#### MSigDB interrogation of immune pathways______####

# # 1. Pull the official Hallmark immune sets (re-defining to ensure safe exact-match names)
# msigdb_hallmark_sets <- msigdbr(species = "human", category = "H")
# ifn_alpha_list <- msigdb_hallmark_sets[msigdb_hallmark_sets$gs_name == "HALLMARK_INTERFERON_ALPHA_RESPONSE", ]$gene_symbol
# ifn_gamma_list <- msigdb_hallmark_sets[msigdb_hallmark_sets$gs_name == "HALLMARK_INTERFERON_GAMMA_RESPONSE", ]$gene_symbol
# inflam_list <- msigdb_hallmark_sets[msigdb_hallmark_sets$gs_name == "HALLMARK_INFLAMMATORY_RESPONSE", ]$gene_symbol
# 
# target_pathways <- list(
#   ifn_alpha = ifn_alpha_list,
#   ifn_Gamma = ifn_gamma_list,
#   inflammatory = inflam_list
# )

# 2. Extract VST normalized counts and convert Ensembl IDs to Gene Symbols
vst_coad <- vst(coad_deseq2_test$dds, blind=FALSE)
coad_vst_mat <- assay(vst_coad)
rownames(coad_vst_mat) <- coad_deseq2_test$res_df$gene_name[match(rownames(coad_vst_mat), rownames(coad_deseq2_test$res_df))]
coad_vst_mat <- coad_vst_mat[!is.na(rownames(coad_vst_mat)), ]

# 3. Score the COAD samples against ONLY these immune pathways
coad_immune_gsva <- gsva(expr = coad_vst_mat, 
                         gset.idx.list = target_pathways, 
                         method = "gsva", 
                         verbose = FALSE)

# 4. Format all three pathways into a dataframe
coad_immune_scores <- data.frame(
  case_id = colnames(coad_vst_mat),
  ifn_alpha = as.numeric(coad_immune_gsva["ifn_alpha", ]),
  ifn_gamma = as.numeric(coad_immune_gsva["ifn_gamma", ]),
  inflammatory = as.numeric(coad_immune_gsva["inflammatory", ]),
  group = coad_deseq2_test$dds$engager_group
)

# Reshape data for ggplot faceting
coad_immune_long <- coad_immune_scores %>%
  pivot_longer(cols = c(ifn_alpha, ifn_gamma, inflammatory),
               names_to = "pathway",
               values_to = "score")

# 5. Plot all three pathways side-by-side with a shared y-axis
ggplot(coad_immune_long, aes(x = group, y = score, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  geom_jitter(aes(color = group), width = 0.2, size = 2, alpha = 0.8) +
  facet_wrap(~ pathway) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(
    title = "COAD: Targeted Immune Pathway GSVA Scores", 
    y = "GSVA Enrichment Score", 
    x = NULL
  ) +
  theme(
    legend.position = "none", 
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 11, color = "black"),
    strip.text = element_text(size = 12, face = "bold"),
    panel.grid.minor = element_blank()
  ) +
  stat_compare_means(method = "wilcox.test", label = "p.format", label.x = 1.3)


#### Is there overlap of up genes with other cancertypes?____####
# 2. See which of 33 genes overlap with ifn_genes
coad_deseq2_ifn_overlap <- intersect(coad_deseq2_test$up_genes, ifn_genes)

# 3. Check the results
print(paste("Number of overlapping IFN genes:", length(coad_deseq2_ifn_overlap)))
print(coad_deseq2_ifn_overlap)

# Check overlap with brca up genes  
intersect(coad_deseq2_test$up_genes, brca_deseq2_test$up_genes)




```


#### Check if the 11 genes identified by edgeR in BRCA as upgenes (core genes) and score
```{r}

# Will see if we can use these 11 genes as a pan-cancer since the edgeR GO term enrichment in EnrichR showed Rig-i etc. and many of these genes were inflammatory/immunostimulatory
# Question: Why did edgeR or how did it narrow it down to these genes while DESeq introduced other genes that separated non_engagers from engagers but encapsulated different oncogenic processes?

# 1. Define the core 11-gene index from the BRCA edgeR/DESeq2 overlap
core_11_genes <- c("ABCC2", "EPN3", "LLGL2", "PSMD3", "LRRC59", 
                   "RAP2C", "PCDH10", "CXCR1", "CPA6", "PHB1", "ZNF652")

# 2. Score BRCA using ONLY the 11 core genes
brca_core_scores <- calculate_sample_scores(brca_deseq2_test, custom_index = core_11_genes)

# Plot BRCA Core Scores
p_brca_core <- ggplot(brca_core_scores, aes(x = group, y = custom_metric, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  geom_jitter(aes(color = group), width = 0.2, size = 2.5, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(
    title = "BRCA: 11-Gene Core Signature", 
    y = "GSVA Enrichment Score", 
    x = NULL
  ) +
  theme(legend.position = "none", plot.title = element_text(face = "bold")) +
  stat_compare_means(method = "wilcox.test", label.x = 1.5)

# 3. Score COAD using ONLY the 11 core genes
coad_core_scores <- calculate_sample_scores(coad_deseq2_test, custom_index = core_11_genes)

# Plot COAD Core Scores
p_coad_core <- ggplot(coad_core_scores, aes(x = group, y = custom_metric, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  geom_jitter(aes(color = group), width = 0.2, size = 2.5, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(
    title = "COAD: 11-Gene Core Signature", 
    y = "GSVA Enrichment Score", 
    x = NULL
  ) +
  theme(legend.position = "none", plot.title = element_text(face = "bold")) +
  stat_compare_means(method = "wilcox.test", label.x = 1.5)

# 4. Display side-by-side
plot_grid(p_brca_core, p_coad_core, ncol = 2)

```

### KIRP 
```{r}
#### Module 1 ####
kirp_deseq2_test <- run_deseq2_core("KIRP", candidate_sample_groups, pol3_loci_long, ifn_genes)
kirp_deseq2_test$n_per_group
# non_engager     engager 
#           9           8 

# Check how many genes passed the filter
length(kirp_deseq2_test$up_genes)
# 12

#### Module 2 ####
# Generate the PCA and Dispersion plots
generate_deseq_qc_plots(kirp_deseq2_test)


#### Module 3 ####
# Generate the custom sample-by-sample metric
kirp_gsva_scores <- calculate_sample_scores(kirp_deseq2_test)

# Visualize the custom metric
ggplot(kirp_gsva_scores, aes(x = group, y = custom_metric, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  geom_jitter(aes(color = group), width = 0.2, size = 2.5, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(
    title = "KIRP: Sample-Level Gene Set Scoring",
    subtitle = paste0("Based on ", length(kirp_deseq2_test$up_genes), "-gene Signature"),
    y = "GSVA Enrichment Score",
    x = NULL 
  ) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 11, color = "black"),
    panel.grid.minor = element_blank()
  ) +
  stat_compare_means(method = "wilcox.test", label.x = 1.5, label.y = 0.6)


#### MSigDB interrogation of immune pathways______####
# 1. Extract VST normalized counts and convert Ensembl IDs to Gene Symbols
vst_kirp <- vst(kirp_deseq2_test$dds, blind=FALSE)
kirp_vst_mat <- assay(vst_kirp)
rownames(kirp_vst_mat) <- kirp_deseq2_test$res_df$gene_name[match(rownames(kirp_vst_mat), rownames(kirp_deseq2_test$res_df))]
kirp_vst_mat <- kirp_vst_mat[!is.na(rownames(kirp_vst_mat)), ]

# 2. Score the KIRP samples against ONLY these immune pathways
# (Relies on target_pathways list defined in the BRCA chunk)
kirp_immune_gsva <- gsva(expr = kirp_vst_mat, 
                         gset.idx.list = target_pathways, 
                         method = "gsva", 
                         verbose = FALSE)

# 3. Format all three pathways into a dataframe
kirp_immune_scores <- data.frame(
  case_id = colnames(kirp_vst_mat),
  ifn_alpha = as.numeric(kirp_immune_gsva["ifn_alpha", ]),
  ifn_gamma = as.numeric(kirp_immune_gsva["ifn_gamma", ]),
  inflammatory = as.numeric(kirp_immune_gsva["inflammatory", ]),
  group = kirp_deseq2_test$dds$engager_group
)

# Reshape data for ggplot faceting
kirp_immune_long <- kirp_immune_scores %>%
  pivot_longer(cols = c(ifn_alpha, ifn_gamma, inflammatory),
               names_to = "pathway",
               values_to = "score")

# 4. Plot all three pathways side-by-side with a shared y-axis
ggplot(kirp_immune_long, aes(x = group, y = score, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  geom_jitter(aes(color = group), width = 0.2, size = 2, alpha = 0.8) +
  facet_wrap(~ pathway) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(
    title = "KIRP: Targeted Immune Pathway GSVA Scores", 
    y = "GSVA Enrichment Score", 
    x = NULL
  ) +
  theme(
    legend.position = "none", 
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 11, color = "black"),
    strip.text = element_text(size = 12, face = "bold"),
    panel.grid.minor = element_blank()
  ) +
  stat_compare_means(method = "wilcox.test", label = "p.format", label.x = 1.3)


#### Is there overlap of up genes with other cancertypes?____####
kirp_deseq2_ifn_overlap <- intersect(kirp_deseq2_test$up_genes, ifn_genes)

print(paste("Number of overlapping IFN genes:", length(kirp_deseq2_ifn_overlap)))
print(kirp_deseq2_ifn_overlap)

# Check overlap with brca up genes  
intersect(kirp_deseq2_test$up_genes, brca_deseq2_test$up_genes)


#### Check if the 11 genes identified by edgeR in BRCA score high here ____####
# (Relies on core_11_genes vector defined in previous chunks)
kirp_core_scores <- calculate_sample_scores(kirp_deseq2_test, custom_index = core_11_genes)

ggplot(kirp_core_scores, aes(x = group, y = custom_metric, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  geom_jitter(aes(color = group), width = 0.2, size = 2.5, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(
    title = "KIRP: 11-Gene Core Signature", 
    y = "GSVA Enrichment Score", 
    x = NULL
  ) +
  theme(legend.position = "none", plot.title = element_text(face = "bold")) +
  stat_compare_means(method = "wilcox.test", label.x = 1.5)
```


### STAD
```{r}
#### Module 1 ####
stad_deseq2_test <- run_deseq2_core("STAD", candidate_sample_groups, pol3_loci_long, ifn_genes)
stad_deseq2_test$n_per_group
# non_engager     engager 
#           4           4 

# Check how many genes passed the filter
length(stad_deseq2_test$up_genes)
# 383

#### Module 2 ####
# Generate the PCA and Dispersion plots
generate_deseq_qc_plots(stad_deseq2_test)


#### Module 3 ####
# Generate the custom sample-by-sample metric
stad_gsva_scores <- calculate_sample_scores(stad_deseq2_test)

# Visualize the custom metric
ggplot(stad_gsva_scores, aes(x = group, y = custom_metric, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  geom_jitter(aes(color = group), width = 0.2, size = 2.5, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(
    title = "STAD: Sample-Level Gene Set Scoring",
    subtitle = paste0("Based on ", length(stad_deseq2_test$up_genes), "-gene Signature"),
    y = "GSVA Enrichment Score",
    x = NULL 
  ) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 11, color = "black"),
    panel.grid.minor = element_blank()
  ) +
  stat_compare_means(method = "wilcox.test", label.x = 1.5, label.y = 0.6)


#### MSigDB interrogation of immune pathways______####
# 1. Extract VST normalized counts and convert Ensembl IDs to Gene Symbols
vst_stad <- vst(stad_deseq2_test$dds, blind=FALSE)
stad_vst_mat <- assay(vst_stad)
rownames(stad_vst_mat) <- stad_deseq2_test$res_df$gene_name[match(rownames(stad_vst_mat), rownames(stad_deseq2_test$res_df))]
stad_vst_mat <- stad_vst_mat[!is.na(rownames(stad_vst_mat)), ]

# 2. Score the STAD samples against ONLY these immune pathways
# (Relies on target_pathways list defined in the BRCA chunk)
stad_immune_gsva <- gsva(expr = stad_vst_mat, 
                         gset.idx.list = target_pathways, 
                         method = "gsva", 
                         verbose = FALSE)

# 3. Format all three pathways into a dataframe
stad_immune_scores <- data.frame(
  case_id = colnames(stad_vst_mat),
  ifn_alpha = as.numeric(stad_immune_gsva["ifn_alpha", ]),
  ifn_gamma = as.numeric(stad_immune_gsva["ifn_gamma", ]),
  inflammatory = as.numeric(stad_immune_gsva["inflammatory", ]),
  group = stad_deseq2_test$dds$engager_group
)

# Reshape data for ggplot faceting
stad_immune_long <- stad_immune_scores %>%
  pivot_longer(cols = c(ifn_alpha, ifn_gamma, inflammatory),
               names_to = "pathway",
               values_to = "score")

# 4. Plot all three pathways side-by-side with a shared y-axis
ggplot(stad_immune_long, aes(x = group, y = score, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  geom_jitter(aes(color = group), width = 0.2, size = 2, alpha = 0.8) +
  facet_wrap(~ pathway) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(
    title = "STAD: Targeted Immune Pathway GSVA Scores", 
    y = "GSVA Enrichment Score", 
    x = NULL
  ) +
  theme(
    legend.position = "none", 
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 11, color = "black"),
    strip.text = element_text(size = 12, face = "bold"),
    panel.grid.minor = element_blank()
  ) +
  stat_compare_means(method = "wilcox.test", label = "p.format", label.x = 1.3)


#### Is there overlap of up genes with other cancertypes?____####
stad_deseq2_ifn_overlap <- intersect(stad_deseq2_test$up_genes, ifn_genes)

print(paste("Number of overlapping IFN genes:", length(stad_deseq2_ifn_overlap)))
print(stad_deseq2_ifn_overlap)

# Check overlap with brca up genes  
intersect(stad_deseq2_test$up_genes, brca_deseq2_test$up_genes)
# [1] "CSMD3"    "B4GALNT2" "UGT3A2" 

#### Check if the 11 genes identified by edgeR in BRCA score high here ____####
# (Relies on core_11_genes vector defined in previous chunks)
stad_core_scores <- calculate_sample_scores(stad_deseq2_test, custom_index = core_11_genes)

ggplot(stad_core_scores, aes(x = group, y = custom_metric, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  geom_jitter(aes(color = group), width = 0.2, size = 2.5, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(
    title = "STAD: 11-Gene Core Signature", 
    y = "GSVA Enrichment Score", 
    x = NULL
  ) +
  theme(legend.position = "none", plot.title = element_text(face = "bold")) +
  stat_compare_means(method = "wilcox.test", label.x = 1.5)
```

KIRC
```{r}
#### Module 1 ####
kirc_deseq2_test <- run_deseq2_core("KIRC", candidate_sample_groups, pol3_loci_long, ifn_genes)
kirc_deseq2_test$n_per_group
# non_engager     engager 
#           3           2 
# Check how many genes passed the filter
length(kirc_deseq2_test$up_genes)
# 425

#### Module 2 ####
generate_deseq_qc_plots(kirc_deseq2_test)

#### Module 3 ####
kirc_gsva_scores <- calculate_sample_scores(kirc_deseq2_test)

ggplot(kirc_gsva_scores, aes(x = group, y = custom_metric, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  geom_jitter(aes(color = group), width = 0.2, size = 2.5, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(title = "KIRC: Sample-Level Gene Set Scoring",
       subtitle = paste0("Based on ", length(kirc_deseq2_test$up_genes), "-gene Signature"),
       y = "GSVA Enrichment Score", x = NULL) +
  theme(legend.position = "none", plot.title = element_text(face = "bold")) +
  stat_compare_means(method = "wilcox.test", label.x = 1.5)

#### MSigDB interrogation ####
vst_kirc <- vst(kirc_deseq2_test$dds, blind=FALSE)
kirc_vst_mat <- assay(vst_kirc)
rownames(kirc_vst_mat) <- kirc_deseq2_test$res_df$gene_name[match(rownames(kirc_vst_mat), rownames(kirc_deseq2_test$res_df))]
kirc_vst_mat <- kirc_vst_mat[!is.na(rownames(kirc_vst_mat)), ]

kirc_immune_gsva <- gsva(expr = kirc_vst_mat, gset.idx.list = target_pathways, method = "gsva", verbose = FALSE)

kirc_immune_scores <- data.frame(
  case_id = colnames(kirc_vst_mat),
  ifn_alpha = as.numeric(kirc_immune_gsva["ifn_alpha", ]),
  ifn_gamma = as.numeric(kirc_immune_gsva["ifn_gamma", ]),
  inflammatory = as.numeric(kirc_immune_gsva["inflammatory", ]),
  group = kirc_deseq2_test$dds$engager_group
)

kirc_immune_long <- kirc_immune_scores %>%
  pivot_longer(cols = c(ifn_alpha, ifn_gamma, inflammatory), names_to = "pathway", values_to = "score")

ggplot(kirc_immune_long, aes(x = group, y = score, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  geom_jitter(aes(color = group), width = 0.2, size = 2, alpha = 0.8) +
  facet_wrap(~ pathway) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(title = "KIRC: Targeted Immune Pathway GSVA Scores", y = "GSVA Enrichment Score", x = NULL) +
  theme(legend.position = "none", plot.title = element_text(face = "bold")) +
  stat_compare_means(method = "wilcox.test", label = "p.format", label.x = 1.3)

#### Overlap and Core Signature ####
kirc_deseq2_ifn_overlap <- intersect(kirc_deseq2_test$up_genes, ifn_genes)
print(paste("Number of overlapping IFN genes:", length(kirc_deseq2_ifn_overlap)))
intersect(kirc_deseq2_test$up_genes, brca_deseq2_test$up_genes)
# [1] "DHRS2" "WNT2"  "PADI3"

kirc_core_scores <- calculate_sample_scores(kirc_deseq2_test, custom_index = core_11_genes)
ggplot(kirc_core_scores, aes(x = group, y = custom_metric, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  geom_jitter(aes(color = group), width = 0.2, size = 2.5, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(title = "KIRC: 11-Gene Core Signature", y = "GSVA Enrichment Score", x = NULL) +
  theme(legend.position = "none", plot.title = element_text(face = "bold")) +
  stat_compare_means(method = "wilcox.test", label.x = 1.5)
```

#### BLCA
```{r}
#### BLCA 
#### Module 1 ####
blca_deseq2_test <- run_deseq2_core("BLCA", candidate_sample_groups, pol3_loci_long, ifn_genes)
print("Sample Sizes for BLCA:")
print(blca_deseq2_test$n_per_group)
# non_engager     engager 
#           2           2 

print(paste("Number of Upregulated Genes (FDR < 0.05):", length(blca_deseq2_test$up_genes)))
int(paste("Number of Upregulated Genes (FDR < 0.05):", length(blca_deseq2_test$up_genes)))
#[1] "Number of Upregulated Genes (FDR < 0.05): 163"

#### Module 2 ####
generate_deseq_qc_plots(blca_deseq2_test)

#### Module 3 ####
blca_gsva_scores <- calculate_sample_scores(blca_deseq2_test)

ggplot(blca_gsva_scores, aes(x = group, y = custom_metric, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  geom_jitter(aes(color = group), width = 0.2, size = 2.5, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(
    title = "BLCA: Sample-Level Gene Set Scoring",
    subtitle = paste0("Based on ", length(blca_deseq2_test$up_genes), "-gene Signature"),
    y = "GSVA Enrichment Score",
    x = NULL 
  ) +
  theme(legend.position = "none", plot.title = element_text(face = "bold")) +
  stat_compare_means(method = "wilcox.test", label.x = 1.5)

#### MSigDB interrogation ####
vst_blca <- vst(blca_deseq2_test$dds, blind=FALSE)
blca_vst_mat <- assay(vst_blca)
rownames(blca_vst_mat) <- blca_deseq2_test$res_df$gene_name[match(rownames(blca_vst_mat), rownames(blca_deseq2_test$res_df))]
blca_vst_mat <- blca_vst_mat[!is.na(rownames(blca_vst_mat)), ]

blca_immune_gsva <- gsva(expr = blca_vst_mat, gset.idx.list = target_pathways, method = "gsva", verbose = FALSE)

blca_immune_scores <- data.frame(
  case_id = colnames(blca_vst_mat),
  ifn_alpha = as.numeric(blca_immune_gsva["ifn_alpha", ]),
  ifn_gamma = as.numeric(blca_immune_gsva["ifn_gamma", ]),
  inflammatory = as.numeric(blca_immune_gsva["inflammatory", ]),
  group = blca_deseq2_test$dds$engager_group
)

blca_immune_long <- blca_immune_scores %>%
  pivot_longer(cols = c(ifn_alpha, ifn_gamma, inflammatory), names_to = "pathway", values_to = "score")

ggplot(blca_immune_long, aes(x = group, y = score, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  geom_jitter(aes(color = group), width = 0.2, size = 2, alpha = 0.8) +
  facet_wrap(~ pathway) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(title = "BLCA: Targeted Immune Pathway GSVA Scores", y = "GSVA Enrichment Score", x = NULL) +
  theme(legend.position = "none", plot.title = element_text(face = "bold")) +
  stat_compare_means(method = "wilcox.test", label = "p.format", label.x = 1.3)

#### Overlap and Core Signature ####
blca_deseq2_ifn_overlap <- intersect(blca_deseq2_test$up_genes, ifn_genes)
print(paste("Number of overlapping IFN genes:", length(blca_deseq2_ifn_overlap)))
print("Intersection with BRCA up_genes:")
print(intersect(blca_deseq2_test$up_genes, brca_deseq2_test$up_genes))

blca_core_scores <- calculate_sample_scores(blca_deseq2_test, custom_index = core_11_genes)
ggplot(blca_core_scores, aes(x = group, y = custom_metric, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  geom_jitter(aes(color = group), width = 0.2, size = 2.5, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(title = "BLCA: 11-Gene Core Signature", y = "GSVA Enrichment Score", x = NULL) +
  theme(legend.position = "none", plot.title = element_text(face = "bold")) +
  stat_compare_means(method = "wilcox.test", label.x = 1.5)
```

