---
title: "de_analysis.Rmd"
output: html_document
date: "2026-04-09"
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(results = "hide")


```


```{r packages}
install.packages("BiocManager")

# Install the necessary bioinformatics packages
# TCGAbiolinks: for data access
# GSVA: for the 94-gene signature scoring
# SummarizedExperiment: the standard data container for TCGA data
BiocManager::install(c("TCGAbiolinks", "GSVA", "SummarizedExperiment"))
BiocManager::install("IlluminaHumanMethylation450kanno.ilmn12.hg19")
BiocManager::install("sesameData")
BiocManager::install("sesame")
BiocManager::install("sesame")
BiocManager::install("edgeR")
BiocManager::install("org.Hs.eg.db")
BiocManager::install("DESeq2")
BiocManager::install("apeglm", update = FALSE, ask = FALSE)
BiocManager::install("survminer")
BiocManager::install(c("BiocGenerics", "S4Vectors", "IRanges"), force = TRUE, update = TRUE, ask = FALSE)
BiocManager::install("DESeq2", force=TRUE, update = FALSE, ask = FALSE)

install.packages("nloptr")
install.packages("ggpubr")
install.packages("cowplot")
install.packages("gt")
install.packages("minifi")
install.packages("glmnet")
install.packages("ggrepel")


# 4/15/26 trying to get DESeq2
# 1. Remove the mismatched packages from your local Linux library
remove.packages(c("BiocGenerics", "S4Vectors", "IRanges"))

# 2. Force BiocManager to use the correct repository for R 4.3
BiocManager::install(version = "3.18", ask = FALSE)

# 3. Install DESeq2 cleanly from the 3.18 repository
BiocManager::install("DESeq2", update = FALSE, ask = FALSE)

BiocManager::install(c("BiocGenerics", "IRanges", "S4Vectors", "DESeq2"), update = FALSE, ask = FALSE)

# 1. Delete the corrupted phantom folder directly from the file system
unlink("/home/nvsharm/R/x86_64-pc-linux-gnu-library/4.3/S4Vectors", recursive = TRUE)

# 2. Install S4Vectors by pointing explicitly to the Bioconductor 3.18 repository
install.packages("S4Vectors", repos = "https://bioconductor.org/packages/3.18/bioc")

install.packages("IRanges", repos = "https://bioconductor.org/packages/3.18/bioc")

# 3. Test DESeq2
library(DESeq2)




# Install clusterProfiler and msigdbr
BiocManager::install("clusterProfiler")
install.packages("msigdbr")

install.packages("PRROC")
install.packages("webshot2")

```


```{r load packages}
library(BiocManager)
library(TCGAbiolinks)
library(GSVA)
library(SummarizedExperiment)
library(tidyverse)
library(nloptr)
library(ggpubr)
library(cowplot)
library(gt)
library(webshot2)
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
library(sesameData)
library(sesame)
library(edgeR)
library(org.Hs.eg.db)
library(AnnotationDbi)
library(S4Vectors)
library(IRanges)
library(DESeq2)
library(apeglm)
library(clusterProfiler)
library(msigdbr)
library(glmnet)
library(survminer)
library(ggrepel)
library(PRROC)
library(conflicted)
library(webshot2)

conflict_prefer("select", "dplyr")
conflict_prefer("count", "dplyr")
conflict_prefer("filter", "dplyr")
```

```{r read objects and files}

candidate_sample_groups <- readRDS("candidate_sample_groups.rds")
pol3_loci_long <- readRDS("pol3_loci_long.rds")
pol3_gr <- readRDS("pol3_gr.rds")
ifn_genes <- readRDS("~/ifn_genes.rds")
gmm_sample_groups <- readr::read_tsv("~/gmm_grouping/gmm_sample_groups.tsv")
gdc_sample_metadata <- read_csv("~/gdc_sample_sheet_1.txt") 

# Reformat gdc_sample_metadata and join it to gmm_groups, then reformat so it will work with downstream functions
gdc_sample_metadata_clean <- gdc_sample_metadata %>% 
  mutate(sample = str_replace(project_id, "TCGA-", "") %>% str_replace("-", ".")) 

# Map gmm sample ids to case_id, dedup, and add engager group classification
gmm_sample_groups_clean <- gmm_sample_groups %>%
  filter(grouping_label %in% c("Engager", "Non_Engager")) %>%
  mutate(engager_group = if_else(grouping_label == "Engager", "engager", "non_engager")) %>%
  left_join(gdc_sample_metadata_clean %>% dplyr::select(sample, case_id, sample_id, sample_type),
            by = "sample") %>%
  group_by(case_id) %>%
  slice_sample(n = 1) %>%
  ungroup()

gmm_sample_groups_clean %>% filter(is.na(case_id)) %>% nrow()
gmm_sample_groups_clean %>% count(cancer_type, engager_group)

gmm_sample_groups %>%
  filter(grouping_label %in% c("Engager", "Non_Engager")) %>%
  mutate(engager_group = if_else(grouping_label == "Engager", "engager", "non_engager")) %>%
  left_join(gdc_sample_metadata_clean %>% dplyr::select(sample, case_id), by = "sample") %>%
  group_by(case_id) %>%
  filter(n_distinct(engager_group) > 1) %>%
  ungroup()

```

# Extract raw counts for de analysis
Note: HNSC has the strongest biological signal but the smallest ATAC-seq sample size in this dataset.
```{r}


candidates <- c("BRCA", "COAD", "KIRP", "STAD", "KIRC", "BLCA")



prepare_rnaseq <- function(cancer_type, matched_samples) {
  message("Processing ", cancer_type)
  
  query <- GDCquery(
    project = paste0("TCGA-", cancer_type),
    data.category = "Transcriptome Profiling",
    data.type = "Gene Expression Quantification",
    workflow.type = "STAR - Counts",
    sample.type = "Primary Tumor"
  )
  
  se <- GDCprepare(query)
  
  # Filter to ATAC-seq matched samples only
  matched_barcodes <- matched_samples %>%
    filter(cancer_type == .env$cancer_type) %>%
    pull(case_id)
  
  #se_matched <- se[, se$barcode %in% matched_barcodes]
  se_matched <- se[, substr(colData(se)$barcode, 1, 12) %in% substr(matched_barcodes, 1, 12)]
  se_matched <- se_matched[, !duplicated(substr(colData(se_matched)$barcode, 1, 12))]
  
  message("  Matched ", ncol(se_matched), " of ", length(matched_barcodes), " samples")
  
  list(
    counts = assay(se_matched, "unstranded"),
    se = se_matched,
    meta = colData(se_matched) %>%
      as.data.frame() %>%
      mutate(case_id = substr(barcode, 1, 12)) %>%
      left_join(
        matched_samples %>%
          filter(cancer_type == .env$cancer_type) %>%
          dplyr::select(case_id, engager_group),
        by = "case_id"
      )
  )
}
```

```{r}

prepare_and_cache_rnaseq <- function(cancer_type, matched_samples = NULL, matched_only = TRUE, cache_dir = "tcga_cache") {
  
  if(!dir.exists(cache_dir)) dir.create(cache_dir)
  
  # 1. Generate unique cache paths
  suffix <- if(matched_only) "_matched.rds" else "_full.rds"
  cache_file <- file.path(cache_dir, paste0(cancer_type, suffix))
  
  # 2. Return cached data if available
  if(file.exists(cache_file)) {
    message("\n>>> Loading CACHED RNA-seq data: ", cancer_type, " (", if(matched_only) "Matched" else "Full", ")")
    return(readRDS(cache_file))
  }
  
  # 3. Download from GDC
  message("\n>>> Downloading NEW RNA-seq data for ", cancer_type)
  query <- GDCquery(
    project = paste0("TCGA-", cancer_type),
    data.category = "Transcriptome Profiling",
    data.type = "Gene Expression Quantification",
    workflow.type = "STAR - Counts",
    sample.type = "Primary Tumor"
  )
  GDCdownload(query)
  se <- GDCprepare(query)
  
  # 4. Standardize Patient IDs (12-char barcode)
  colData(se)$case_id <- substr(colData(se)$barcode, 1, 12)
  
  # 5. Execute Branching Logic
  if(matched_only) {
    # Discovery Mode: Filter to the 29 matched samples
    if(is.null(matched_samples)) stop("Error: matched_samples required when matched_only = TRUE")
    
    target_ids <- matched_samples %>%
      filter(cancer_type == .env$cancer_type) %>%
      pull(case_id)
    
    se_final <- se[, colData(se)$case_id %in% target_ids]
    se_final <- se_final[, !duplicated(colData(se_final)$case_id)]
    
    meta_df <- as.data.frame(colData(se_final)) %>%
      left_join(matched_samples %>% 
                  filter(cancer_type == .env$cancer_type) %>% 
                  dplyr::select(case_id, engager_group), 
                by = "case_id")
  } else {
    # Validation Mode: Keep all patients (~1,100 for BRCA)
    se_final <- se[, !duplicated(colData(se)$case_id)]
    meta_df <- as.data.frame(colData(se_final))
  }
  
  # 6. Save and Return
  final_data <- list(
    counts = assay(se_final, "unstranded"),
    se = se_final,
    meta = meta_df
  )
  
  saveRDS(final_data, cache_file)
  message("  Saved to cache: ", cache_file)
  return(final_data)
}

# prepare_and_cache_rnaseq <- function(cancer_type, matched_samples = NULL, cache_dir = "tcga_cache") {
#   
#   # 1. Setup cache directory
#   if(!dir.exists(cache_dir)) dir.create(cache_dir)
#   suffix <- if(matched_only) "_matched.rds" else "_full.rds"
#   cache_file <- file.path(cache_dir, paste0(cancer_type, suffix))
#   #cache_file <- file.path(cache_dir, paste0(cancer_type, "_rnaseq_matched.rds"))
#   
#   # 2. Check for cached data
#   # if(file.exists(cache_file)) {
#   #   message("\n>>> Loading CACHED RNA-seq data for ", cancer_type)
#   #   return(readRDS(cache_file))
#   # }
#   if(file.exists(cache_file)) {
#     message("\n>>> Loading CACHED RNA-seq data for ", cancer_type, " (", if(matched_only) "Matched" else "Full", ")")
#     return(readRDS(cache_file))
#   }
#   
#   # 3. If no cache, perform the download and processing
#   message("\n>>> Downloading NEW RNA-seq data for ", cancer_type)
#   
#   query <- GDCquery(
#     project = paste0("TCGA-", cancer_type),
#     data.category = "Transcriptome Profiling",
#     data.type = "Gene Expression Quantification",
#     workflow.type = "STAR - Counts",
#     sample.type = "Primary Tumor"
#   )
#   
#   GDCdownload(query = query)
#   se <- GDCprepare(query)
#   
#   # CHANGE 2: Standardize IDs immediately to 12-char format (TCGA-XX-XXXX)
#   colData(se)$case_id <- substr(colData(se)$barcode, 1, 12)
#   
#   # Filtering logic changed for running full cohort
#   # 4. FILTERING LOGIC
#   if(matched_only) {
#     # CHANGE 3: This is your original logic for the 29 matched samples
#     matched_ids <- matched_samples %>%
#       filter(cancer_type == .env$cancer_type) %>%
#       pull(case_id)
#     
#     se_final <- se[, colData(se)$case_id %in% matched_ids]
#     # Deduplicate to ensure 1 sample per patient for the model
#     se_final <- se_final[, !duplicated(colData(se_final)$case_id)]
#     
#     # Metadata Join (Only for Matched Samples because they have Ground Truth labels)
#     meta_df <- colData(se_final) %>%
#       as.data.frame() %>%
#       left_join(
#         matched_samples %>%
#           filter(cancer_type == .env$cancer_type) %>%
#           dplyr::select(case_id, engager_group),
#         by = "case_id"
#       )
#   } else {
#     # CHANGE 4: Logic for the Full Cohort (Validation)
#     # No filtering by matched_samples; just deduplicate patients
#     se_final <- se[, !duplicated(colData(se)$case_id)]
#     meta_df <- as.data.frame(colData(se_final))
#   }
#   
#   message("  Final dataset contains: ", ncol(se_final), " samples")
#   
#   # 5. SAVE AND RETURN
#   final_data <- list(
#     counts = assay(se_final, "unstranded"),
#     se = se_final,
#     meta = meta_df
#   )
#   
#   saveRDS(final_data, cache_file)
#   message("  Saved to cache: ", cache_file)
#   
#   return(final_data)
# 
# #   matched_barcodes <- matched_samples %>%
# #     filter(cancer_type == .env$cancer_type) %>%
# #     pull(case_id)
# #   
# #   se_matched <- se[, substr(colData(se)$barcode, 1, 12) %in% substr(matched_barcodes, 1, 12)]
# #   se_matched <- se_matched[, !duplicated(substr(colData(se_matched)$barcode, 1, 12))]
# #   
# #   message("  Matched ", ncol(se_matched), " of ", length(matched_barcodes), " samples")
# #   
# #   # Compile the final object
# #   final_data <- list(
# #     counts = assay(se_matched, "unstranded"),
# #     se = se_matched,
# #     meta = colData(se_matched) %>%
# #       as.data.frame() %>%
# #       mutate(case_id = substr(barcode, 1, 12)) %>%
# #       left_join(
# #         matched_samples %>%
# #           filter(cancer_type == .env$cancer_type) %>%
# #           dplyr::select(case_id, engager_group),
# #         by = "case_id"
# #       )
# #   )
# #   
# #   # 4. Save to cache
# #   saveRDS(final_data, cache_file)
# #   message("  Saved to cache: ", cache_file)
# #   
# #   return(final_data)
# # }
# }
```


# DE analysis
1) run edgeR engagers vs non_engagers
2) get up and down genes
3) Are any up genes in IFN signature genes?
4) Do up genes have engager loci within them?

Then we can test in BRCA 
## edgeR
```{r}
run_de_analysis <- function(cancer_type, matched_samples, pol3_loci_long, ifn_genes) {
  
  message("Running DE for ", cancer_type)
  
  # Prepare RNA-seq data
  data <- prepare_rnaseq(cancer_type, matched_samples)
  
  # 1. EdgeR DE engager vs non_engager ####
  dge <- DGEList(counts = data$counts, group = data$meta$engager_group)
  keep <- filterByExpr(dge)
  dge <- dge[keep, , keep.lib.sizes = FALSE]
  dge <- calcNormFactors(dge)
  
  design <- model.matrix(~ 0 + engager_group, data = data$meta)
  colnames(design) <- levels(factor(data$meta$engager_group))
  dge <- estimateDisp(dge, design)
  fit <- glmQLFit(dge, design)
  
  con <- makeContrasts(engager - non_engager, levels = design)
  qlf <- glmQLFTest(fit, contrast = con)
  
  results <- topTags(qlf, n = Inf)$table
  results$ensembl_id <- rownames(results) %>% str_remove("\\..*")
  results$gene_name <- mapIds(org.Hs.eg.db,
                              keys = results$ensembl_id,
                              column = "SYMBOL",
                              keytype = "ENSEMBL",
                              multiVals = "first")
  
  # 2. Upregulated and downregulated genes ####
  up_genes <- results %>%
    filter(FDR < 0.05, !is.na(gene_name), logFC > 0) %>%
    pull(gene_name)
  
  down_genes <- results %>%
    filter(FDR < 0.05, !is.na(gene_name), logFC < 0) %>%
    pull(gene_name)
  
  # 3. IFN overlap ####
  ifn_overlap <- intersect(up_genes, ifn_genes)
  
  # 4. Engager loci overlap ####
  engager_loci <- pol3_loci_long %>%
    filter(cancer_type == .env$cancer_type,
           is_5ppp == 1,
           accessibility == 1) %>%
    distinct(V1, V2, V3)
  
  engager_gr <- GRanges(
    seqnames = engager_loci$V1,
    ranges = IRanges(start = as.integer(engager_loci$V2),
                     end = as.integer(engager_loci$V3))
  )
  
  up_gr <- rowRanges(data$se)[rowData(data$se)$gene_name %in% up_genes]
  overlaps <- findOverlaps(up_gr, engager_gr)
  genes_with_engager_loci <- names(up_gr)[queryHits(overlaps)]
  
  list(
    de_results = results,
    up_genes = up_genes,
    down_genes = down_genes,
    ifn_overlap = ifn_overlap,
    genes_with_engager_loci = genes_with_engager_loci,
    n_samples = nrow(data$meta),
    n_per_group = table(data$meta$engager_group)
  )
}
```

### For one cancer type
```{r}
brca_de <- run_de_analysis("BRCA", candidate_sample_groups, pol3_loci_long, ifn_genes)

# How many DE genes
length(brca_de$up_genes) #11
length(brca_de$down_genes) #20

# Any IFN overlap
brca_de$ifn_overlap

# Any genes with engager loci
brca_de$genes_with_engager_loci

# Sample counts
brca_de$n_per_group

# Are there any IFN genes that are even in the expression matrix and were filtered out?
intersect(ifn_genes, rownames(brca_de$de_results %>% filter(!is.na(gene_name)))) # 0 in expression matrix


```

### For all cancer types
```{r}
candidates <- c("COAD", "KIRP", "STAD", "KIRC", "BLCA") #"BRCA", 

de_results_all <- lapply(candidates, run_de_analysis,
                         matched_samples = candidate_sample_groups,
                         pol3_loci_long = pol3_loci_long,
                         ifn_genes = ifn_genes)
names(de_results_all) <- candidates

de_results_all[["BRCA"]] <- brca_de

# Rename in active environment b/c doing DESeq2 also
#de_results_all_edgeR <- de_results_all

# Save object to disk
#saveRDS(de_results_all_edgeR, "de_results_all_edgeR.rds")

```

### Dig into DE results for all cancer types
```{r}
lapply(names(de_results_all), function(ct) {
  cat("\n", ct, "\n")
  cat("N per group:", de_results_all[[ct]]$n_per_group, "\n")
  cat("N up genes FDR<0.05:", length(de_results_all[[ct]]$up_genes), "\n")
  cat("N down genes FDR<0.05:", length(de_results_all[[ct]]$down_genes), "\n")
  cat("IFN overlap:", paste(de_results_all[[ct]]$ifn_overlap, collapse=", "), "\n")
  cat("Genes with engager loci:", paste(de_results_all[[ct]]$genes_with_engager_loci, collapse=", "), "\n")
})


```

## DESeq2
### Table of how many samples are in each cancer type for DE analysis
```{r}
candidate_sample_groups %>%
  count(cancer_type, engager_group)

candidate_sample_groups %>%
  count(cancer_type, engager_group) %>%
  pivot_wider(names_from = engager_group, values_from = n) %>%
  mutate(total = engager + non_engager) %>%
  gt() %>%
  cols_label(
    cancer_type = "Cancer Type",
    engager = "Engager",
    non_engager = "Non-Engager",
    total = "Total"
  ) %>%
  tab_header(title = "Sample counts by RIG-I engager group")
```


### Module 1 Core DESeq2 Function
```{r}

# Function handles matrix setup, variance stabilization, and DE testing 
# returns a list containing the raw dds object, the full results df, and intersection logic
run_deseq2_core <- function(cancer_type, matched_samples, pol3_loci_long, ifn_genes) {
  
  message("\n>>> Running DESeq2 Core for ", cancer_type)
  data <- prepare_and_cache_rnaseq(cancer_type, matched_samples)
  
  # Ensure reference level
  data$meta$engager_group <- factor(data$meta$engager_group, levels = c("non_engager", "engager"))
  
  dds <- DESeqDataSetFromMatrix(countData = data$counts,
                                colData = data$meta,
                                design = ~ engager_group)
  
  keep <- rowSums(counts(dds)) >= 10
  dds <- dds[keep,]
  dds <- DESeq(dds)
  res <- lfcShrink(dds, coef="engager_group_engager_vs_non_engager", type="apeglm")
  
  res_df <- as.data.frame(res)
  res_df$ensembl_id <- rownames(res_df) %>% str_remove("\\..*")
  res_df$gene_name <- mapIds(org.Hs.eg.db,
                             keys = res_df$ensembl_id,
                             column = "SYMBOL",
                             keytype = "ENSEMBL",
                             multiVals = "first")
  
  up_genes <- res_df %>% filter(padj < 0.05, !is.na(gene_name), log2FoldChange > 0.5) %>% pull(gene_name)
  down_genes <- res_df %>% filter(padj < 0.05, !is.na(gene_name), log2FoldChange < -0.5) %>% pull(gene_name)
  
  # Engager loci overlap logic
  engager_loci <- pol3_loci_long %>%
    filter(cancer_type == .env$cancer_type, is_5ppp == 1, accessibility == 1) %>%
    distinct(V1, V2, V3)
  
  engager_gr <- GRanges(seqnames = engager_loci$V1, ranges = IRanges(start = as.integer(engager_loci$V2), end = as.integer(engager_loci$V3)))
  up_gr <- rowRanges(data$se)[rowData(data$se)$gene_name %in% up_genes]
  overlaps <- findOverlaps(up_gr, engager_gr)
  genes_with_engager_loci <- names(up_gr)[queryHits(overlaps)]
  
  return(list(
    cancer_type = cancer_type,
    dds = dds,
    res_df = res_df,
    up_genes = up_genes,
    down_genes = down_genes,
    ifn_overlap = intersect(up_genes, ifn_genes),
    genes_with_engager_loci = genes_with_engager_loci,
    n_samples = nrow(data$meta),
    n_per_group = table(data$meta$engager_group)
  ))
}

```




### Module 2 QC Plotting
Dispersion plot, Volcano plot, PCA plot
```{r}
generate_deseq_qc_plots <- function(deseq_output) {
  ct <- deseq_output$cancer_type
  dds <- deseq_output$dds
  
  dir.create("DESeq_QC_Plots", showWarnings = FALSE)
  
  # Dispersion Plot
  png(paste0("DESeq_QC_Plots/", ct, "_dispersion.png"), width=600, height=500)
  plotDispEsts(dds, main=paste(ct, "Dispersion Estimates"))
  dev.off()
  
  # PCA Plot
  vsd <- vst(dds, blind=FALSE)
  pca_data <- plotPCA(vsd, intgroup="engager_group", returnData=TRUE)
  percentVar <- round(100 * attr(pca_data, "percentVar"))
  
  pca_plot <- ggplot(pca_data, aes(PC1, PC2, color=engager_group)) +
    geom_point(size=3, alpha=0.8) +
    theme_minimal() +
    labs(title=paste(ct, "PCA"),
         x=paste0("PC1: ", percentVar[1], "%"),
         y=paste0("PC2: ", percentVar[2], "%")) +
    scale_color_manual(values=c("non_engager"="#00BFC4", "engager"="#F8766D"))
  
  ggsave(paste0("DESeq_QC_Plots/", ct, "_PCA.png"), plot=pca_plot, width=5, height=4)
  
  # --- NEW VOLCANO PLOT SECTION ---
  # Identify Top 20 significant genes
  top_20 <- as.data.frame(deseq_output$res_df) %>%
    filter(!is.na(padj)) %>%
    arrange(padj) %>%
    slice_head(n = 20) %>%
    pull(gene_name)
  
  # Combine with biological overlap lists
  biological_overlaps <- c(deseq_output$ifn_overlap, deseq_output$genes_with_engager_loci)
  genes_to_label <- unique(c(top_20, biological_overlaps))
  
  # Prepare data: define colors and which genes to label
  volcano_data <- as.data.frame(deseq_output$res_df) %>%
    mutate(
      significance = case_when(
        padj < 0.05 & log2FoldChange > 0.5 ~ "Up",
        padj < 0.05 & log2FoldChange < -0.5 ~ "Down",
        TRUE ~ "NS"
      ),
      label = if_else(gene_name %in% genes_to_label & padj < 0.05, gene_name, NA_character_)
    )

  volcano_plot <- ggplot(volcano_data, aes(x=log2FoldChange, y=-log10(padj), color=significance)) +
    geom_point(alpha=0.4, size=1.5) +
    scale_color_manual(values=c("Up"="#F8766D", "Down"="#00BFC4", "NS"="grey70")) +
    geom_text_repel(aes(label=label), color="black", size=3, max.overlaps = 30) +
    theme_minimal() +
    labs(title=paste(ct, "Volcano Plot"), 
         subtitle="Labeled: Top 20 Significant + Biological Overlaps",
         x="log2 Fold Change", y="-log10 Adjusted P-value")
  
  ggsave(paste0("DESeq_QC_Plots/", ct, "_volcano.png"), plot=volcano_plot, width=7, height=6)
  
  # --- END VOLCANO SECTION ---
  
  message("Saved QC plots for ", ct)
}
```


### Module 3 GSVA sample-by-sample scoring
```{r}
calculate_sample_scores <- function(deseq_output, custom_index = NULL) {
  
  # Default to the cancer-specific up_genes if no index is provided
  if(is.null(custom_index)) {
    custom_index <- deseq_output$up_genes
  }
  
  if(length(custom_index) == 0) {
    message("No index genes available for ", deseq_output$cancer_type)
    return(NULL)
  }
  
  # Extract VST normalized counts
  vsd <- vst(deseq_output$dds, blind=FALSE)
  vst_mat <- assay(vsd)
  
  # Map rownames from Ensembl to Symbol for GSVA matching
  rownames(vst_mat) <- deseq_output$res_df$gene_name[match(rownames(vst_mat), rownames(deseq_output$res_df))]
  vst_mat <- vst_mat[!is.na(rownames(vst_mat)), ] # Drop any genes that didn't map correctly to a symbol (NAs)
  
  # Put the up genes (for brca = 133 up genes) into a list format for GSVA package
  index_list <- list(Target_Index = custom_index)
  
  # run GSVA to get one score per column (per sample)
  gsva_res <- gsva(expr = vst_mat, gset.idx.list = index_list, method = "gsva", verbose = FALSE)
  
  # format into a table that can be used to make a boxplot
  sample_scores <- data.frame(
    case_id = colnames(vst_mat),
    custom_metric = as.numeric(gsva_res["Target_Index", ]),
    group = deseq_output$dds$engager_group
  )
  
  return(sample_scores)
}
```

### BRCA 
```{r}
#### Module 1_____ ####
brca_deseq2_test <- run_deseq2_core("BRCA", candidate_sample_groups, pol3_loci_long, ifn_genes)
brca_deseq2_test$n_per_group
# non_engager     engager 
#          15          14
length(brca_deseq2_test$up_genes)
# [1] 169
brca_deseq2_test$res_df %>% View


#### Module 2______ ####
# Generate the PCA and Dispersion plots and volcano plots
generate_deseq_qc_plots(brca_deseq2_test)


#### Module 3______ ####
# Generate the custom sample-by-sample metric
brca_gsva_scores <- calculate_sample_scores(brca_deseq2_test)

# Visualize the custom 133 gene score
group_colors <- c("non_engager" = "#4575b4", "engager" = "#d73027")

ggplot(brca_gsva_scores, aes(x = group, y = custom_metric, fill = group)) +
  # Create the boxplot with a bit of transparency
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  # Add the sample points (dots) and color them by group
  geom_jitter(aes(color = group), width = 0.2, size = 2.5, alpha = 0.8) +
  # Use a clean, professional theme
  theme_minimal() +
  # Map the colors we defined
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  # Refine labels and titles
  labs(
    title = "BRCA: Sample-Level Gene Set Scoring",
    subtitle = paste0("Based on ", length(brca_deseq2_test$up_genes), "-gene Signature"),
    y = "GSVA Enrichment Score",
    x = NULL # Removing 'group' label as the axis categories make it obvious
  ) +
  # Remove the legend since the x-axis already labels the groups
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 11, color = "black"),
    panel.grid.minor = element_blank()
  ) +
  stat_compare_means(method = "wilcox.test", label.x = 1.5, label.y = 0.6)




ggplot(brca_gsva_scores, aes(x = group, y = custom_metric, fill = group)) +
  geom_boxplot() +
  geom_jitter(width = 0.2) +
  theme_minimal() +
  labs(title = "BRCA: Sample-by-Sample Custom Metric (133-Gene Index)", y = "GSVA Score")


#### biological proof: enrichment of immunostimulatory pathways______ ####
# 2. Pull the Hallmark Interferon Alpha list
# ifn_alpha <- msigdbr(species = "human", category = "H") %>% 
#   filter(gs_name == "HALLMARK_INTERFERON_ALPHA_RESPONSE") %>% 
#   pull(gene_symbol)

# 2. See which of 33 genes overlap with ifn_genes
brca_deseq2_ifn_overlap <- intersect(brca_deseq2_test$up_genes, ifn_genes)

# 3. Check the results
print(paste("Number of overlapping IFN genes:", length(brca_deseq2_ifn_overlap)))
print(brca_deseq2_ifn_overlap)

# Check overlap with edgeR genes
intersect(brca_deseq2_test$up_genes, brca_de$up_genes)


#### MSigDB interrogation of up genes______####

# 1. Pull the official Hallmark immune sets
msigdb_hallmark_sets <- msigdbr(species = "human", category = "H")
ifn_alpha_list <- msigdb_hallmark_sets[msigdb_hallmark_sets$gs_name == "HALLMARK_INTERFERON_ALPHA_RESPONSE", ]$gene_symbol
ifn_gamma_list <- msigdb_hallmark_sets[msigdb_hallmark_sets$gs_name == "HALLMARK_INTERFERON_GAMMA_RESPONSE", ]$gene_symbol
inflam_list <- msigdb_hallmark_sets[msigdb_hallmark_sets$gs_name == "HALLMARK_INFLAMMATORY_RESPONSE", ]$gene_symbol

# Create a list of these specific pathways
target_pathways <- list(
  ifn_alpha = ifn_alpha_list,
  ifn_gamma = ifn_gamma_list,
  inflammatory = inflam_list
)

# 2. Extract VST normalized counts and convert Ensembl IDs to Gene Symbols
vst_brca <- vst(brca_deseq2_test$dds, blind=FALSE)
brca_vst_mat <- assay(vst_brca)
rownames(brca_vst_mat) <- brca_deseq2_test$res_df$gene_name[match(rownames(brca_vst_mat), rownames(brca_deseq2_test$res_df))]
brca_vst_mat <- brca_vst_mat[!is.na(rownames(brca_vst_mat)), ]

# 3. Score the BRCA samples against ONLY these immune pathways
brca_immune_gsva <- gsva(expr = brca_vst_mat, 
                         gset.idx.list = target_pathways, 
                         method = "gsva", 
                         verbose = FALSE)


# 4. Format all three pathways into a dataframe
brca_immune_scores <- data.frame(
  case_id = colnames(brca_vst_mat),
  ifn_alpha = as.numeric(brca_immune_gsva["ifn_alpha", ]),
  ifn_gamma = as.numeric(brca_immune_gsva["ifn_gamma", ]), 
  inflammatory = as.numeric(brca_immune_gsva["inflammatory", ]),
  group = brca_deseq2_test$dds$engager_group
)

# Reshape data for ggplot faceting
brca_immune_long <- brca_immune_scores %>%
  pivot_longer(cols = c(ifn_alpha, ifn_gamma, inflammatory),
               names_to = "pathway",
               values_to = "score")

# 5. Plot the result for IFN Alpha
# 5. Plot all three pathways side-by-side
ggplot(brca_immune_long, aes(x = group, y = score, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  geom_jitter(aes(color = group), width = 0.2, size = 2, alpha = 0.8) +
  facet_wrap(~ pathway) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(
    title = "BRCA: Targeted Immune Pathway GSVA Scores", 
    y = "GSVA Enrichment Score", 
    x = NULL
  ) +
  theme(
    legend.position = "none", 
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 11, color = "black"),
    strip.text = element_text(size = 12, face = "bold"),
    panel.grid.minor = element_blank()
  ) +
  stat_compare_means(method = "wilcox.test", label = "p.format", label.x = 1.3)
```


### COAD 
```{r}
#### Module 1 ####
coad_deseq2_test <- run_deseq2_core("COAD", candidate_sample_groups, pol3_loci_long, ifn_genes)
coad_deseq2_test$n_per_group
# non_engager     engager 
#          9          8
length(coad_deseq2_test$up_genes)
# [1] 26
coad_deseq2_test$res_df %>% View


#### Module 2 ####
# Generate the PCA and Dispersion plots
generate_deseq_qc_plots(coad_deseq2_test)


#### Module 3 ####
# Generate the custom sample-by-sample metric
coad_gsva_scores <- calculate_sample_scores(coad_deseq2_test)


# Visualize the custom 133 gene score
group_colors <- c("non_engager" = "#4575b4", "engager" = "#d73027")

ggplot(coad_gsva_scores, aes(x = group, y = custom_metric, fill = group)) +
  # Create the boxplot with a bit of transparency
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  # Add the sample points (dots) and color them by group
  geom_jitter(aes(color = group), width = 0.2, size = 2.5, alpha = 0.8) +
  # Use a clean, professional theme
  theme_minimal() +
  # Map the colors we defined
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  # Refine labels and titles
  labs(
    title = "COAD: Sample-Level Gene Set Scoring",
    subtitle = paste0("Based on ", length(brca_deseq2_test$up_genes), "-gene Signature"),
    y = "GSVA Enrichment Score",
    x = NULL # Removing 'group' label as the axis categories make it obvious
  ) +
  # Remove the legend since the x-axis already labels the groups
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 11, color = "black"),
    panel.grid.minor = element_blank()
  ) +
  stat_compare_means(method = "wilcox.test", label.x = 1.5, label.y = 0.6)

#### MSigDB interrogation of immune pathways______####

# # 1. Pull the official Hallmark immune sets (re-defining to ensure safe exact-match names)
# msigdb_hallmark_sets <- msigdbr(species = "human", category = "H")
# ifn_alpha_list <- msigdb_hallmark_sets[msigdb_hallmark_sets$gs_name == "HALLMARK_INTERFERON_ALPHA_RESPONSE", ]$gene_symbol
# ifn_gamma_list <- msigdb_hallmark_sets[msigdb_hallmark_sets$gs_name == "HALLMARK_INTERFERON_GAMMA_RESPONSE", ]$gene_symbol
# inflam_list <- msigdb_hallmark_sets[msigdb_hallmark_sets$gs_name == "HALLMARK_INFLAMMATORY_RESPONSE", ]$gene_symbol
# 
# target_pathways <- list(
#   ifn_alpha = ifn_alpha_list,
#   ifn_Gamma = ifn_gamma_list,
#   inflammatory = inflam_list
# )

# 2. Extract VST normalized counts and convert Ensembl IDs to Gene Symbols
vst_coad <- vst(coad_deseq2_test$dds, blind=FALSE)
coad_vst_mat <- assay(vst_coad)
rownames(coad_vst_mat) <- coad_deseq2_test$res_df$gene_name[match(rownames(coad_vst_mat), rownames(coad_deseq2_test$res_df))]
coad_vst_mat <- coad_vst_mat[!is.na(rownames(coad_vst_mat)), ]

# 3. Score the COAD samples against ONLY these immune pathways
coad_immune_gsva <- gsva(expr = coad_vst_mat, 
                         gset.idx.list = target_pathways, 
                         method = "gsva", 
                         verbose = FALSE)

# 4. Format all three pathways into a dataframe
coad_immune_scores <- data.frame(
  case_id = colnames(coad_vst_mat),
  ifn_alpha = as.numeric(coad_immune_gsva["ifn_alpha", ]),
  ifn_gamma = as.numeric(coad_immune_gsva["ifn_gamma", ]),
  inflammatory = as.numeric(coad_immune_gsva["inflammatory", ]),
  group = coad_deseq2_test$dds$engager_group
)

# Reshape data for ggplot faceting
coad_immune_long <- coad_immune_scores %>%
  pivot_longer(cols = c(ifn_alpha, ifn_gamma, inflammatory),
               names_to = "pathway",
               values_to = "score")

# 5. Plot all three pathways side-by-side with a shared y-axis
ggplot(coad_immune_long, aes(x = group, y = score, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  geom_jitter(aes(color = group), width = 0.2, size = 2, alpha = 0.8) +
  facet_wrap(~ pathway) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(
    title = "COAD: Targeted Immune Pathway GSVA Scores", 
    y = "GSVA Enrichment Score", 
    x = NULL
  ) +
  theme(
    legend.position = "none", 
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 11, color = "black"),
    strip.text = element_text(size = 12, face = "bold"),
    panel.grid.minor = element_blank()
  ) +
  stat_compare_means(method = "wilcox.test", label = "p.format", label.x = 1.3)


#### Is there overlap of up genes with other cancertypes?____####
# 2. See which of 33 genes overlap with ifn_genes
coad_deseq2_ifn_overlap <- intersect(coad_deseq2_test$up_genes, ifn_genes)

# 3. Check the results
print(paste("Number of overlapping IFN genes:", length(coad_deseq2_ifn_overlap)))
print(coad_deseq2_ifn_overlap)

# Check overlap with brca up genes  
intersect(coad_deseq2_test$up_genes, brca_deseq2_test$up_genes)




```


#### Check if the 11 genes identified by edgeR in BRCA as upgenes (core genes) and score
```{r}

# Will see if we can use these 11 genes as a pan-cancer since the edgeR GO term enrichment in EnrichR showed Rig-i etc. and many of these genes were inflammatory/immunostimulatory
# Question: Why did edgeR or how did it narrow it down to these genes while DESeq introduced other genes that separated non_engagers from engagers but encapsulated different oncogenic processes?

# 1. Define the core 11-gene index from the BRCA edgeR/DESeq2 overlap
core_11_genes <- c("ABCC2", "EPN3", "LLGL2", "PSMD3", "LRRC59", 
                   "RAP2C", "PCDH10", "CXCR1", "CPA6", "PHB1", "ZNF652")

# 2. Score BRCA using ONLY the 11 core genes
brca_core_scores <- calculate_sample_scores(brca_deseq2_test, custom_index = core_11_genes)

# Plot BRCA Core Scores
p_brca_core <- ggplot(brca_core_scores, aes(x = group, y = custom_metric, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  geom_jitter(aes(color = group), width = 0.2, size = 2.5, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(
    title = "BRCA: 11-Gene Core Signature", 
    y = "GSVA Enrichment Score", 
    x = NULL
  ) +
  theme(legend.position = "none", plot.title = element_text(face = "bold")) +
  stat_compare_means(method = "wilcox.test", label.x = 1.5)

# 3. Score COAD using ONLY the 11 core genes
coad_core_scores <- calculate_sample_scores(coad_deseq2_test, custom_index = core_11_genes)

# Plot COAD Core Scores
p_coad_core <- ggplot(coad_core_scores, aes(x = group, y = custom_metric, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  geom_jitter(aes(color = group), width = 0.2, size = 2.5, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(
    title = "COAD: 11-Gene Core Signature", 
    y = "GSVA Enrichment Score", 
    x = NULL
  ) +
  theme(legend.position = "none", plot.title = element_text(face = "bold")) +
  stat_compare_means(method = "wilcox.test", label.x = 1.5)

# 4. Display side-by-side
plot_grid(p_brca_core, p_coad_core, ncol = 2)

```

### KIRP 
```{r}
#### Module 1 ####
kirp_deseq2_test <- run_deseq2_core("KIRP", candidate_sample_groups, pol3_loci_long, ifn_genes)
kirp_deseq2_test$n_per_group
# non_engager     engager 
#           9           8 

# Check how many genes passed the filter
length(kirp_deseq2_test$up_genes)
# 12

#### Module 2 ####
# Generate the PCA and Dispersion plots
generate_deseq_qc_plots(kirp_deseq2_test)


#### Module 3 ####
# Generate the custom sample-by-sample metric
kirp_gsva_scores <- calculate_sample_scores(kirp_deseq2_test)

# Visualize the custom metric
ggplot(kirp_gsva_scores, aes(x = group, y = custom_metric, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  geom_jitter(aes(color = group), width = 0.2, size = 2.5, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(
    title = "KIRP: Sample-Level Gene Set Scoring",
    subtitle = paste0("Based on ", length(kirp_deseq2_test$up_genes), "-gene Signature"),
    y = "GSVA Enrichment Score",
    x = NULL 
  ) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 11, color = "black"),
    panel.grid.minor = element_blank()
  ) +
  stat_compare_means(method = "wilcox.test", label.x = 1.5, label.y = 0.6)


#### MSigDB interrogation of immune pathways______####
# 1. Extract VST normalized counts and convert Ensembl IDs to Gene Symbols
vst_kirp <- vst(kirp_deseq2_test$dds, blind=FALSE)
kirp_vst_mat <- assay(vst_kirp)
rownames(kirp_vst_mat) <- kirp_deseq2_test$res_df$gene_name[match(rownames(kirp_vst_mat), rownames(kirp_deseq2_test$res_df))]
kirp_vst_mat <- kirp_vst_mat[!is.na(rownames(kirp_vst_mat)), ]

# 2. Score the KIRP samples against ONLY these immune pathways
# (Relies on target_pathways list defined in the BRCA chunk)
kirp_immune_gsva <- gsva(expr = kirp_vst_mat, 
                         gset.idx.list = target_pathways, 
                         method = "gsva", 
                         verbose = FALSE)

# 3. Format all three pathways into a dataframe
kirp_immune_scores <- data.frame(
  case_id = colnames(kirp_vst_mat),
  ifn_alpha = as.numeric(kirp_immune_gsva["ifn_alpha", ]),
  ifn_gamma = as.numeric(kirp_immune_gsva["ifn_gamma", ]),
  inflammatory = as.numeric(kirp_immune_gsva["inflammatory", ]),
  group = kirp_deseq2_test$dds$engager_group
)

# Reshape data for ggplot faceting
kirp_immune_long <- kirp_immune_scores %>%
  pivot_longer(cols = c(ifn_alpha, ifn_gamma, inflammatory),
               names_to = "pathway",
               values_to = "score")

# 4. Plot all three pathways side-by-side with a shared y-axis
ggplot(kirp_immune_long, aes(x = group, y = score, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  geom_jitter(aes(color = group), width = 0.2, size = 2, alpha = 0.8) +
  facet_wrap(~ pathway) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(
    title = "KIRP: Targeted Immune Pathway GSVA Scores", 
    y = "GSVA Enrichment Score", 
    x = NULL
  ) +
  theme(
    legend.position = "none", 
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 11, color = "black"),
    strip.text = element_text(size = 12, face = "bold"),
    panel.grid.minor = element_blank()
  ) +
  stat_compare_means(method = "wilcox.test", label = "p.format", label.x = 1.3)


#### Is there overlap of up genes with other cancertypes?____####
kirp_deseq2_ifn_overlap <- intersect(kirp_deseq2_test$up_genes, ifn_genes)

print(paste("Number of overlapping IFN genes:", length(kirp_deseq2_ifn_overlap)))
print(kirp_deseq2_ifn_overlap)

# Check overlap with brca up genes  
intersect(kirp_deseq2_test$up_genes, brca_deseq2_test$up_genes)


#### Check if the 11 genes identified by edgeR in BRCA score high here ____####
# (Relies on core_11_genes vector defined in previous chunks)
kirp_core_scores <- calculate_sample_scores(kirp_deseq2_test, custom_index = core_11_genes)

ggplot(kirp_core_scores, aes(x = group, y = custom_metric, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  geom_jitter(aes(color = group), width = 0.2, size = 2.5, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(
    title = "KIRP: 11-Gene Core Signature", 
    y = "GSVA Enrichment Score", 
    x = NULL
  ) +
  theme(legend.position = "none", plot.title = element_text(face = "bold")) +
  stat_compare_means(method = "wilcox.test", label.x = 1.5)
```


### STAD
```{r}
#### Module 1 ####
stad_deseq2_test <- run_deseq2_core("STAD", candidate_sample_groups, pol3_loci_long, ifn_genes)
stad_deseq2_test$n_per_group
# non_engager     engager 
#           4           4 

# Check how many genes passed the filter
length(stad_deseq2_test$up_genes)
# 383

#### Module 2 ####
# Generate the PCA and Dispersion plots
generate_deseq_qc_plots(stad_deseq2_test)


#### Module 3 ####
# Generate the custom sample-by-sample metric
stad_gsva_scores <- calculate_sample_scores(stad_deseq2_test)

# Visualize the custom metric
ggplot(stad_gsva_scores, aes(x = group, y = custom_metric, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  geom_jitter(aes(color = group), width = 0.2, size = 2.5, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(
    title = "STAD: Sample-Level Gene Set Scoring",
    subtitle = paste0("Based on ", length(stad_deseq2_test$up_genes), "-gene Signature"),
    y = "GSVA Enrichment Score",
    x = NULL 
  ) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 11, color = "black"),
    panel.grid.minor = element_blank()
  ) +
  stat_compare_means(method = "wilcox.test", label.x = 1.5, label.y = 0.6)


#### MSigDB interrogation of immune pathways______####
# 1. Extract VST normalized counts and convert Ensembl IDs to Gene Symbols
vst_stad <- vst(stad_deseq2_test$dds, blind=FALSE)
stad_vst_mat <- assay(vst_stad)
rownames(stad_vst_mat) <- stad_deseq2_test$res_df$gene_name[match(rownames(stad_vst_mat), rownames(stad_deseq2_test$res_df))]
stad_vst_mat <- stad_vst_mat[!is.na(rownames(stad_vst_mat)), ]

# 2. Score the STAD samples against ONLY these immune pathways
# (Relies on target_pathways list defined in the BRCA chunk)
stad_immune_gsva <- gsva(expr = stad_vst_mat, 
                         gset.idx.list = target_pathways, 
                         method = "gsva", 
                         verbose = FALSE)

# 3. Format all three pathways into a dataframe
stad_immune_scores <- data.frame(
  case_id = colnames(stad_vst_mat),
  ifn_alpha = as.numeric(stad_immune_gsva["ifn_alpha", ]),
  ifn_gamma = as.numeric(stad_immune_gsva["ifn_gamma", ]),
  inflammatory = as.numeric(stad_immune_gsva["inflammatory", ]),
  group = stad_deseq2_test$dds$engager_group
)

# Reshape data for ggplot faceting
stad_immune_long <- stad_immune_scores %>%
  pivot_longer(cols = c(ifn_alpha, ifn_gamma, inflammatory),
               names_to = "pathway",
               values_to = "score")

# 4. Plot all three pathways side-by-side with a shared y-axis
ggplot(stad_immune_long, aes(x = group, y = score, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  geom_jitter(aes(color = group), width = 0.2, size = 2, alpha = 0.8) +
  facet_wrap(~ pathway) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(
    title = "STAD: Targeted Immune Pathway GSVA Scores", 
    y = "GSVA Enrichment Score", 
    x = NULL
  ) +
  theme(
    legend.position = "none", 
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 11, color = "black"),
    strip.text = element_text(size = 12, face = "bold"),
    panel.grid.minor = element_blank()
  ) +
  stat_compare_means(method = "wilcox.test", label = "p.format", label.x = 1.3)


#### Is there overlap of up genes with other cancertypes?____####
stad_deseq2_ifn_overlap <- intersect(stad_deseq2_test$up_genes, ifn_genes)

print(paste("Number of overlapping IFN genes:", length(stad_deseq2_ifn_overlap)))
print(stad_deseq2_ifn_overlap)

# Check overlap with brca up genes  
intersect(stad_deseq2_test$up_genes, brca_deseq2_test$up_genes)
# [1] "CSMD3"    "B4GALNT2" "UGT3A2" 

#### Check if the 11 genes identified by edgeR in BRCA score high here ____####
# (Relies on core_11_genes vector defined in previous chunks)
stad_core_scores <- calculate_sample_scores(stad_deseq2_test, custom_index = core_11_genes)

ggplot(stad_core_scores, aes(x = group, y = custom_metric, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  geom_jitter(aes(color = group), width = 0.2, size = 2.5, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(
    title = "STAD: 11-Gene Core Signature", 
    y = "GSVA Enrichment Score", 
    x = NULL
  ) +
  theme(legend.position = "none", plot.title = element_text(face = "bold")) +
  stat_compare_means(method = "wilcox.test", label.x = 1.5)
```

KIRC
```{r}
#### Module 1 ####
kirc_deseq2_test <- run_deseq2_core("KIRC", candidate_sample_groups, pol3_loci_long, ifn_genes)
kirc_deseq2_test$n_per_group
# non_engager     engager 
#           3           2 
# Check how many genes passed the filter
length(kirc_deseq2_test$up_genes)
# 425

#### Module 2 ####
generate_deseq_qc_plots(kirc_deseq2_test)

#### Module 3 ####
kirc_gsva_scores <- calculate_sample_scores(kirc_deseq2_test)

ggplot(kirc_gsva_scores, aes(x = group, y = custom_metric, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  geom_jitter(aes(color = group), width = 0.2, size = 2.5, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(title = "KIRC: Sample-Level Gene Set Scoring",
       subtitle = paste0("Based on ", length(kirc_deseq2_test$up_genes), "-gene Signature"),
       y = "GSVA Enrichment Score", x = NULL) +
  theme(legend.position = "none", plot.title = element_text(face = "bold")) +
  stat_compare_means(method = "wilcox.test", label.x = 1.5)

#### MSigDB interrogation ####
vst_kirc <- vst(kirc_deseq2_test$dds, blind=FALSE)
kirc_vst_mat <- assay(vst_kirc)
rownames(kirc_vst_mat) <- kirc_deseq2_test$res_df$gene_name[match(rownames(kirc_vst_mat), rownames(kirc_deseq2_test$res_df))]
kirc_vst_mat <- kirc_vst_mat[!is.na(rownames(kirc_vst_mat)), ]

kirc_immune_gsva <- gsva(expr = kirc_vst_mat, gset.idx.list = target_pathways, method = "gsva", verbose = FALSE)

kirc_immune_scores <- data.frame(
  case_id = colnames(kirc_vst_mat),
  ifn_alpha = as.numeric(kirc_immune_gsva["ifn_alpha", ]),
  ifn_gamma = as.numeric(kirc_immune_gsva["ifn_gamma", ]),
  inflammatory = as.numeric(kirc_immune_gsva["inflammatory", ]),
  group = kirc_deseq2_test$dds$engager_group
)

kirc_immune_long <- kirc_immune_scores %>%
  pivot_longer(cols = c(ifn_alpha, ifn_gamma, inflammatory), names_to = "pathway", values_to = "score")

ggplot(kirc_immune_long, aes(x = group, y = score, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  geom_jitter(aes(color = group), width = 0.2, size = 2, alpha = 0.8) +
  facet_wrap(~ pathway) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(title = "KIRC: Targeted Immune Pathway GSVA Scores", y = "GSVA Enrichment Score", x = NULL) +
  theme(legend.position = "none", plot.title = element_text(face = "bold")) +
  stat_compare_means(method = "wilcox.test", label = "p.format", label.x = 1.3)

#### Overlap and Core Signature ####
kirc_deseq2_ifn_overlap <- intersect(kirc_deseq2_test$up_genes, ifn_genes)
print(paste("Number of overlapping IFN genes:", length(kirc_deseq2_ifn_overlap)))
intersect(kirc_deseq2_test$up_genes, brca_deseq2_test$up_genes)
# [1] "DHRS2" "WNT2"  "PADI3"

kirc_core_scores <- calculate_sample_scores(kirc_deseq2_test, custom_index = core_11_genes)
ggplot(kirc_core_scores, aes(x = group, y = custom_metric, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  geom_jitter(aes(color = group), width = 0.2, size = 2.5, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(title = "KIRC: 11-Gene Core Signature", y = "GSVA Enrichment Score", x = NULL) +
  theme(legend.position = "none", plot.title = element_text(face = "bold")) +
  stat_compare_means(method = "wilcox.test", label.x = 1.5)
```

#### BLCA
```{r}
#### BLCA 
#### Module 1 ####
blca_deseq2_test <- run_deseq2_core("BLCA", candidate_sample_groups, pol3_loci_long, ifn_genes)
print("Sample Sizes for BLCA:")
print(blca_deseq2_test$n_per_group)
# non_engager     engager 
#           2           2 

print(paste("Number of Upregulated Genes (FDR < 0.05):", length(blca_deseq2_test$up_genes)))
int(paste("Number of Upregulated Genes (FDR < 0.05):", length(blca_deseq2_test$up_genes)))
#[1] "Number of Upregulated Genes (FDR < 0.05): 163"

#### Module 2 ####
generate_deseq_qc_plots(blca_deseq2_test)

#### Module 3 ####
blca_gsva_scores <- calculate_sample_scores(blca_deseq2_test)

ggplot(blca_gsva_scores, aes(x = group, y = custom_metric, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  geom_jitter(aes(color = group), width = 0.2, size = 2.5, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(
    title = "BLCA: Sample-Level Gene Set Scoring",
    subtitle = paste0("Based on ", length(blca_deseq2_test$up_genes), "-gene Signature"),
    y = "GSVA Enrichment Score",
    x = NULL 
  ) +
  theme(legend.position = "none", plot.title = element_text(face = "bold")) +
  stat_compare_means(method = "wilcox.test", label.x = 1.5)

#### MSigDB interrogation ####
vst_blca <- vst(blca_deseq2_test$dds, blind=FALSE)
blca_vst_mat <- assay(vst_blca)
rownames(blca_vst_mat) <- blca_deseq2_test$res_df$gene_name[match(rownames(blca_vst_mat), rownames(blca_deseq2_test$res_df))]
blca_vst_mat <- blca_vst_mat[!is.na(rownames(blca_vst_mat)), ]

blca_immune_gsva <- gsva(expr = blca_vst_mat, gset.idx.list = target_pathways, method = "gsva", verbose = FALSE)

blca_immune_scores <- data.frame(
  case_id = colnames(blca_vst_mat),
  ifn_alpha = as.numeric(blca_immune_gsva["ifn_alpha", ]),
  ifn_gamma = as.numeric(blca_immune_gsva["ifn_gamma", ]),
  inflammatory = as.numeric(blca_immune_gsva["inflammatory", ]),
  group = blca_deseq2_test$dds$engager_group
)

blca_immune_long <- blca_immune_scores %>%
  pivot_longer(cols = c(ifn_alpha, ifn_gamma, inflammatory), names_to = "pathway", values_to = "score")

ggplot(blca_immune_long, aes(x = group, y = score, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  geom_jitter(aes(color = group), width = 0.2, size = 2, alpha = 0.8) +
  facet_wrap(~ pathway) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(title = "BLCA: Targeted Immune Pathway GSVA Scores", y = "GSVA Enrichment Score", x = NULL) +
  theme(legend.position = "none", plot.title = element_text(face = "bold")) +
  stat_compare_means(method = "wilcox.test", label = "p.format", label.x = 1.3)

#### Overlap and Core Signature ####
blca_deseq2_ifn_overlap <- intersect(blca_deseq2_test$up_genes, ifn_genes)
print(paste("Number of overlapping IFN genes:", length(blca_deseq2_ifn_overlap)))
print("Intersection with BRCA up_genes:")
print(intersect(blca_deseq2_test$up_genes, brca_deseq2_test$up_genes))

blca_core_scores <- calculate_sample_scores(blca_deseq2_test, custom_index = core_11_genes)
ggplot(blca_core_scores, aes(x = group, y = custom_metric, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  geom_jitter(aes(color = group), width = 0.2, size = 2.5, alpha = 0.8) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(title = "BLCA: 11-Gene Core Signature", y = "GSVA Enrichment Score", x = NULL) +
  theme(legend.position = "none", plot.title = element_text(face = "bold")) +
  stat_compare_means(method = "wilcox.test", label.x = 1.5)
```

# Predictive signature 
# 1) Model training
```{r}
model_results_list <- list()
```

## BRCA
```{r}
brca_data_counts <- prepare_and_cache_rnaseq("BRCA", candidate_sample_groups)

names(brca_data_counts)
dim(brca_data_counts$counts)
length(brca_deseq2_test$up_genes)

brca_vst_mat <- vst(brca_deseq2_test$dds, blind = TRUE)
brca_vst_counts <- assay(brca_vst_mat)
dim(brca_vst_counts)
# [1] 39234    29

# brca_vst_133 <- brca_vst_counts %>%
#   as.data.frame() %>%
#   rownames_to_column("gene_name") %>%
#   filter(gene_name %in% brca_deseq2_test$up_genes) %>%
#   column_to_rownames("gene_name")



# 1. Map the Ensembl IDs in brca_vst_counts to Symbols using existing res_df
# Using match ensures the order remains identical to the matrix columns
gene_symbols <- brca_deseq2_test$res_df$gene_name[match(rownames(brca_vst_counts), rownames(brca_deseq2_test$res_df))]
rownames(brca_vst_counts) <- gene_symbols


# 2. Subset to the 133 genes (now that names match) and clean NAs
brca_vst_133 <- brca_vst_counts[rownames(brca_vst_counts) %in% brca_deseq2_test$up_genes, ]
brca_vst_133 <- brca_vst_133[!is.na(rownames(brca_vst_133)), ]
dim(brca_vst_133)

# Align Labels and Run LOOCV______####

# 3. Setup labels (Engager = 1, Non-Engager = 0) 
# colData is already aligned to the matrix columns
y <- ifelse(colData(brca_deseq2_test$dds)$engager_group == "engager", 1, 0)
X <- t(brca_vst_133) # glmnet requires samples as rows
n_samples <- nrow(X)
brca_predicted_probs <- numeric(n_samples)

# LOOCV Loop for 133 genes____####
set.seed(42) # For reproducibility in cv.glmnet folds
for(i in 1:n_samples) {
  X_train <- X[-i, ]
  y_train <- y[-i]
  X_test  <- X[i, , drop = FALSE]
  
  # alpha=1 for LASSO; binomial for logistic
  # Using 5 folds for internal CV because training N is small (28)
  cv_fit <- cv.glmnet(X_train, y_train, family = "binomial", alpha = 1, nfolds = 5)
  
  # Predict probability for the held-out sample
  brca_predicted_probs[i] <- predict(cv_fit, X_test, s = "lambda.min", type = "response")
}

# Evaluation and Plotting_____####

library(pROC)
auc_val <- as.numeric(auc(roc(y, brca_predicted_probs, quiet = TRUE)))
cor_val <- cor(brca_predicted_probs, y)


results_df <- data.frame(
  true_label = factor(y, labels = c("Non-Engager", "Engager")),
  pred_prob = brca_predicted_probs
)

# Save results_df for brca
model_results_list$brca <- results_df

# Boxplot with pearson correlation ####
# 2. Map colors directly by position: 1st level (Non-Engager) = blue, 2nd level (Engager) = red
lasso_boxplot <- ggplot(model_results_list$brca, aes(x = true_label, y = pred_prob, fill = true_label)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.4, width = 0.5) +
  geom_jitter(aes(color = true_label), width = 0.2, size = 2.5) +
  theme_minimal() +
  scale_fill_manual(values = c("blue", "red")) +
  scale_color_manual(values = c("blue", "red")) +
  labs(
    title = paste0("BRCA LOOCV Across 133 DE Genes (AUC = ", round(auc_val, 3), ")"),
    subtitle = paste0("Pearson Correlation: ", round(cor_val, 3)),
    y = "Predicted Probability of Engager Status",
    x = NULL
  ) +
  theme(legend.position = "none", 
        plot.title = element_text(face = "bold", size = 10),
        plot.subtitle = element_text(size = 9)) +
  stat_compare_means(method = "wilcox.test", label = "p.format", label.x = 1.4)

ggsave(
  filename = "model_plots/v2/brca_lasso_boxplot.png", 
  plot = lasso_boxplot, 
  width = 5.5,      # Width in inches
  height = 5,     # Height in inches
  units = "in",
  dpi = 300       # 300 DPI is the print standard
)


# ROC Plot ####
# Calculate the ROC object
# Ensures "Engager" is treated as the positive class
roc_obj <- roc(model_results_list$brca$true_label, model_results_list$brca$pred_prob, levels = c("Non-Engager", "Engager"))

# Extract coordinates for ggplot
roc_data <- data.frame(
  FPR = 1 - roc_obj$specificities,
  TPR = roc_obj$sensitivities
)

roc_plot <- ggplot(roc_data, aes(x = FPR, y = TPR)) +
  #geom_line(color = "red", size = 1) +
  geom_path(color = "red", size = 1) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey") +
  theme_minimal() +
  labs(
    title = "ROC Curve: BRCA LOOCV Across 133 DE Genes",
    subtitle = paste0("AUC = ", round(auc(roc_obj), 3)),
    x = "False Positive Rate (1 - Specificity)",
    y = "True Positive Rate (Sensitivity)"
  ) +
  theme(plot.title = element_text(face = "bold"))

ggsave(
  filename = "model_plots/v2/brca_lasso_roc_curve.png", 
  plot = roc_plot, 
  width = 6,      # Width in inches
  height = 5,     # Height in inches
  units = "in",
  dpi = 300       # 300 DPI is the print standard
)


# PR curve ####
# Calculate the pr object
brca_pr_obj <- pr.curve(
  scores.class0 = model_results_list$brca %>% 
    filter(true_label == "Engager") %>% 
    pull(pred_prob),
  scores.class1 = model_results_list$brca %>% 
    filter(true_label == "Non-Engager") %>% 
    pull(pred_prob),
  curve = TRUE
)

pr_plot <- as.data.frame(brca_pr_obj$curve) %>% 
  setNames(c("recall", "precision", "threshold")) %>% 
  ggplot(aes(x = recall, y = precision)) +
  geom_path(color = "blue", linewidth = 1) + 
  theme_minimal() +
  labs(
    title = "Precision-Recall Curve: BRCA LOOCV Across 133 DE Genes",
    subtitle = paste0("PR AUC = ", round(brca_pr_obj$auc.integral, 3)),
    x = "Recall (True Positive Rate)",
    y = "Precision (Positive Predictive Value)"
  ) +
  theme(plot.title = element_text(face = "bold")) +
  geom_hline(
    yintercept = sum(model_results_list$brca$true_label == "Engager") / nrow(model_results_list$brca), 
    linetype = "dashed", color = "grey"
  )

ggsave(
  filename = "model_plots/v2/brca_lasso_pr_curve.png", 
  plot = pr_plot, 
  width = 6,      # Width in inches
  height = 5,     # Height in inches
  units = "in",
  dpi = 300       # 300 DPI is the print standard
)

# LOOCV Loop for Core 11 Genes____####

# 1. Define the 11-gene subset from your counts
# (Ensures rownames are already symbols from previous step)
brca_vst_11 <- brca_vst_counts[rownames(brca_vst_counts) %in% core_11_genes, ]
brca_vst_11 <- brca_vst_11[!is.na(rownames(brca_vst_11)), ]

# 2. Setup training data
X_11 <- t(brca_vst_11) 
y <- ifelse(colData(brca_deseq2_test$dds)$engager_group == "engager", 1, 0)
n_samples <- nrow(X_11)
predicted_probs_11 <- numeric(n_samples)

# 3. The LOOCV Loop
set.seed(42) 
for(i in 1:n_samples) {
  # Hide one, train on the rest
  X_train <- X_11[-i, ]
  y_train <- y[-i]
  X_test  <- X_11[i, , drop = FALSE]
  
  # Lasso Logistic Regression
  # nfolds=5 for internal stability in the small training set
  cv_fit <- cv.glmnet(X_train, y_train, family = "binomial", alpha = 1, nfolds = 5)
  
  # Get the "Confidence Score" for the held-out sample
  predicted_probs_11[i] <- predict(cv_fit, X_test, s = "lambda.min", type = "response")
}

# 4. Calculate Final Stats
library(pROC)
auc_11 <- as.numeric(auc(roc(y, predicted_probs_11, quiet = TRUE)))
cor_11 <- cor(predicted_probs_11, y)

# 5. Format and Plot
results_11_df <- data.frame(
  true_label = factor(y, labels = c("Non-Engager", "Engager")),
  pred_prob = predicted_probs_11
)

ggplot(results_11_df, aes(x = true_label, y = pred_prob, fill = true_label)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.4, width = 0.5) +
  geom_jitter(aes(color = true_label), width = 0.2, size = 3) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(
    title = paste0("BRCA LOOCV: Core 11-Gene Signature (AUC = ", round(auc_11, 3), ")"),
    subtitle = paste0("Pearson Correlation: ", round(cor_11, 3)),
    y = "Predicted Probability of Engagement",
    x = NULL
  ) +
  theme(
    legend.position = "none", 
    plot.title = element_text(face = "bold", size = 14)
  ) +
  stat_compare_means(method = "wilcox.test", label = "p.format", label.x = 1.4)

# Extract coefficients at the optimal lambda (lambda.min)
# This returns a sparse matrix of all genes and their weights
weights_matrix <- coef(cv_fit, s = "lambda.min")

# Convert to a standard data frame and filter for non-zero weights
# (LASSO forces unimportant genes to 0)
gene_weights <- data.frame(
  gene = rownames(weights_matrix),
  weight = as.numeric(weights_matrix)
) %>%
  filter(weight != 0) %>%
  arrange(desc(abs(weight)))

# View the result
print(gene_weights)
```

## COAD
COAD LOOCV: 26-Gene Signature (AUC = 0.889)
```{r}
coad_data_counts <- prepare_and_cache_rnaseq("COAD", candidate_sample_groups)

names(coad_data_counts)
dim(coad_data_counts$counts)
length(coad_deseq2_test$up_genes) # 26 up genes

coad_vst_mat <- vst(coad_deseq2_test$dds, blind = TRUE)
coad_vst_counts <- assay(coad_vst_mat)
dim(coad_vst_counts)
# [1] 33414    17


# 1. Map the Ensembl IDs in coad_vst_counts to symbols using res_df
# Using match ensures the order remains identical to the matrix columns
coad_gene_symbols <- coad_deseq2_test$res_df$gene_name[match(rownames(coad_vst_counts), rownames(coad_deseq2_test$res_df))]
rownames(coad_vst_counts) <- coad_gene_symbols


# 2. Subset to the de genes (now that names match) and clean NAs
coad_vst <- coad_vst_counts[rownames(coad_vst_counts) %in% coad_deseq2_test$up_genes, ]
coad_vst <- coad_vst[!is.na(rownames(coad_vst)), ]
dim(coad_vst)
# [1] 26 17

# Align Labels and Run LOOCV______####

# 3. Setup labels (Engager = 1, Non-Engager = 0) 
# colData is already aligned to the matrix columns
y <- ifelse(colData(coad_deseq2_test$dds)$engager_group == "engager", 1, 0)
X <- t(coad_vst) # glmnet requires samples as rows
n_samples <- nrow(X)
coad_predicted_probs <- numeric(n_samples)

# LOOCV Loop for 26 genes____####
set.seed(42) # For reproducibility in cv.glmnet folds
for(i in 1:n_samples) {
  X_train <- X[-i, ]
  y_train <- y[-i]
  X_test  <- X[i, , drop = FALSE]
  
  # alpha=1 for LASSO; binomial for logistic
  # Using 5 folds for internal CV because training N is small (28)
  cv_fit <- cv.glmnet(X_train, y_train, family = "binomial", alpha = 1, nfolds = 5)
  
  # Predict probability for the held-out sample
  coad_predicted_probs[i] <- predict(cv_fit, X_test, s = "lambda.min", type = "response")
}

# Evaluation and Plotting_____####

# Boxplot with Pearson Correlation ####
library(pROC)
auc_val <- as.numeric(auc(roc(y, coad_predicted_probs, quiet = TRUE)))
cor_val <- cor(coad_predicted_probs, y)

results_df <- data.frame(
  true_label = factor(y, labels = c("Non-Engager", "Engager")),
  pred_prob = coad_predicted_probs
)

model_results_list$coad <- results_df
rm(results_df)

lasso_boxplot <- ggplot(model_results_list$coad, aes(x = true_label, y = pred_prob, fill = true_label)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.4, width = 0.5) +
  geom_jitter(aes(color = true_label), width = 0.2, size = 2.5) +
  theme_minimal() +
  scale_fill_manual(values = c("Non-Engager" = "#4575b4", "Engager" = "#d73027")) +
  scale_color_manual(values = c("Non-Engager" = "#4575b4", "Engager" = "#d73027")) +
  labs(
    title = paste0("COAD LOOCV Across 26 DE Genes (AUC = ", round(auc_val, 3), ")"),
    subtitle = paste0("Pearson Correlation: ", round(cor_val, 3)),
    y = "Predicted Probability of Engager Status",
    x = NULL
  ) +
  theme(legend.position = "none", 
        plot.title = element_text(face = "bold", size = 10),
        plot.subtitle = element_text(size = 9)) +
  stat_compare_means(method = "wilcox.test", label = "p.format", label.x = 1.4)

ggsave(
  filename = "model_plots/v2/coad_lasso_boxplot.png", 
  plot = lasso_boxplot, 
  width = 5.5,      # Width in inches
  height = 5,     # Height in inches
  units = "in",
  dpi = 300       # 300 DPI is the print standard
)



# AUC Plot
# Calculate the ROC object
# Ensures "Engager" is treated as the positive class
roc_obj <- roc(model_results_list$coad$true_label, model_results_list$coad$pred_prob, levels = c("Non-Engager", "Engager"))

# Extract coordinates for ggplot
roc_data <- data.frame(
  FPR = 1 - roc_obj$specificities,
  TPR = roc_obj$sensitivities
)

# ROC plot ####
roc_plot <- ggplot(roc_data, aes(x = FPR, y = TPR)) +
  #geom_line(color = "red", size = 1) +
  geom_path(color = "red", size = 1) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey") +
  theme_minimal() +
  labs(
    title = "ROC Curve: COAD LOOCV Across 26 Genes",
    subtitle = paste0("AUC = ", round(auc(roc_obj), 3)),
    x = "False Positive Rate (1 - Specificity)",
    y = "True Positive Rate (Sensitivity)"
  ) +
  theme(plot.title = element_text(face = "bold"))

ggsave(
  filename = "model_plots/v2/coad_lasso_roc_curve.png", 
  plot = roc_plot, 
  width = 6,      # Width in inches
  height = 5,     # Height in inches
  units = "in",
  dpi = 300       # 300 DPI is the print standard
)

# PR Curve ####
# Calculate the pr object
coad_pr_obj <- pr.curve(
  scores.class0 = model_results_list$coad %>% 
    filter(true_label == "Engager") %>% 
    pull(pred_prob),
  scores.class1 = model_results_list$coad %>% 
    filter(true_label == "Non-Engager") %>% 
    pull(pred_prob),
  curve = TRUE
)


pr_plot <- as.data.frame(coad_pr_obj$curve) %>% 
  setNames(c("recall", "precision", "threshold")) %>% 
  ggplot(aes(x = recall, y = precision)) +
  geom_path(color = "blue", linewidth = 1) + 
  theme_minimal() +
  labs(
    title = "Precision-Recall Curve: COAD LOOCV Across 26 DE Genes",
    subtitle = paste0("PR AUC = ", round(coad_pr_obj$auc.integral, 3)),
    x = "Recall (True Positive Rate)",
    y = "Precision (Positive Predictive Value)"
  ) +
  theme(plot.title = element_text(face = "bold")) +
  geom_hline(
    yintercept = sum(model_results_list$coad$true_label == "Engager") / nrow(model_results_list$coad), 
    linetype = "dashed", color = "grey"
  )

ggsave(
  filename = "model_plots/v2/coad_lasso_pr_curve.png", 
  plot = pr_plot, 
  width = 6,      # Width in inches
  height = 5,     # Height in inches
  units = "in",
  dpi = 300       # 300 DPI is the print standard
)


```

# 2) Gene signature to full TCGA RNA cohort
## BRCA
### Pull rnaseq data
```{r}
# Load full transcriptome and clinical data
brca_full_data <- prepare_and_cache_rnaseq("BRCA", matched_only = FALSE)

# Extract the metadata from the RNA-seq object
brca_clinical_check <- as.data.frame(colData(brca_full_data$se))
# Check if the vital_status and time columns are already there
target_cols <- c("vital_status", "days_to_death", "days_to_last_follow_up")
found_cols <- target_cols[target_cols %in% colnames(brca_clinical_check)]
print(found_cols)

brca_clinical <- brca_clinical_check
rm(brca_clinical_check)


```

### vst normalize 
To make sure gene expression values are the same scale as the N=29 training set
```{r}
# 1. Create the DESeq container for the full cohort
brca_dds_full <- DESeqDataSetFromMatrix(
    countData = brca_full_data$counts,
    colData = brca_full_data$meta,
    design = ~ 1 # Intercept only for population-wide normalization
)

# 2. Run vst
# This will take ~10-15 minutes on 1,111 samples
brca_vst_full <- vst(brca_dds_full, blind = TRUE)
brca_mat_full <- assay(brca_vst_full)

# 3. Map Gene IDs to Symbols
# Ensuring the rownames match your 133-gene model (e.g., RAP2C)
rownames(brca_mat_full) <- rowData(brca_full_data$se)$gene_name[match(rownames(brca_mat_full), rowData(brca_full_data$se)$gene_id)]
```

### Get final model
train on all 29 samples now that we know that the 133 genes are useful for predicting engager/non-engager status from the training set
```{r}

# 1. Train the Master Model on all 29 Discovery Samples
brca_x_train_133 <- t(brca_vst_133) 
brca_y_labels_133 <- y              

set.seed(42)
brca_model_133_final <- cv.glmnet(
    x = brca_x_train_133, 
    y = brca_y_labels_133, 
    family = "binomial", 
    alpha = 1, 
    nfolds = 5
)

# 2. Extract the non-zero active weights for reference
brca_final_weights <- as.matrix(coef(brca_model_133_final, s = "lambda.min"))
brca_active_genes <- brca_final_weights[brca_final_weights[, 1] != 0, , drop = FALSE]
print(brca_active_genes) # Every gene has a positive coefficient!

# Save as tsv
brca_active_genes %>% 
  as.data.frame() %>% 
  rename(coefficient = 1) %>% 
  mutate(gene_symbol = rownames(.)) %>% 
  select(gene_symbol, coefficient) %>% 
  arrange(desc(abs(coefficient))) %>% 
  write_tsv("~/model_plots/v2/brca_lasso_gene_signature.tsv")

# Get genes and weights ####
brca_gene_signature_table <- brca_active_genes %>% 
  as.data.frame() %>% 
  rename(coefficient = 1) %>% 
  mutate(gene_symbol = rownames(.)) %>% 
  select(gene_symbol, coefficient) %>% 
  arrange(desc(abs(coefficient))) %>% 
  # Make table
  gt() %>%
  tab_header(
    title = "BRCA 19-Gene LASSO Signature Coefficients",
    subtitle = "Selected from 133 Candidate DE Genes (BRCA Cohort)"
  ) %>%
  cols_label(
    gene_symbol = "Gene Symbol",
    coefficient = "LASSO Coefficient (β)"
  ) %>%
  fmt_number(
    columns = coefficient,
    decimals = 4
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = gene_symbol == "(Intercept)")
  ) %>%
  tab_options(
    table.width = px(400),
    column_labels.background.color = "whitesmoke"
  ) 

brca_gene_signature_table




# 3. Align the 1,095 Validation Samples
# Subsetting the large matrix using the exact row order of the discovery matrix
brca_x_val_133 <- t(brca_mat_full[rownames(brca_vst_133), ])

# error in one of the 133 genes doesn't exist in the validation matrix (brca_mat_full). maybe an issue with gene annotations
target_genes <- rownames(brca_vst_133)
available_genes <- rownames(brca_mat_full)

missing_genes <- setdiff(target_genes, available_genes)
print(missing_genes)

# fix
# 1. Extract the genes we DO have, transpose to patients-as-rows ####
available_data <- brca_mat_full[rownames(brca_mat_full) %in% target_genes, ] %>% 
  t()

# 2. Create a zero-filled matrix for the 7 missing genes ####
missing_data <- matrix(
  0, 
  nrow = nrow(available_data), 
  ncol = length(missing_genes), 
  dimnames = list(rownames(available_data), missing_genes)
)

# 3. Bind them together and forcefully reorder columns to match the discovery set exactly ####
brca_x_val_133 <- cbind(available_data, missing_data)[, target_genes]

# Confirm dimensions are exactly 1095 rows, 133 columns
dim(brca_x_val_133)

# Clean up temporary debugging variables
rm(target_genes, available_genes, missing_genes, available_data, missing_data)

# 4. Generate the Engagement Probabilities ####
brca_final_probs_133 <- predict(
  brca_model_133_final, 
  newx = brca_x_val_133, 
  s = "lambda.min", 
  type = "response"
)

# 3. Consolidate into a clean, tidy tibble ####
brca_engagement_scores <- tibble(
  case_id = rownames(brca_final_probs_133),
  engagement_prob = as.numeric(brca_final_probs_133)
)

# 4. Sanity Check ####
summary(brca_engagement_scores$engagement_prob) # Model successfully identified extremes (near 0 and near 1) and gradient meaning the signature is present and varies in the full cohort population
```

### Histogram and density plot of engager status probabilities to see if there are multiple populations
looks like there are two populations at the extremes. will take the extreme populations and not cutoff by the median
```{r}
ggplot(brca_engagement_scores, aes(x = engagement_prob)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "steelblue", color = "black", alpha = 0.7) +
  geom_density(color = "darkred", linewidth = 1) +
  theme_minimal() +
  labs(
    title = "Distribution of Engager Status Probabilities in BRCA",
    x = "Probability of Engager Status (0 to 1)",
    y = "Density"
  )
```

## COAD
### Pull rnaseq data
```{r}
# Load full transcriptome and clinical data
coad_full_data <- prepare_and_cache_rnaseq("COAD", matched_only = FALSE)

# Extract the metadata from the RNA-seq object
coad_clinical_check <- as.data.frame(colData(coad_full_data$se))
# Check if the vital_status and time columns are already there
target_cols <- c("vital_status", "days_to_death", "days_to_last_follow_up")
found_cols <- target_cols[target_cols %in% colnames(coad_clinical_check)]
print(found_cols)

coad_clinical <- coad_clinical_check
rm(coad_clinical_check)


```

### vst normalize 
To make sure gene expression values are the same scale as the N=17 training set
```{r}
# 1. Create the DESeq container for the full cohort
coad_dds_full <- DESeqDataSetFromMatrix(
    countData = coad_full_data$counts,
    colData = coad_full_data$meta,
    design = ~ 1 # Intercept only for population-wide normalization
)

# 2. Run vst
coad_vst_full <- vst(coad_dds_full, blind = TRUE)
coad_mat_full <- assay(coad_vst_full)

# 3. Map Gene IDs to Symbols
# Ensuring the rownames match your 133-gene model (e.g., RAP2C)
rownames(coad_mat_full) <- rowData(coad_full_data$se)$gene_name[match(rownames(coad_mat_full), rowData(coad_full_data$se)$gene_id)]
```

### Get final model
train on all 17 samples now that we know that the de genes are useful for predicting engager/non-engager status from the training set
```{r}
# 1. Train the Master Model on all 17 Discovery Samples
coad_x_train <- t(coad_vst) 
coad_y_labels <- y               

set.seed(42)
coad_model_final <- cv.glmnet(
    x = coad_x_train, 
    y = coad_y_labels, 
    family = "binomial", 
    alpha = 1, 
    nfolds = 5
)

# 2. Extract the non-zero active weights for reference
coad_final_weights <- as.matrix(coef(coad_model_final, s = "lambda.min"))
coad_active_genes <- coad_final_weights[coad_final_weights[, 1] != 0, , drop = FALSE]
print(coad_active_genes) 

#              lambda.min
# (Intercept) -19.10543555
# CCDC175       0.46150323
# ZNF572        0.54155029
# FAM181B       0.08039022
# DRICH1        0.47364495
# PDPK2P        1.01272774
# EOLA1-DT      0.39995442
# CHMP1B2P      0.51472922

# Save as tsv
coad_active_genes %>% 
  as.data.frame() %>% 
  rename(coefficient = 1) %>% 
  mutate(gene_symbol = rownames(.)) %>% 
  select(gene_symbol, coefficient) %>% 
  arrange(desc(abs(coefficient))) %>% 
  write_tsv("~/model_plots/v2/coad_lasso_gene_signature.tsv")

# Get genes and weights ####
coad_gene_signature_table <- coad_active_genes %>% 
  as.data.frame() %>% 
  rename(coefficient = 1) %>% 
  mutate(gene_symbol = rownames(.)) %>% 
  select(gene_symbol, coefficient) %>% 
  arrange(desc(abs(coefficient))) %>% 
  # Make table
  gt() %>%
  tab_header(
    title = "COAD 26-Gene LASSO Signature Coefficients",
    subtitle = "Selected from 133 Candidate DE Genes (BRCA Cohort)"
  ) %>%
  cols_label(
    gene_symbol = "Gene Symbol",
    coefficient = "LASSO Coefficient (β)"
  ) %>%
  fmt_number(
    columns = coefficient,
    decimals = 4
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = gene_symbol == "(Intercept)")
  ) %>%
  tab_options(
    table.width = px(400),
    column_labels.background.color = "whitesmoke"
  ) 

coad_gene_signature_table



# 3. Align the Validation Samples
target_genes <- rownames(coad_vst)
available_genes <- rownames(coad_mat_full)

missing_genes <- setdiff(target_genes, available_genes)
print(paste("Missing genes:", paste(missing_genes, collapse=", ")))
# [1] "Missing genes: EOLA1-DT, LOC105379362"

# 1. Extract the genes we DO have, transpose to patients-as-rows ####
available_data <- coad_mat_full[rownames(coad_mat_full) %in% target_genes, ] %>% 
  t()

# 2. Create a zero-filled matrix for the missing genes (if any) ####
if(length(missing_genes) > 0) {
  missing_data <- matrix(
    0, 
    nrow = nrow(available_data), 
    ncol = length(missing_genes), 
    dimnames = list(rownames(available_data), missing_genes)
  )
  # 3. Bind them together and forcefully reorder columns to match the discovery set exactly ####
  coad_x_val <- cbind(available_data, missing_data)[, target_genes]
} else {
  coad_x_val <- available_data[, target_genes]
}

# Confirm dimensions 
dim(coad_x_val)
# [1] 458  26

# Clean up temporary debugging variables
rm(target_genes, available_genes, missing_genes, available_data)
if(exists("missing_data")) rm(missing_data)

# 4. Generate the Engagement Probabilities ####
coad_final_probs <- predict(
  coad_model_final, 
  newx = coad_x_val, 
  s = "lambda.min", 
  type = "response"
)

# 5. Consolidate into a clean, tidy tibble ####
coad_engagement_scores <- tibble(
  case_id = rownames(coad_final_probs),
  engagement_prob = as.numeric(coad_final_probs)
)

# 6. Sanity Check ####
summary(coad_engagement_scores$engagement_prob)
#> dim(coad_x_val)
#    Min.  1st Qu.   Median     Mean  3rd Qu.     Max. 
# 0.003887 0.045553 0.102098 0.160451 0.231534 0.912739 
```

### Histogram and density plot of engager status probabilities to see if there are multiple populations
looks like there are two populations at the extremes. will take the extreme populations and not cutoff by the median
```{r}
ggplot(coad_engagement_scores, aes(x = engagement_prob)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "steelblue", color = "black", alpha = 0.7) +
  geom_density(color = "darkred", linewidth = 1) +
  theme_minimal() +
  labs(
    title = "Distribution of Engager Status Probabilities in COAD",
    x = "Probability of Engager Status",
    y = "Density"
  )
```


# 3) Prognostic Analysis (survival curves, coxph)
## BRCA
### Survival curves
Get poopulations by taking the top and bottom quartiles
```{r}

# 1. Define the 25% (quartile) cutoffs
lower_cutoff <- quantile(brca_engagement_scores$engagement_prob, 0.25)
upper_cutoff <- quantile(brca_engagement_scores$engagement_prob, 0.75)

# 2. Isolate extremes, strictly filter for primary tumors, and merge
brca_survival_data <- brca_engagement_scores %>%
  filter(engagement_prob <= lower_cutoff | engagement_prob >= upper_cutoff) %>%
  mutate(
    engagement_group = if_else(engagement_prob >= upper_cutoff, "Engager", "Non-Engager"),
    patient_id = substr(case_id, 1, 12),
    sample_type = substr(case_id, 14, 15) # Extract the 2-digit sample code
  ) %>%
  # Safeguard 1: Strictly keep only Primary Solid Tumors ("01")
  filter(sample_type == "01") %>%
  # Safeguard 2: Now safely drop replicate vials (A vs B)
  distinct(patient_id, .keep_all = TRUE) %>%
  
  # Join and build survival variables
  inner_join(
    brca_clinical %>% distinct(submitter_id, .keep_all = TRUE), 
    by = c("patient_id" = "submitter_id")
  ) %>%
  mutate(
    os_time = if_else(vital_status == "Dead", days_to_death, days_to_last_follow_up),
    os_status = if_else(vital_status == "Dead", 1, 0)
  ) %>%
  filter(!is.na(os_time) & os_time > 0 & !is.na(os_status))

# 3. Verify the final dataset
dim(brca_survival_data)
#  500 101
table(brca_survival_data$engagement_group, brca_survival_data$os_status)
# Non-Engager 208  38

brca_survival_data %>%
  select(patient_id, engagement_prob, engagement_group, os_time, os_status,
         ajcc_pathologic_stage, age_at_index) %>%
  write_tsv("survival_analysis_plots/brca_quartile_km_data.tsv")


```

### Plot KM curves
```{r}
# See percentage of censored data
brca_survival_data %>%
  group_by(engagement_group) %>%
  summarize(
    total_n = n(),
    deaths = sum(os_status == 1),
    censored = sum(os_status == 0),
    percent_censored = round((censored / total_n) * 100, 1)
  )

# # A tibble: 2 × 5
#   engagement_group total_n deaths censored percent_censored
#   <chr>              <int>  <int>    <int>            <dbl>
# 1 Engager              254     41      213             83.9
# 2 Non-Engager          246     38      208             84.6


# 1. Fit the survival model using your existing variable name
brca_survival_data$engagement_group <- factor(
  brca_survival_data$engagement_group, 
  levels = c("Engager", "Non-Engager")
)
brca_km_fit <- survfit(Surv(os_time, os_status) ~ engagement_group, data = brca_survival_data)


# 2. Generate the Kaplan-Meier Plot
brca_surv_plot <- ggsurvplot(
  brca_km_fit,
  data = brca_survival_data,
  pval = TRUE,
  risk.table = TRUE,
  palette = c("#E41A1C", "#377EB8"), 
  title = "BRCA Overall Survival: Top vs Bottom Quartile",
  xlab = "Time (Days)",
  ylab = "Survival Probability",
  
  # Standardize legend position and labels
  legend = "top",
  legend.title = "Status",
  legend.labs = c("Engager", "Non-Engager"),
  
  # Granular time axis
  break.time.by = 1000,
  ggtheme = theme_minimal()
)

png("~/survival_analysis_plots/brca_survplot.png", width = 7, height = 7, units = "in", res = 300)
print(brca_surv_plot)
dev.off()

```

### CoxPH
with covariates
```{r}

# Force Non-Engager to be the reference group (baseline)
brca_survival_data$engagement_group <- relevel(factor(brca_survival_data$engagement_group), ref = "Non-Engager")

# 1. Fit a Multivariate Cox Model
# include clinical factors
brca_cox_model <- coxph(
  Surv(os_time, os_status) ~ engagement_group + ajcc_pathologic_stage + age_at_index, 
  data = brca_survival_data
)

# 2. Look at the result
summary(brca_cox_model)

# Call:
# coxph(formula = Surv(os_time, os_status) ~ engagement_group + 
#     ajcc_pathologic_stage + age_at_index, data = brca_survival_data)
# 
#   n= 473, number of events= 55 
#    (27 observations deleted due to missingness)
# 
#                                       coef  exp(coef)   se(coef)      z Pr(>|z|)    
# engagement_groupEngager          6.711e-01  1.956e+00  2.965e-01  2.264 0.023596 *  
# ajcc_pathologic_stageStage IA   -9.873e-02  9.060e-01  8.751e-01 -0.113 0.910173    
# ajcc_pathologic_stageStage IB   -1.631e+01  8.250e-08  1.045e+04 -0.002 0.998755    
# ajcc_pathologic_stageStage II    2.516e+00  1.238e+01  1.152e+00  2.185 0.028915 *  
# ajcc_pathologic_stageStage IIA  -6.014e-02  9.416e-01  5.867e-01 -0.103 0.918357    
# ajcc_pathologic_stageStage IIB   2.870e-01  1.332e+00  5.848e-01  0.491 0.623613    
# ajcc_pathologic_stageStage IIIA  7.354e-01  2.086e+00  6.038e-01  1.218 0.223199    
# ajcc_pathologic_stageStage IIIB  1.234e+00  3.434e+00  7.112e-01  1.735 0.082789 .  
# ajcc_pathologic_stageStage IIIC -1.624e+01  8.881e-08  3.274e+03 -0.005 0.996044    
# ajcc_pathologic_stageStage IV    2.613e+00  1.364e+01  6.019e-01  4.341 1.42e-05 ***
# ajcc_pathologic_stageStage X     2.155e+00  8.624e+00  1.136e+00  1.897 0.057803 .  
# age_at_index                     3.709e-02  1.038e+00  1.093e-02  3.392 0.000693 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
#                                 exp(coef) exp(-coef) lower .95 upper .95
# engagement_groupEngager         1.956e+00  5.112e-01    1.0942     3.498
# ajcc_pathologic_stageStage IA   9.060e-01  1.104e+00    0.1630     5.035
# ajcc_pathologic_stageStage IB   8.250e-08  1.212e+07    0.0000       Inf
# ajcc_pathologic_stageStage II   1.238e+01  8.079e-02    1.2953   118.280
# ajcc_pathologic_stageStage IIA  9.416e-01  1.062e+00    0.2982     2.973
# ajcc_pathologic_stageStage IIB  1.332e+00  7.505e-01    0.4235     4.192
# ajcc_pathologic_stageStage IIIA 2.086e+00  4.793e-01    0.6389     6.813
# ajcc_pathologic_stageStage IIIB 3.434e+00  2.912e-01    0.8520    13.841
# ajcc_pathologic_stageStage IIIC 8.881e-08  1.126e+07    0.0000       Inf
# ajcc_pathologic_stageStage IV   1.364e+01  7.330e-02    4.1931    44.388
# ajcc_pathologic_stageStage X    8.624e+00  1.160e-01    0.9312    79.867
# age_at_index                    1.038e+00  9.636e-01    1.0158     1.060
# 
# Concordance= 0.783  (se = 0.035 )
# Likelihood ratio test= 58.52  on 12 df,   p=4e-08
# Wald test            = 53.3  on 12 df,   p=4e-07
# Score (logrank) test = 107.4  on 12 df,   p=<2e-16

# Add subtype and hormone receptor status as covariates ####
# Subtypes_____________________
# Look for the right columns
TCGAquery_subtype(tumor = "brca") %>% View()
  colnames(.)
  
# Pull official TCGA-BRCA molecular subtypes
brca_subtypes <- TCGAquery_subtype(tumor = "brca") %>%
  dplyr::select(patient, subtype = BRCA_Subtype_PAM50)


# Hormone receptor status_________________________
# 1. Query for the Biotab (text table) version of the clinical data
query_brca_biotab <- GDCquery(
    project = "TCGA-BRCA",
    data.category = "Clinical",
    data.type = "Clinical Supplement",
    data.format = "BCR Biotab"
)

# 2. Download the small .txt files (takes seconds)
GDCdownload(query_brca_biotab)

# 3. Load the tables into a list
brca_biotab_list <- GDCprepare(query_brca_biotab)

# 4. Extract the specific patient clinical table
# This is where the hormone receptor status is stored
brca_hr_status_data <- brca_biotab_list$clinical_patient_brca

# Look for the receptor status columns
colnames(brca_hr_status_data)[grep("er_status|pr_status|her2_status", colnames(brca_hr_status_data), ignore.case = TRUE)]

# Prepare the clean HR status table
# 1. Clean the HR table by keeping only real TCGA barcodes
brca_hr_clean <- brca_hr_status_data %>%
  filter(grepl("^TCGA-", bcr_patient_barcode)) %>%
  dplyr::select(
    patient_id = bcr_patient_barcode, 
    er_status = er_status_by_ihc, 
    pr_status = pr_status_by_ihc, 
    her2_status = her2_status_by_ihc
  )

# Count how many times each patient appears
dupe_check <- brca_hr_clean %>%
  count(patient_id) %>%
  filter(n > 1)
print(dupe_check)


# Join to survival data
brca_survival_adj <- brca_survival_data %>%
  left_join(brca_hr_clean, by = "patient_id")

# Check the first few rows to see the new columns
head(brca_survival_adj)

# Check the distribution of ER status
table(brca_survival_adj$er_status, useNA = "always")
# Not Evaluated]        Negative        Positive            <NA> 
#              25             123             352               0 

# Check Progesterone Receptor status
table(brca_survival_adj$pr_status, useNA = "always")
# [Not Evaluated]   Indeterminate        Negative        Positive            <NA> 
#              26               1             179             294               0 

# Check HER2 status
table(brca_survival_adj$her2_status, useNA = "always")
# [Not Available] [Not Evaluated]       Equivocal   Indeterminate        Negative        Positive            <NA> 
#               5              86              85               7             236              81               0 

# Only get patients that have definitive positive and negative for multivariate coxph model
# Create a clean subset for the multivariate model
brca_survival_final <- brca_survival_adj %>%
  # Filter to keep only definitive clinical results for all three
  filter(
    er_status %in% c("Positive", "Negative"),
    pr_status %in% c("Positive", "Negative"),
    her2_status %in% c("Positive", "Negative")
  ) %>%
  # Convert to factors so the model treats 'Negative' as the baseline
  mutate(
    er_status = factor(er_status, levels = c("Negative", "Positive")),
    pr_status = factor(pr_status, levels = c("Negative", "Positive")),
    her2_status = factor(her2_status, levels = c("Negative", "Positive"))
  )
nrow(brca_survival_final)

# Ensure engagement_group has the correct baseline
brca_survival_final$engagement_group <- relevel(factor(brca_survival_final$engagement_group), ref = "Non-Engager")

# Run the final adjusted model
brca_cox_triple_adj <- coxph(
  Surv(os_time, os_status) ~ engagement_group + er_status + pr_status + her2_status + ajcc_pathologic_stage + age_at_index, 
  data = brca_survival_final
)

# Generate the results
summary(brca_cox_triple_adj)

# Call:
# coxph(formula = Surv(os_time, os_status) ~ engagement_group + 
#     er_status + pr_status + her2_status + ajcc_pathologic_stage + 
#     age_at_index, data = brca_survival_final)
# 
#   n= 304, number of events= 28 
#    (11 observations deleted due to missingness)
# 
#                                       coef  exp(coef)   se(coef)      z Pr(>|z|)    
# engagement_groupEngager          1.111e+00  3.038e+00  5.082e-01  2.186 0.028784 *  
# er_statusPositive               -1.410e+00  2.442e-01  7.800e-01 -1.807 0.070723 .  
# pr_statusPositive                6.139e-02  1.063e+00  7.269e-01  0.084 0.932695    
# her2_statusPositive              6.348e-02  1.066e+00  4.650e-01  0.137 0.891410    
# ajcc_pathologic_stageStage IA    6.282e-01  1.874e+00  1.256e+00  0.500 0.616966    
# ajcc_pathologic_stageStage II   -1.712e+01  3.656e-08  3.475e+04  0.000 0.999607    
# ajcc_pathologic_stageStage IIA   9.921e-02  1.104e+00  8.165e-01  0.121 0.903298    
# ajcc_pathologic_stageStage IIB   6.193e-02  1.064e+00  9.139e-01  0.068 0.945974    
# ajcc_pathologic_stageStage IIIA  1.130e+00  3.095e+00  8.599e-01  1.314 0.188869    
# ajcc_pathologic_stageStage IIIB  1.594e+00  4.925e+00  9.485e-01  1.681 0.092795 .  
# ajcc_pathologic_stageStage IIIC -1.644e+01  7.277e-08  6.089e+03 -0.003 0.997846    
# ajcc_pathologic_stageStage IV    3.071e+00  2.156e+01  9.179e-01  3.345 0.000822 ***
# ajcc_pathologic_stageStage X    -1.680e+01  5.054e-08  2.664e+04 -0.001 0.999497    
# age_at_index                     4.382e-02  1.045e+00  1.674e-02  2.617 0.008870 ** 
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
#                                 exp(coef) exp(-coef) lower .95 upper .95
# engagement_groupEngager         3.038e+00  3.292e-01   1.12198     8.226
# er_statusPositive               2.442e-01  4.095e+00   0.05295     1.126
# pr_statusPositive               1.063e+00  9.405e-01   0.25580     4.420
# her2_statusPositive             1.066e+00  9.385e-01   0.42834     2.651
# ajcc_pathologic_stageStage IA   1.874e+00  5.336e-01   0.15985    21.974
# ajcc_pathologic_stageStage II   3.656e-08  2.735e+07   0.00000       Inf
# ajcc_pathologic_stageStage IIA  1.104e+00  9.056e-01   0.22287     5.472
# ajcc_pathologic_stageStage IIB  1.064e+00  9.399e-01   0.17741     6.380
# ajcc_pathologic_stageStage IIIA 3.095e+00  3.231e-01   0.57377    16.697
# ajcc_pathologic_stageStage IIIB 4.925e+00  2.030e-01   0.76738    31.609
# ajcc_pathologic_stageStage IIIC 7.277e-08  1.374e+07   0.00000       Inf
# ajcc_pathologic_stageStage IV   2.156e+01  4.639e-02   3.56692   130.300
# ajcc_pathologic_stageStage X    5.054e-08  1.979e+07   0.00000       Inf
# age_at_index                    1.045e+00  9.571e-01   1.01106     1.080
# 
# Concordance= 0.857  (se = 0.031 )
# Likelihood ratio test= 36.51  on 14 df,   p=9e-04
# Wald test            = 29.98  on 14 df,   p=0.008
# Score (logrank) test = 57.77  on 14 df,   p=3e-07
# 

# PROBLEM 
# coef  exp(coef)   se(coef)      z Pr(>|z|)    
# ajcc_pathologic_stageStage II   -1.712e+01  3.656e-08  3.475e+04  0.000 0.999607
# the standard error for Stage II (as well as stage IIIC and stage X) is very high indicating zero events (deaths) happened in this subgroup. will collapse the groups to Stage 1,2,3,4,5

# 1. Collapse sub-stages into major stages to fix zero-event categories
brca_survival_stable <- brca_survival_final %>%
  mutate(stage_simple = case_when(
    grepl("Stage I$|Stage IA|Stage IB", ajcc_pathologic_stage) ~ "Stage_I",
    grepl("Stage II$|Stage IIA|Stage IIB", ajcc_pathologic_stage) ~ "Stage_II",
    grepl("Stage III$|Stage IIIA|Stage IIIB|Stage IIIC", ajcc_pathologic_stage) ~ "Stage_III",
    grepl("Stage IV", ajcc_pathologic_stage) ~ "Stage_IV",
    TRUE ~ NA_character_
  )) %>%
  # Filter out 'Unknown' or 'Stage X' which have no prognostic meaning
  filter(!is.na(stage_simple)) %>%
  mutate(stage_simple = factor(stage_simple, levels = c("Stage_I", "Stage_II", "Stage_III", "Stage_IV")))

# 2. Run the stable multivariate Cox model
brca_cox_stable <- coxph(
  Surv(os_time, os_status) ~ engagement_group + er_status + pr_status + her2_status + stage_simple + age_at_index, 
  data = brca_survival_stable
)

# 3. View Summary
summary(brca_cox_stable)
# Call:
# coxph(formula = Surv(os_time, os_status) ~ engagement_group + 
#     er_status + pr_status + her2_status + stage_simple + age_at_index, 
#     data = brca_survival_stable)
# 
#   n= 302, number of events= 28 
# 
#                             coef exp(coef) se(coef)      z Pr(>|z|)    
# engagement_groupEngager  1.10846   3.02969  0.49734  2.229  0.02583 *  
# er_statusPositive       -1.35035   0.25915  0.77583 -1.741  0.08177 .  
# pr_statusPositive       -0.01790   0.98226  0.72957 -0.025  0.98043    
# her2_statusPositive      0.03269   1.03323  0.45249  0.072  0.94241    
# stage_simpleStage_II    -0.07471   0.92802  0.66758 -0.112  0.91090    
# stage_simpleStage_III    0.97823   2.65974  0.69116  1.415  0.15697    
# stage_simpleStage_IV     2.92377  18.61127  0.80272  3.642  0.00027 ***
# age_at_index             0.04701   1.04813  0.01470  3.197  0.00139 ** 
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
#                         exp(coef) exp(-coef) lower .95 upper .95
# engagement_groupEngager    3.0297    0.33007   1.14304     8.030
# er_statusPositive          0.2592    3.85877   0.05664     1.186
# pr_statusPositive          0.9823    1.01806   0.23508     4.104
# her2_statusPositive        1.0332    0.96784   0.42564     2.508
# stage_simpleStage_II       0.9280    1.07757   0.25079     3.434
# stage_simpleStage_III      2.6597    0.37598   0.68631    10.308
# stage_simpleStage_IV      18.6113    0.05373   3.85921    89.754
# age_at_index               1.0481    0.95408   1.01835     1.079
# 
# Concordance= 0.829  (se = 0.036 )
# Likelihood ratio test= 33.37  on 8 df,   p=5e-05
# Wald test            = 35.31  on 8 df,   p=2e-05
# Score (logrank) test = 50.27  on 8 df,   p=4e-08
# 

# 4. Run Proportional Hazards Test (Assumption Check) to determine if the relative risk (HR) is constant over the entire follow-up period
ph_test <- cox.zph(brca_cox_stable)
print(ph_test)
#                  chisq df     p
# engagement_group  4.452  1 0.035
# er_status         0.519  1 0.471
# pr_status         0.455  1 0.500
# her2_status       1.036  1 0.309
# stage_simple      2.308  3 0.511
# age_at_index      1.662  1 0.197
# GLOBAL           17.856  8 0.022

# Visualizes how the hazard changes over time
plot(ph_test[1]) # [1] refers to the first variable: engagement_group
abline(h = 0, col = "red")
# Result: the risk associated with the engager group is not constant over time but is higher as time increases meaning has more long-term impact than short term impact
# 

```

## COAD
### Survival curves
Get poopulations by taking the top and bottom quartiles
```{r}

# 1. Define the 25% (quartile) cutoffs
lower_cutoff <- quantile(coad_engagement_scores$engagement_prob, 0.25)
upper_cutoff <- quantile(coad_engagement_scores$engagement_prob, 0.75)

# 2. Isolate extremes, strictly filter for primary tumors, and merge
coad_survival_data <- coad_engagement_scores %>%
  filter(engagement_prob <= lower_cutoff | engagement_prob >= upper_cutoff) %>%
  mutate(
    engagement_group = if_else(engagement_prob >= upper_cutoff, "Engager", "Non-Engager"),
    patient_id = substr(case_id, 1, 12),
    sample_type = substr(case_id, 14, 15) # Extract the 2-digit sample code
  ) %>%
  # Safeguard 1: Strictly keep only Primary Solid Tumors ("01")
  filter(sample_type == "01") %>%
  # Safeguard 2: Now safely drop replicate vials (A vs B)
  distinct(patient_id, .keep_all = TRUE) %>%
  
  # Join and build survival variables
  inner_join(
    coad_clinical %>% distinct(submitter_id, .keep_all = TRUE), 
    by = c("patient_id" = "submitter_id")
  ) %>%
  mutate(
    os_time = if_else(vital_status == "Dead", days_to_death, days_to_last_follow_up),
    os_status = if_else(vital_status == "Dead", 1, 0)
  ) %>%
  filter(!is.na(os_time) & os_time > 0 & !is.na(os_status))

# 3. Verify the final dataset
dim(coad_survival_data)
#  [1] 192 114

table(coad_survival_data$engagement_group, coad_survival_data$os_status)
# Non-Engager 
  #            0  1
  # Engager     71 30
  # Non-Engager 67 24

coad_survival_data %>%
  select(patient_id, engagement_prob, engagement_group, os_time, os_status,
         ajcc_pathologic_stage, age_at_index) %>%
  write_tsv("survival_analysis_plots/coad_quartile_km_data.tsv")

```

### Plot KM curves
```{r}
# See percentage of censored data
coad_survival_data %>%
  group_by(engagement_group) %>%
  summarize(
    total_n = n(),
    deaths = sum(os_status == 1),
    censored = sum(os_status == 0),
    percent_censored = round((censored / total_n) * 100, 1)
  )

# # A tibble: 2 × 5
# engagement_group total_n deaths censored percent_censored
#   <chr>              <int>  <int>    <int>            <dbl>
# 1 Engager              101     30       71             70.3
# 2 Non-Engager           91     24       67             73.6


# 1. Fit the survival model using your existing variable name
coad_survival_data$engagement_group <- factor(
  coad_survival_data$engagement_group, 
  levels = c("Engager", "Non-Engager")
)
coad_km_fit <- survfit(Surv(os_time, os_status) ~ engagement_group, data = coad_survival_data)


# 2. Generate the Kaplan-Meier Plot
coad_surv_plot <- ggsurvplot(
  coad_km_fit,
  data = coad_survival_data,
  pval = TRUE,
  risk.table = TRUE,
  palette = c("#E41A1C", "#377EB8"), 
  title = "COAD Overall Survival: Top vs Bottom Quartile",
  xlab = "Time (Days)",
  ylab = "Survival Probability",
  
  # Standardize legend position and labels
  legend = "top",
  legend.title = "Status",
  legend.labs = c("Engager", "Non-Engager"),
  
  # Granular time axis
  break.time.by = 1000,
  ggtheme = theme_minimal()
)

png("~/survival_analysis_plots/coad_survplot.png", width = 7, height = 7, units = "in", res = 300)
print(coad_surv_plot)
dev.off()
```

### CoxPH
with covariates
```{r}

# Force Non-Engager to be the reference group (baseline)
coad_survival_data$engagement_group <- relevel(factor(coad_survival_data$engagement_group), ref = "Non-Engager")

# 1. Fit a Multivariate Cox Model
# include clinical factors
coad_cox_model <- coxph(
  Surv(os_time, os_status) ~ engagement_group + ajcc_pathologic_stage + age_at_index, 
  data = coad_survival_data
)

# 2. Look at the result
summary(coad_cox_model)

# Call:
# coxph(formula = Surv(os_time, os_status) ~ engagement_group + 
#     ajcc_pathologic_stage + age_at_index, data = coad_survival_data)
# 
#   n= 172, number of events= 37 
#    (20 observations deleted due to missingness)
# 
#                                       coef  exp(coef)   se(coef)      z Pr(>|z|)    
# engagement_groupEngager         -8.464e-02  9.188e-01  3.484e-01 -0.243  0.80804    
# ajcc_pathologic_stageStage II    4.861e-01  1.626e+00  8.425e-01  0.577  0.56395    
# ajcc_pathologic_stageStage IIA   1.829e-01  1.201e+00  6.851e-01  0.267  0.78951    
# ajcc_pathologic_stageStage IIB   9.664e-01  2.629e+00  9.238e-01  1.046  0.29550    
# ajcc_pathologic_stageStage III   1.259e+00  3.520e+00  9.336e-01  1.348  0.17761    
# ajcc_pathologic_stageStage IIIA -1.507e+01  2.844e-07  4.387e+03 -0.003  0.99726    
# ajcc_pathologic_stageStage IIIB -1.399e-01  8.694e-01  8.311e-01 -0.168  0.86629    
# ajcc_pathologic_stageStage IIIC  1.228e+00  3.415e+00  7.088e-01  1.733  0.08311 .  
# ajcc_pathologic_stageStage IV    2.039e+00  7.685e+00  6.833e-01  2.985  0.00284 ** 
# ajcc_pathologic_stageStage IVA   2.554e+00  1.286e+01  9.679e-01  2.638  0.00833 ** 
# ajcc_pathologic_stageStage IVB  -1.456e+01  4.773e-07  7.280e+03 -0.002  0.99840    
# age_at_index                     8.409e-02  1.088e+00  2.047e-02  4.108 3.99e-05 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
#                                 exp(coef) exp(-coef) lower .95 upper .95
# engagement_groupEngager         9.188e-01  1.088e+00    0.4642     1.819
# ajcc_pathologic_stageStage II   1.626e+00  6.150e-01    0.3119     8.477
# ajcc_pathologic_stageStage IIA  1.201e+00  8.329e-01    0.3135     4.598
# ajcc_pathologic_stageStage IIB  2.629e+00  3.804e-01    0.4299    16.072
# ajcc_pathologic_stageStage III  3.520e+00  2.841e-01    0.5649    21.941
# ajcc_pathologic_stageStage IIIA 2.844e-07  3.516e+06    0.0000       Inf
# ajcc_pathologic_stageStage IIIB 8.694e-01  1.150e+00    0.1705     4.433
# ajcc_pathologic_stageStage IIIC 3.415e+00  2.928e-01    0.8513    13.700
# ajcc_pathologic_stageStage IV   7.685e+00  1.301e-01    2.0139    29.326
# ajcc_pathologic_stageStage IVA  1.286e+01  7.779e-02    1.9285    85.703
# ajcc_pathologic_stageStage IVB  4.773e-07  2.095e+06    0.0000       Inf
# age_at_index                    1.088e+00  9.193e-01    1.0450     1.132
# 
# Concordance= 0.776  (se = 0.043 )
# Likelihood ratio test= 43.54  on 12 df,   p=2e-05
# Wald test            = 37.56  on 12 df,   p=2e-04
# Score (logrank) test = 48.49  on 12 df,   p=3e-06

```


