---
title: "de_analysis_model_training_pkr.Rmd"
output: html_document
date: "2026-07-01"
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(results = "hide")


```


```{r packages}
install.packages("BiocManager")

# Install the necessary bioinformatics packages
# TCGAbiolinks: for data access
# GSVA: for the 94-gene signature scoring
# SummarizedExperiment: the standard data container for TCGA data
BiocManager::install(c("TCGAbiolinks", "GSVA", "SummarizedExperiment"))
BiocManager::install("IlluminaHumanMethylation450kanno.ilmn12.hg19")
BiocManager::install("sesameData")
BiocManager::install("sesame")
BiocManager::install("sesame")
BiocManager::install("edgeR")
BiocManager::install("org.Hs.eg.db")
BiocManager::install("DESeq2")
BiocManager::install("apeglm", update = FALSE, ask = FALSE)
BiocManager::install("survminer")
BiocManager::install(c("BiocGenerics", "S4Vectors", "IRanges"), force = TRUE, update = TRUE, ask = FALSE)
BiocManager::install("DESeq2", force=TRUE, update = FALSE, ask = FALSE)

install.packages("nloptr")
install.packages("ggpubr")
install.packages("cowplot")
install.packages("gt")
install.packages("minifi")
install.packages("glmnet")
install.packages("ggrepel")


# 4/15/26 trying to get DESeq2
# 1. Remove the mismatched packages from your local Linux library
remove.packages(c("BiocGenerics", "S4Vectors", "IRanges"))

# 2. Force BiocManager to use the correct repository for R 4.3
BiocManager::install(version = "3.18", ask = FALSE)

# 3. Install DESeq2 cleanly from the 3.18 repository
BiocManager::install("DESeq2", update = FALSE, ask = FALSE)

BiocManager::install(c("BiocGenerics", "IRanges", "S4Vectors", "DESeq2"), update = FALSE, ask = FALSE)

# 1. Delete the corrupted phantom folder directly from the file system
unlink("/home/nvsharm/R/x86_64-pc-linux-gnu-library/4.3/S4Vectors", recursive = TRUE)

# 2. Install S4Vectors by pointing explicitly to the Bioconductor 3.18 repository
install.packages("S4Vectors", repos = "https://bioconductor.org/packages/3.18/bioc")

install.packages("IRanges", repos = "https://bioconductor.org/packages/3.18/bioc")

# 3. Test DESeq2
library(DESeq2)




# Install clusterProfiler and msigdbr
BiocManager::install("clusterProfiler")
install.packages("msigdbr")

install.packages("PRROC")
install.packages("webshot2")

```


```{r load packages}
library(BiocManager)
library(TCGAbiolinks)
library(GSVA)
library(SummarizedExperiment)
library(tidyverse)
library(nloptr)
library(ggpubr)
library(cowplot)
library(gt)
library(webshot2)
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
library(sesameData)
library(sesame)
library(edgeR)
library(org.Hs.eg.db)
library(AnnotationDbi)
library(S4Vectors)
library(IRanges)
library(DESeq2)
library(apeglm)
#library(clusterProfiler)
library(msigdbr)
library(glmnet)
library(survminer)
library(ggrepel)
library(PRROC)
library(conflicted)
library(webshot2)
library(pROC)

conflict_prefer("select", "dplyr")
conflict_prefer("count", "dplyr")
conflict_prefer("filter", "dplyr")
conflicts_prefer(SummarizedExperiment::rowRanges)
conflicts_prefer(base::intersect)
```

```{r read objects and files}

ifn_genes <- readRDS("~/ifn_genes.rds")
pkr_sample_groups <- readr::read_tsv("~/pkr/pkr_quartile_grouping/cluster_definitions/pkr_sample_groups_quartile_grouping.tsv")
gdc_sample_metadata <- read_csv("~/gdc_sample_sheet_1.txt") 

# Reformat gdc_sample_metadata and join it to gmm_groups, then reformat so it will work with downstream functions
gdc_sample_metadata_clean <- gdc_sample_metadata %>% 
  mutate(sample = str_replace(project_id, "TCGA-", "") %>% str_replace("-", ".")) 

# Map gmm sample ids to case_id, dedup, and add engager group classification
pkr_sample_groups_clean <- pkr_sample_groups %>%
  filter(grouping_label %in% c("Engager", "Non_Engager")) %>%
  mutate(engager_group = if_else(grouping_label == "Engager", "engager", "non_engager")) %>%
  left_join(gdc_sample_metadata_clean %>% dplyr::select(sample, case_id, sample_id, sample_type),
            by = "sample") %>%
  group_by(case_id) %>%
  arrange(sample, .by_group = TRUE) %>% 
  slice_head(n = 1) %>%
  ungroup()

pkr_sample_groups_clean %>% filter(is.na(case_id)) %>% nrow()
pkr_sample_groups_clean %>% count(cancer_type, engager_group)
# A tibble: 12 × 3
#    cancer_type engager_group     n
#    <chr>       <chr>         <int>
#  1 BLCA        engager           2
#  2 BLCA        non_engager       2
#  3 BRCA        engager          16
#  4 BRCA        non_engager      15
#  5 COAD        engager           7
#  6 COAD        non_engager       9
#  7 KIRC        engager           3
#  8 KIRC        non_engager       3
#  9 KIRP        engager           9
# 10 KIRP        non_engager       9
# 11 STAD        engager           4
# 12 STAD        non_engager       4


pkr_sample_groups %>%
  filter(grouping_label %in% c("Engager", "Non_Engager")) %>%
  mutate(engager_group = if_else(grouping_label == "Engager", "engager", "non_engager")) %>%
  left_join(gdc_sample_metadata_clean %>% dplyr::select(sample, case_id), by = "sample") %>%
  group_by(case_id) %>%
  filter(n_distinct(engager_group) > 1) %>%
  ungroup()

```

# Extract raw counts for de analysis

```{r}

prepare_and_cache_rnaseq <- function(cancer_type, matched_samples = NULL, matched_only = TRUE, cache_dir = "tcga_cache_pkr") {
  
  if(!dir.exists(cache_dir)) dir.create(cache_dir)
  
  # 1. Generate unique cache paths
  suffix <- if(matched_only) "_matched.rds" else "_full.rds"
  cache_file <- file.path(cache_dir, paste0(cancer_type, suffix))
  
  # 2. Return cached data if available
  if(file.exists(cache_file)) {
    message("\n>>> Loading CACHED RNA-seq data: ", cancer_type, " (", if(matched_only) "Matched" else "Full", ")")
    return(readRDS(cache_file))
  }
  
  # 3. Download from GDC
  message("\n>>> Downloading NEW RNA-seq data for ", cancer_type)
  query <- GDCquery(
    project = paste0("TCGA-", cancer_type),
    data.category = "Transcriptome Profiling",
    data.type = "Gene Expression Quantification",
    workflow.type = "STAR - Counts",
    sample.type = "Primary Tumor"
  )
  GDCdownload(query)
  se <- GDCprepare(query)
  
  # 4. Standardize Patient IDs (12-char barcode)
  colData(se)$case_id <- substr(colData(se)$barcode, 1, 12)
  
  # 5. Execute Branching Logic
  if(matched_only) {
    # Discovery Mode: Filter to the 29 matched samples
    if(is.null(matched_samples)) stop("Error: matched_samples required when matched_only = TRUE")
    
    target_ids <- matched_samples %>%
      filter(cancer_type == .env$cancer_type) %>%
      pull(case_id)
    
    se_final <- se[, colData(se)$case_id %in% target_ids]
    se_final <- se_final[, !duplicated(colData(se_final)$case_id)]
    
    meta_df <- as.data.frame(colData(se_final)) %>%
      left_join(matched_samples %>% 
                  filter(cancer_type == .env$cancer_type) %>% 
                  dplyr::select(case_id, engager_group), 
                by = "case_id")
  } else {
    # Validation Mode: Keep all patients (~1,100 for BRCA)
    se_final <- se[, !duplicated(colData(se)$case_id)]
    meta_df <- as.data.frame(colData(se_final))
  }
  
  # 6. Save and Return
  final_data <- list(
    counts = assay(se_final, "unstranded"),
    se = se_final,
    meta = meta_df
  )
  
  saveRDS(final_data, cache_file)
  message("  Saved to cache: ", cache_file)
  return(final_data)
}


```


# DE analysis
1) engagers vs non_engagers
2) get up and down genes
3) Are any up genes in IFN signature genes?

## DESeq2
### Table of how many samples are in each cancer type for DE analysis
```{r}
pkr_sample_groups_clean %>%
  count(cancer_type, engager_group)

pkr_sample_groups_clean %>%
  count(cancer_type, engager_group) %>%
  pivot_wider(names_from = engager_group, values_from = n) %>%
  mutate(total = engager + non_engager) %>%
  gt() %>%
  cols_label(
    cancer_type = "Cancer Type",
    engager = "Engager",
    non_engager = "Non-Engager",
    total = "Total"
  ) %>%
  tab_header(title = "Sample counts by PKR engager group")
```


### Module 1 Core DESeq2 Function
```{r}

# Function handles matrix setup, variance stabilization, and DE testing 
# returns a list containing the raw dds object, the full results df, and intersection logic
run_deseq2_core <- function(cancer_type, matched_samples, ifn_genes, cache_dir = "tcga_cache_pkr") {
  
  message("\n>>> Running DESeq2 Core for ", cancer_type)
  data <- prepare_and_cache_rnaseq(cancer_type, matched_samples, cache_dir = cache_dir)
  
  # Ensure reference level
  data$meta$engager_group <- factor(data$meta$engager_group, levels = c("non_engager", "engager"))
  
  dds <- DESeqDataSetFromMatrix(countData = data$counts,
                                colData = data$meta,
                                design = ~ engager_group)
  
  keep <- rowSums(counts(dds)) >= 10
  dds <- dds[keep,]
  dds <- DESeq(dds)
  res <- lfcShrink(dds, coef="engager_group_engager_vs_non_engager", type="apeglm")
  
  res_df <- as.data.frame(res)
  res_df$ensembl_id <- rownames(res_df) %>% str_remove("\\..*")
  res_df$gene_name <- mapIds(org.Hs.eg.db,
                             keys = res_df$ensembl_id,
                             column = "SYMBOL",
                             keytype = "ENSEMBL",
                             multiVals = "first")
  
  up_genes <- res_df %>% filter(padj < 0.05, !is.na(gene_name), log2FoldChange > 0.5) %>% pull(gene_name)
  down_genes <- res_df %>% filter(padj < 0.05, !is.na(gene_name), log2FoldChange < -0.5) %>% pull(gene_name)
  
  
  return(list(
    cancer_type = cancer_type,
    dds = dds,
    res_df = res_df,
    up_genes = up_genes,
    down_genes = down_genes,
    ifn_overlap = intersect(up_genes, ifn_genes),
    n_samples = nrow(data$meta),
    n_per_group = table(data$meta$engager_group)
  ))
}

```




### Module 2 QC Plotting
Dispersion plot, Volcano plot, PCA plot
```{r}
generate_deseq_qc_plots <- function(deseq_output, output_dir = "pkr/pkr_quartile_grouping/de_qc_plots") {
  ct <- deseq_output$cancer_type
  dds <- deseq_output$dds
  
  dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)
  
  # Dispersion Plot
  #png(paste0(output_dir, ct, "_dispersion.png"), width=600, height=500)
  png(file.path(output_dir, paste0(ct, "_dispersion.png")), width=600, height=500)
  plotDispEsts(dds, main=paste(ct, "Dispersion Estimates"))
  dev.off()
  
  # PCA Plot
  vsd <- vst(dds, blind=FALSE)
  pca_data <- plotPCA(vsd, intgroup="engager_group", returnData=TRUE)
  percentVar <- round(100 * attr(pca_data, "percentVar"))
  
  pca_plot <- ggplot(pca_data, aes(PC1, PC2, color=engager_group)) +
    geom_point(size=3, alpha=0.8) +
    theme_minimal() +
    labs(title=paste(ct, "PCA"),
         x=paste0("PC1: ", percentVar[1], "%"),
         y=paste0("PC2: ", percentVar[2], "%")) +
    scale_color_manual(values=c("non_engager"="#00BFC4", "engager"="#F8766D"))
  
  #ggsave(paste0("DESeq_QC_Plots/", ct, "_PCA.png"), plot=pca_plot, width=5, height=4)
  ggsave(file.path(output_dir, paste0(ct, "_PCA.png")), plot=pca_plot, width=5, height=4)
  
  # --- NEW VOLCANO PLOT SECTION ---
  # Identify Top 20 significant genes
  top_20 <- as.data.frame(deseq_output$res_df) %>%
    filter(!is.na(padj)) %>%
    arrange(padj) %>%
    slice_head(n = 20) %>%
    pull(gene_name)
  
  # Combine with biological overlap lists
  biological_overlaps <- deseq_output$ifn_overlap
  genes_to_label <- unique(c(top_20, biological_overlaps))
  
  # Prepare data: define colors and which genes to label
  volcano_data <- as.data.frame(deseq_output$res_df) %>%
    mutate(
      significance = case_when(
        padj < 0.05 & log2FoldChange > 0.5 ~ "Up",
        padj < 0.05 & log2FoldChange < -0.5 ~ "Down",
        TRUE ~ "NS"
      ),
      label = if_else(gene_name %in% genes_to_label & padj < 0.05, gene_name, NA_character_)
    )

  volcano_plot <- ggplot(volcano_data, aes(x=log2FoldChange, y=-log10(padj), color=significance)) +
    geom_point(alpha=0.4, size=1.5) +
    scale_color_manual(values=c("Up"="#F8766D", "Down"="#00BFC4", "NS"="grey70")) +
    geom_text_repel(aes(label=label), color="black", size=3, max.overlaps = 30) +
    theme_minimal() +
    labs(title=paste(ct, "Volcano Plot"), 
         subtitle="Labeled: Top 20 Significant + IFN Signature Overlaps", 
         x="log2 Fold Change", y="-log10 Adjusted P-value")
  
  #ggsave(paste0("DESeq_QC_Plots/", ct, "_volcano.png"), plot=volcano_plot, width=7, height=6)
  ggsave(file.path(output_dir, paste0(ct, "_volcano.png")), plot=volcano_plot, width=7, height=6)
  
  # --- END VOLCANO SECTION ---
  
  message("Saved QC plots for ", ct)
}
```


### Module 3 GSVA sample-by-sample scoring
```{r}
calculate_sample_scores <- function(deseq_output, custom_index = NULL) {
  
  # Default to the cancer-specific up_genes if no index is provided
  if(is.null(custom_index)) {
    custom_index <- deseq_output$up_genes
  }
  
  if(length(custom_index) == 0) {
    message("No index genes available for ", deseq_output$cancer_type)
    return(NULL)
  }
  
  # Extract VST normalized counts
  vsd <- vst(deseq_output$dds, blind=FALSE)
  vst_mat <- assay(vsd)
  
  # Map rownames from Ensembl to Symbol for GSVA matching
  rownames(vst_mat) <- deseq_output$res_df$gene_name[match(rownames(vst_mat), rownames(deseq_output$res_df))]
  vst_mat <- vst_mat[!is.na(rownames(vst_mat)), ] # Drop any genes that didn't map correctly to a symbol (NAs)
  
  # Put the up genes into a list format for GSVA package
  index_list <- list(Target_Index = custom_index)
  
  # run GSVA to get one score per column (per sample)
  gsva_res <- gsva(expr = vst_mat, gset.idx.list = index_list, method = "gsva", verbose = FALSE)
  
  # format into a table that can be used to make a boxplot
  sample_scores <- data.frame(
    case_id = colnames(vst_mat),
    custom_metric = as.numeric(gsva_res["Target_Index", ]),
    group = deseq_output$dds$engager_group
  )
  
  return(sample_scores)
}
```

### BRCA 
```{r}
#### Module 1_____ ####
brca_deseq2_test <- run_deseq2_core("BRCA", pkr_sample_groups_clean, ifn_genes)
brca_deseq2_test$n_per_group
# non_engager     engager 
#         15          16 
length(brca_deseq2_test$up_genes)
# [1] 125
brca_deseq2_test$res_df %>% View

brca_deseq2_test$ifn_overlap
# [1] "PARP12" "IFIT2"  "CASP1"  "BATF2"  "CXCL10" "CXCL11" "USP18" 



#### Module 2______ ####
# Generate the PCA and Dispersion plots and volcano plots
generate_deseq_qc_plots(brca_deseq2_test)


#### Module 3______ ####
# Generate the custom sample-by-sample metric
brca_gsva_scores <- calculate_sample_scores(brca_deseq2_test)

# Visualize the gene score
group_colors <- c("non_engager" = "#4575b4", "engager" = "#d73027")

brca_gsva_boxplot <- ggplot(brca_gsva_scores, aes(x = group, y = custom_metric, fill = group)) +
  # Create the boxplot with a bit of transparency
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  # Add the sample points (dots) and color them by group
  geom_jitter(aes(color = group), width = 0.2, size = 2.5, alpha = 0.8) +
  # Use a clean, professional theme
  theme_minimal() +
  # Map the colors we defined
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  # Refine labels and titles
  labs(
    title = "BRCA: Sample-Level Gene Set Scoring",
    subtitle = paste0("Based on ", length(brca_deseq2_test$up_genes), "-gene Signature"),
    y = "GSVA Enrichment Score",
    x = NULL # Removing 'group' label as the axis categories make it obvious
  ) +
  # Remove the legend since the x-axis already labels the groups
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 11, color = "black"),
    panel.grid.minor = element_blank()
  ) +
  stat_compare_means(method = "wilcox.test", label.x = 1.5, label.y = 0.6)

ggsave("pkr/pkr_quartile_grouping/gsva/brca_gsva_boxplot.png", plot = brca_gsva_boxplot,
       width = 5.5, height = 5, units = "in", dpi = 300)


# 2. See which up genes overlap with ifn_genes
brca_deseq2_ifn_overlap <- intersect(brca_deseq2_test$up_genes, ifn_genes)

# 3. Check the results
print(paste("Number of overlapping IFN genes:", length(brca_deseq2_ifn_overlap)))
print(brca_deseq2_ifn_overlap)


#### MSigDB interrogation of up genes______####
# For the PKR relevant pathways need to find the exact msigdb set names. search for terms and see what's available
all_sets <- msigdbr(species = "human")
all_sets %>%
  filter(str_detect(gs_name, "EIF2|PERK|PKR|INTEGRATED_STRESS|ATF4|UNFOLDED")) %>%
  filter(gs_collection %in% c("H", "C2")) %>%  #Hallmark and curated pathways only %>% 
  distinct(gs_name, gs_collection, gs_subcollection) %>% View()
  #write_tsv(.,"~/pkr/pkr_quartile_grouping/gsva/eif2_perk_pkr_integratedstress_atf4_unfolded_related_msigdb_pathways.tsv")
  #print(n = 100)

all_sets %>%
  filter(str_detect(gs_name, "NFKB|INFLAMMATOR|TNFA")) %>%
  filter(gs_collection %in% c("H", "C2")) %>%  #Hallmark and curated pathways only
  distinct(gs_name, gs_collection, gs_subcollection) %>% View()
  #write_tsv(.,"~/pkr/pkr_quartile_grouping/gsva/nfkb_inflammatory_tnfa_related_msigdb_pathways.tsv")
  #print(n = 100)


# 1. Pull the official Hallmark immune sets
msigdb_hallmark_sets <- msigdbr(species = "human", collection = "H")
ifn_alpha_list <- msigdb_hallmark_sets[msigdb_hallmark_sets$gs_name == "HALLMARK_INTERFERON_ALPHA_RESPONSE", ]$gene_symbol
ifn_gamma_list <- msigdb_hallmark_sets[msigdb_hallmark_sets$gs_name == "HALLMARK_INTERFERON_GAMMA_RESPONSE", ]$gene_symbol
inflam_list <- msigdb_hallmark_sets[msigdb_hallmark_sets$gs_name == "HALLMARK_INFLAMMATORY_RESPONSE", ]$gene_symbol
c2_reactome <- msigdbr(species = "human", collection = "C2", subcollection = "CP:REACTOME")
ifn_sig_list  <- c2_reactome[c2_reactome$gs_name == "REACTOME_INTERFERON_SIGNALING", ]$gene_symbol

# # Pull the RIG-I / RLR-specific sets from C2
# c2_reactome <- msigdbr(species = "human", category = "C2", subcategory = "CP:REACTOME")
# c2_kegg     <- msigdbr(species = "human", category = "C2", subcategory = "CP:KEGG_LEGACY")
# rigi_ifn_list <- c2_reactome[c2_reactome$gs_name == "REACTOME_DDX58_IFIH1_MEDIATED_INDUCTION_OF_INTERFERON_ALPHA_BETA", ]$gene_symbol
# rlr_list      <- c2_kegg[c2_kegg$gs_name == "KEGG_RIG_I_LIKE_RECEPTOR_SIGNALING_PATHWAY", ]$gene_symbol
# ifn_sig_list  <- c2_reactome[c2_reactome$gs_name == "REACTOME_INTERFERON_SIGNALING", ]$gene_symbol

# Pull PKR relevant pathways
c2_all <- msigdbr(species = "human", collection = "C2")
#eif2_list          <- c2_all[c2_all$gs_name == "BIOCARTA_EIF2_PATHWAY", ]$gene_symbol        # Not relevant
#pkr_list           <- c2_all[c2_all$gs_name == "REACTOME_PKR_MEDIATED_SIGNALING", ]$gene_symbol # Not relevant
nfkb_canonical_list <- c2_all[c2_all$gs_name == "WP_CANONICAL_NFKB_PATHWAY", ]$gene_symbol
inflam_wp_list     <- c2_all[c2_all$gs_name == "WP_INFLAMMATORY_RESPONSE_PATHWAY", ]$gene_symbol
nfkb_biocarta_list <- c2_all[c2_all$gs_name == "BIOCARTA_NFKB_PATHWAY", ]$gene_symbol

# Create a list of these specific pathways
target_pathways <- list(
  HALLMARK_INTERFERON_ALPHA_RESPONSE = ifn_alpha_list,
  HALLMARK_INTERFERON_GAMMA_RESPONSE = ifn_gamma_list,
  HALLMARK_INFLAMMATORY_RESPONSE = inflam_list,
  REACTOME_INTERFERON_SIGNALING = ifn_sig_list,
  #BIOCARTA_EIF2_PATHWAY = eif2_list,
  #REACTOME_PKR_MEDIATED_SIGNALING = pkr_list,
  WP_CANONICAL_NFKB_PATHWAY = nfkb_canonical_list,
  WP_INFLAMMATORY_RESPONSE_PATHWAY = inflam_wp_list,
  BIOCARTA_NFKB_PATHWAY = nfkb_biocarta_list
)
# Check none are empty
sapply(target_pathways, length)
# HALLMARK_INTERFERON_ALPHA_RESPONSE HALLMARK_INTERFERON_GAMMA_RESPONSE     HALLMARK_INFLAMMATORY_RESPONSE      REACTOME_INTERFERON_SIGNALING          WP_CANONICAL_NFKB_PATHWAY 
#                                 97                                200                                201                                395                                  8 
#   WP_INFLAMMATORY_RESPONSE_PATHWAY              BIOCARTA_NFKB_PATHWAY 
#                                 30                                 41

# 2. Extract VST normalized counts and convert Ensembl IDs to Gene Symbols
vst_brca <- vst(brca_deseq2_test$dds, blind=FALSE)
brca_vst_mat <- assay(vst_brca)
rownames(brca_vst_mat) <- brca_deseq2_test$res_df$gene_name[match(rownames(brca_vst_mat), rownames(brca_deseq2_test$res_df))]
brca_vst_mat <- brca_vst_mat[!is.na(rownames(brca_vst_mat)), ]

# 3. Score the BRCA samples against ONLY these immune pathways
#gsva_par <- gsvaParam(exprData = brca_vst_mat, geneSets = target_pathways)
brca_immune_gsva <- gsva(expr = brca_vst_mat,
                         gset.idx.list = target_pathways,
                         method = "gsva",
                         verbose = FALSE)
#brca_immune_gsva <- gsva(gsva_par, verbose = FALSE)

# 4. Format all three pathways into a dataframe
brca_immune_scores <- data.frame(
  case_id = colnames(brca_vst_mat),
  HALLMARK_INTERFERON_ALPHA_RESPONSE = as.numeric(brca_immune_gsva["HALLMARK_INTERFERON_ALPHA_RESPONSE", ]),
  HALLMARK_INTERFERON_GAMMA_RESPONSE = as.numeric(brca_immune_gsva["HALLMARK_INTERFERON_GAMMA_RESPONSE", ]),
  HALLMARK_INFLAMMATORY_RESPONSE = as.numeric(brca_immune_gsva["HALLMARK_INFLAMMATORY_RESPONSE", ]),
  REACTOME_INTERFERON_SIGNALING = as.numeric(brca_immune_gsva["REACTOME_INTERFERON_SIGNALING", ]),
  #BIOCARTA_EIF2_PATHWAY = as.numeric(brca_immune_gsva["BIOCARTA_EIF2_PATHWAY", ]),
  #REACTOME_PKR_MEDIATED_SIGNALING = as.numeric(brca_immune_gsva["REACTOME_PKR_MEDIATED_SIGNALING", ]),
  WP_CANONICAL_NFKB_PATHWAY = as.numeric(brca_immune_gsva["WP_CANONICAL_NFKB_PATHWAY", ]),
  WP_INFLAMMATORY_RESPONSE_PATHWAY = as.numeric(brca_immune_gsva["WP_INFLAMMATORY_RESPONSE_PATHWAY", ]),
  BIOCARTA_NFKB_PATHWAY = as.numeric(brca_immune_gsva["BIOCARTA_NFKB_PATHWAY", ]),
  group = brca_deseq2_test$dds$engager_group,
  check.names = FALSE
)

# Reshape data for ggplot faceting
brca_immune_long <- brca_immune_scores %>%
  pivot_longer(cols = c(HALLMARK_INTERFERON_ALPHA_RESPONSE, HALLMARK_INTERFERON_GAMMA_RESPONSE,
                        HALLMARK_INFLAMMATORY_RESPONSE, REACTOME_INTERFERON_SIGNALING,
                        #BIOCARTA_EIF2_PATHWAY, REACTOME_PKR_MEDIATED_SIGNALING,
                        WP_CANONICAL_NFKB_PATHWAY, WP_INFLAMMATORY_RESPONSE_PATHWAY,
                        BIOCARTA_NFKB_PATHWAY),
               names_to = "pathway",
               values_to = "score")

# 5. Plot all pathways side-by-side
brca_msigdb_hallmark_pkr_pathways_boxplot <- ggplot(brca_immune_long, aes(x = group, y = score, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  geom_jitter(aes(color = group), width = 0.2, size = 2, alpha = 0.8) +
  facet_wrap(~ pathway, labeller = label_wrap_gen(width = 22)) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(
    title = "BRCA", 
    y = "GSVA Enrichment Score", 
    x = NULL
  ) +
  theme(
    legend.position = "none", 
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 15, color = "black"),
    strip.text = element_text(size = 16, face = "bold"),
    panel.grid.minor = element_blank()
  ) +
  stat_compare_means(method = "wilcox.test", label = "p.format", label.x = 1.3, size = 5)

ggsave("pkr/pkr_quartile_grouping/gsva/brca_msigdb_hallmark_pkr_pathways_boxplot.png", plot = brca_msigdb_hallmark_pkr_pathways_boxplot,
       width = 22, height = 14, units = "in", dpi = 300)
brca_msigdb_hallmark_pkr_pathways_boxplot




```


### COAD 
```{r}
#### Module 1_____ ####
coad_deseq2_test <- run_deseq2_core("COAD", pkr_sample_groups_clean, ifn_genes)
coad_deseq2_test$n_per_group
# non_engager     engager 
#          9           7 
length(coad_deseq2_test$up_genes)
# 45
coad_deseq2_test$res_df %>% View

coad_deseq2_test$ifn_overlap
# character(0)



#### Module 2______ ####
# Generate the PCA and Dispersion plots and volcano plots
generate_deseq_qc_plots(coad_deseq2_test)


#### Module 3______ ####
# Generate the custom sample-by-sample metric
coad_gsva_scores <- calculate_sample_scores(coad_deseq2_test)

# Visualize the gene score
group_colors <- c("non_engager" = "#4575b4", "engager" = "#d73027")

coad_gsva_boxplot <- ggplot(coad_gsva_scores, aes(x = group, y = custom_metric, fill = group)) +
  # Create the boxplot with a bit of transparency
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  # Add the sample points (dots) and color them by group
  geom_jitter(aes(color = group), width = 0.2, size = 2.5, alpha = 0.8) +
  # Use a clean, professional theme
  theme_minimal() +
  # Map the colors we defined
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  # Refine labels and titles
  labs(
    title = "COAD: Sample-Level Gene Set Scoring",
    subtitle = paste0("Based on ", length(coad_deseq2_test$up_genes), "-gene Signature"),
    y = "GSVA Enrichment Score",
    x = NULL # Removing 'group' label as the axis categories make it obvious
  ) +
  # Remove the legend since the x-axis already labels the groups
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 11, color = "black"),
    panel.grid.minor = element_blank()
  ) +
  stat_compare_means(method = "wilcox.test", label.x = 1.5, label.y = 0.6)

ggsave("pkr/pkr_quartile_grouping/gsva/coad_gsva_boxplot.png", plot = coad_gsva_boxplot,
       width = 5.5, height = 5, units = "in", dpi = 300)


# 2. See which up genes overlap with ifn_genes
coad_deseq2_ifn_overlap <- intersect(coad_deseq2_test$up_genes, ifn_genes)

# 3. Check the results
print(paste("Number of overlapping IFN genes:", length(coad_deseq2_ifn_overlap)))
print(coad_deseq2_ifn_overlap)


#### MSigDB interrogation of up genes______####
# For the PKR relevant pathways need to find the exact msigdb set names. search for terms and see what's available
all_sets <- msigdbr(species = "human")
all_sets %>%
  filter(str_detect(gs_name, "EIF2|PERK|PKR|INTEGRATED_STRESS|ATF4|UNFOLDED")) %>%
  filter(gs_collection %in% c("H", "C2")) %>%  #Hallmark and curated pathways only %>% 
  distinct(gs_name, gs_collection, gs_subcollection) %>% View()
  #write_tsv(.,"~/pkr/pkr_quartile_grouping/gsva/eif2_perk_pkr_integratedstress_atf4_unfolded_related_msigdb_pathways.tsv")
  #print(n = 100)

all_sets %>%
  filter(str_detect(gs_name, "NFKB|INFLAMMATOR|TNFA")) %>%
  filter(gs_collection %in% c("H", "C2")) %>%  #Hallmark and curated pathways only
  distinct(gs_name, gs_collection, gs_subcollection) %>% View()
  #write_tsv(.,"~/pkr/pkr_quartile_grouping/gsva/nfkb_inflammatory_tnfa_related_msigdb_pathways.tsv")
  #print(n = 100)


# 1. Pull the official Hallmark immune sets
msigdb_hallmark_sets <- msigdbr(species = "human", collection = "H")
ifn_alpha_list <- msigdb_hallmark_sets[msigdb_hallmark_sets$gs_name == "HALLMARK_INTERFERON_ALPHA_RESPONSE", ]$gene_symbol
ifn_gamma_list <- msigdb_hallmark_sets[msigdb_hallmark_sets$gs_name == "HALLMARK_INTERFERON_GAMMA_RESPONSE", ]$gene_symbol
inflam_list <- msigdb_hallmark_sets[msigdb_hallmark_sets$gs_name == "HALLMARK_INFLAMMATORY_RESPONSE", ]$gene_symbol
c2_reactome <- msigdbr(species = "human", collection = "C2", subcollection = "CP:REACTOME")
ifn_sig_list  <- c2_reactome[c2_reactome$gs_name == "REACTOME_INTERFERON_SIGNALING", ]$gene_symbol

# # Pull the RIG-I / RLR-specific sets from C2
# c2_reactome <- msigdbr(species = "human", category = "C2", subcategory = "CP:REACTOME")
# c2_kegg     <- msigdbr(species = "human", category = "C2", subcategory = "CP:KEGG_LEGACY")
# rigi_ifn_list <- c2_reactome[c2_reactome$gs_name == "REACTOME_DDX58_IFIH1_MEDIATED_INDUCTION_OF_INTERFERON_ALPHA_BETA", ]$gene_symbol
# rlr_list      <- c2_kegg[c2_kegg$gs_name == "KEGG_RIG_I_LIKE_RECEPTOR_SIGNALING_PATHWAY", ]$gene_symbol
# ifn_sig_list  <- c2_reactome[c2_reactome$gs_name == "REACTOME_INTERFERON_SIGNALING", ]$gene_symbol

# Pull PKR relevant pathways
c2_all <- msigdbr(species = "human", collection = "C2")
#eif2_list          <- c2_all[c2_all$gs_name == "BIOCARTA_EIF2_PATHWAY", ]$gene_symbol        # Not relevant
#pkr_list           <- c2_all[c2_all$gs_name == "REACTOME_PKR_MEDIATED_SIGNALING", ]$gene_symbol # Not relevant
nfkb_canonical_list <- c2_all[c2_all$gs_name == "WP_CANONICAL_NFKB_PATHWAY", ]$gene_symbol
inflam_wp_list     <- c2_all[c2_all$gs_name == "WP_INFLAMMATORY_RESPONSE_PATHWAY", ]$gene_symbol
nfkb_biocarta_list <- c2_all[c2_all$gs_name == "BIOCARTA_NFKB_PATHWAY", ]$gene_symbol

# Create a list of these specific pathways
target_pathways <- list(
  HALLMARK_INTERFERON_ALPHA_RESPONSE = ifn_alpha_list,
  HALLMARK_INTERFERON_GAMMA_RESPONSE = ifn_gamma_list,
  HALLMARK_INFLAMMATORY_RESPONSE = inflam_list,
  REACTOME_INTERFERON_SIGNALING = ifn_sig_list,
  #BIOCARTA_EIF2_PATHWAY = eif2_list,
  #REACTOME_PKR_MEDIATED_SIGNALING = pkr_list,
  WP_CANONICAL_NFKB_PATHWAY = nfkb_canonical_list,
  WP_INFLAMMATORY_RESPONSE_PATHWAY = inflam_wp_list,
  BIOCARTA_NFKB_PATHWAY = nfkb_biocarta_list
)
# Check none are empty
sapply(target_pathways, length)
# HALLMARK_INTERFERON_ALPHA_RESPONSE HALLMARK_INTERFERON_GAMMA_RESPONSE     HALLMARK_INFLAMMATORY_RESPONSE      REACTOME_INTERFERON_SIGNALING          WP_CANONICAL_NFKB_PATHWAY 
#                                 97                                200                                201                                395                                  8 
#   WP_INFLAMMATORY_RESPONSE_PATHWAY              BIOCARTA_NFKB_PATHWAY 
#                                 30                                 41

# 2. Extract VST normalized counts and convert Ensembl IDs to Gene Symbols
vst_coad <- vst(coad_deseq2_test$dds, blind=FALSE)
coad_vst_mat <- assay(vst_coad)
rownames(coad_vst_mat) <- coad_deseq2_test$res_df$gene_name[match(rownames(coad_vst_mat), rownames(coad_deseq2_test$res_df))]
coad_vst_mat <- coad_vst_mat[!is.na(rownames(coad_vst_mat)), ]

# 3. Score the COAD samples against ONLY these immune pathways
coad_immune_gsva <- gsva(expr = coad_vst_mat,
                         gset.idx.list = target_pathways,
                         method = "gsva",
                         verbose = FALSE)


# 4. Format all three pathways into a dataframe
coad_immune_scores <- data.frame(
  case_id = colnames(coad_vst_mat),
  HALLMARK_INTERFERON_ALPHA_RESPONSE = as.numeric(coad_immune_gsva["HALLMARK_INTERFERON_ALPHA_RESPONSE", ]),
  HALLMARK_INTERFERON_GAMMA_RESPONSE = as.numeric(coad_immune_gsva["HALLMARK_INTERFERON_GAMMA_RESPONSE", ]),
  HALLMARK_INFLAMMATORY_RESPONSE = as.numeric(coad_immune_gsva["HALLMARK_INFLAMMATORY_RESPONSE", ]),
  REACTOME_INTERFERON_SIGNALING = as.numeric(coad_immune_gsva["REACTOME_INTERFERON_SIGNALING", ]),
  WP_CANONICAL_NFKB_PATHWAY = as.numeric(coad_immune_gsva["WP_CANONICAL_NFKB_PATHWAY", ]),
  WP_INFLAMMATORY_RESPONSE_PATHWAY = as.numeric(coad_immune_gsva["WP_INFLAMMATORY_RESPONSE_PATHWAY", ]),
  BIOCARTA_NFKB_PATHWAY = as.numeric(coad_immune_gsva["BIOCARTA_NFKB_PATHWAY", ]),
  group = coad_deseq2_test$dds$engager_group,
  check.names = FALSE
)

# Reshape data for ggplot faceting
coad_immune_long <- coad_immune_scores %>%
  pivot_longer(cols = c(HALLMARK_INTERFERON_ALPHA_RESPONSE, HALLMARK_INTERFERON_GAMMA_RESPONSE,
                        HALLMARK_INFLAMMATORY_RESPONSE, REACTOME_INTERFERON_SIGNALING,
                        #BIOCARTA_EIF2_PATHWAY, REACTOME_PKR_MEDIATED_SIGNALING,
                        WP_CANONICAL_NFKB_PATHWAY, WP_INFLAMMATORY_RESPONSE_PATHWAY,
                        BIOCARTA_NFKB_PATHWAY),
               names_to = "pathway",
               values_to = "score")

# 5. Plot all pathways side-by-side
coad_msigdb_hallmark_pkr_pathways_boxplot <- ggplot(coad_immune_long, aes(x = group, y = score, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  geom_jitter(aes(color = group), width = 0.2, size = 2, alpha = 0.8) +
  facet_wrap(~ pathway, labeller = label_wrap_gen(width = 22)) +
  theme_minimal() +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  labs(
    title = "COAD", 
    y = "GSVA Enrichment Score", 
    x = NULL
  ) +
  theme(
    legend.position = "none", 
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 15, color = "black"),
    strip.text = element_text(size = 16, face = "bold"),
    panel.grid.minor = element_blank()
  ) +
  stat_compare_means(method = "wilcox.test", label = "p.format", label.x = 1.3, size = 5)

ggsave("pkr/pkr_quartile_grouping/gsva/coad_msigdb_hallmark_pkr_pathways_boxplot.png", plot = coad_msigdb_hallmark_pkr_pathways_boxplot,
       width = 22, height = 14, units = "in", dpi = 300)
coad_msigdb_hallmark_pkr_pathways_boxplot




```


# Predictive signature 
# 1) Model training
```{r}
model_results_list <- list()
```

## BRCA
```{r}

length(brca_deseq2_test$up_genes)
# 125

brca_vst_transform <- vst(brca_deseq2_test$dds, blind = TRUE)
brca_vst_counts <- assay(brca_vst_transform)
dim(brca_vst_counts)
# [1] 39657    31 - 39657 genes and 31 samples




# 1. Map the Ensembl IDs in brca_vst_counts to Symbols using existing res_df
# Using match ensures the order remains identical to the matrix columns
gene_symbols <- brca_deseq2_test$res_df$gene_name[match(rownames(brca_vst_counts), rownames(brca_deseq2_test$res_df))]
rownames(brca_vst_counts) <- gene_symbols


# 2. Subset to the up genes (now that names match) and clean NAs
brca_vst_sig_de_genes <- brca_vst_counts[rownames(brca_vst_counts) %in% brca_deseq2_test$up_genes, ]
brca_vst_sig_de_genes <- brca_vst_sig_de_genes[!is.na(rownames(brca_vst_sig_de_genes)), ]
dim(brca_vst_sig_de_genes)
# [1] 125  31

# Align Labels and Run LOOCV______####

# 3. Setup labels (Engager = 1, Non-Engager = 0) 
# colData is already aligned to the matrix columns
y <- ifelse(colData(brca_deseq2_test$dds)$engager_group == "engager", 1, 0)
X <- t(brca_vst_sig_de_genes) # glmnet requires samples as rows
n_samples <- nrow(X)
brca_predicted_probs <- numeric(n_samples)

# LOOCV Loop for up genes____####
set.seed(42) # For reproducibility in cv.glmnet folds
#set.seed(7) # Trying different seeds to test robustness
for(i in 1:n_samples) {
  X_train <- X[-i, ]
  y_train <- y[-i]
  X_test  <- X[i, , drop = FALSE]
  
  # alpha=1 for LASSO; binomial for logistic
  # Using 5 folds for internal CV because training N is small 
  cv_fit <- cv.glmnet(X_train, y_train, family = "binomial", alpha = 1, nfolds = 5)
  
  # Predict probability for the held-out sample
  brca_predicted_probs[i] <- predict(cv_fit, X_test, s = "lambda.min", type = "response")
}

# Changing up seeds to check for robustness of AUC and pearson corr values since the groups are imbalanced (see notes)
seeds <- 1:50
brca_seed_results <- map_dfr(seeds, function(s) {
  set.seed(s)
  probs <- numeric(n_samples)
  ok <- TRUE
  for (i in 1:n_samples) {
    fit <- tryCatch(
      cv.glmnet(X[-i, ], y[-i], family = "binomial", alpha = 1, nfolds = 5),
      error = function(e) NULL
    )
    if (is.null(fit)) { ok <- FALSE; break }
    probs[i] <- predict(fit, X[i, , drop = FALSE], s = "lambda.min", type = "response")
  }
  if (!ok) return(NULL)
  tibble(seed = s,
         auc = as.numeric(auc(roc(y, probs, quiet = TRUE))),
         pearson = cor(probs, y))
})

brca_seed_results %>% summarise(
  auc_mean = mean(auc), auc_sd = sd(auc), auc_min = min(auc), auc_max = max(auc),
  r_mean = mean(pearson), r_sd = sd(pearson)
)

# A tibble: 1 × 6
#   auc_mean auc_sd auc_min auc_max r_mean   r_sd
#      <dbl>  <dbl>   <dbl>   <dbl>  <dbl>  <dbl>
# 1    0.879 0.0210   0.825   0.933  0.669 0.0271

# Evaluation and Plotting_____####


auc_val <- as.numeric(auc(roc(y, brca_predicted_probs, quiet = TRUE)))
cor_val <- cor(brca_predicted_probs, y)


results_df <- data.frame(
  true_label = factor(y, labels = c("Non-Engager", "Engager")),
  pred_prob = brca_predicted_probs
)

# Save results_df for brca
model_results_list$brca <- results_df

# Boxplot with pearson correlation ####
# 2. Map colors directly by position: 1st level (Non-Engager) = blue, 2nd level (Engager) = red
lasso_boxplot <- ggplot(model_results_list$brca, aes(x = true_label, y = pred_prob, fill = true_label)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.4, width = 0.5) +
  geom_jitter(aes(color = true_label), width = 0.2, size = 2.5) +
  theme_minimal() +
  scale_fill_manual(values = c("blue", "red")) +
  scale_color_manual(values = c("blue", "red")) +
  labs(
    title = paste0("BRCA LOOCV Across ",length(brca_deseq2_test$up_genes), " DE Genes (AUC = ", round(auc_val, 3), ")"),
    subtitle = paste0("Pearson Correlation: ", round(cor_val, 3)),
    y = "Predicted Probability of Engager Status",
    x = NULL
  ) +
  theme(legend.position = "none", 
        plot.title = element_text(face = "bold", size = 10),
        plot.subtitle = element_text(size = 9)) +
  stat_compare_means(method = "wilcox.test", label = "p.format", label.x = 1.4)

ggsave(
  filename = "pkr/pkr_quartile_grouping/model_plots/brca_lasso_boxplot_quartile.png", 
  plot = lasso_boxplot, 
  width = 5.5,      # Width in inches
  height = 5,     # Height in inches
  units = "in",
  dpi = 300       # 300 DPI is the print standard
)


# ROC Plot ####
# Calculate the ROC object
# Ensures "Engager" is treated as the positive class
roc_obj <- roc(model_results_list$brca$true_label, model_results_list$brca$pred_prob, levels = c("Non-Engager", "Engager"))

# Extract coordinates for ggplot
roc_data <- data.frame(
  FPR = 1 - roc_obj$specificities,
  TPR = roc_obj$sensitivities
)

roc_plot <- ggplot(roc_data, aes(x = FPR, y = TPR)) +
  #geom_line(color = "red", size = 1) +
  geom_path(color = "red", size = 1) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey") +
  theme_minimal() +
  labs(
    title = paste0("ROC Curve: BRCA LOOCV Across ", length(brca_deseq2_test$up_genes), " DE Genes"),
    subtitle = paste0("AUC = ", round(auc(roc_obj), 3)),
    x = "False Positive Rate",
    y = "True Positive Rate"
  ) +
  theme(plot.title = element_text(face = "bold"))

ggsave(
  filename = "pkr/pkr_quartile_grouping/model_plots/brca_lasso_roc_curve_quartile.png", 
  plot = roc_plot, 
  width = 6,      # Width in inches
  height = 5,     # Height in inches
  units = "in",
  dpi = 300       # 300 DPI is the print standard
)


# PR curve ####
# Calculate the pr object
brca_pr_obj <- pr.curve(
  scores.class0 = model_results_list$brca %>% 
    filter(true_label == "Engager") %>% 
    pull(pred_prob),
  scores.class1 = model_results_list$brca %>% 
    filter(true_label == "Non-Engager") %>% 
    pull(pred_prob),
  curve = TRUE
)

pr_plot <- as.data.frame(brca_pr_obj$curve) %>% 
  setNames(c("recall", "precision", "threshold")) %>% 
  ggplot(aes(x = recall, y = precision)) +
  geom_path(color = "blue", linewidth = 1) + 
  theme_minimal() +
  labs(
    title = paste0("Precision-Recall Curve: BRCA LOOCV Across ",length(brca_deseq2_test$up_genes), " DE Genes"),
    subtitle = paste0("PR AUC = ", round(brca_pr_obj$auc.integral, 3)),
    x = "Recall (True Positive Rate)",
    y = "Precision (Positive Predictive Value)"
  ) +
  theme(plot.title = element_text(face = "bold")) +
  geom_hline(
    yintercept = sum(model_results_list$brca$true_label == "Engager") / nrow(model_results_list$brca), 
    linetype = "dashed", color = "grey"
  )

ggsave(
  filename = "pkr/pkr_quartile_grouping/model_plots/brca_lasso_pr_curve_quartile.png", 
  plot = pr_plot, 
  width = 7,      # Width in inches
  height = 5,     # Height in inches
  units = "in",
  dpi = 300       # 300 DPI is the print standard
)


```

## COAD
```{r}

length(coad_deseq2_test$up_genes) # 45 up genes

coad_vst_transform <- vst(coad_deseq2_test$dds, blind = TRUE)
coad_vst_counts <- assay(coad_vst_transform)
dim(coad_vst_counts)
# [1] 36728    32


# 1. Map the Ensembl IDs in coad_vst_counts to symbols using res_df
# Using match ensures the order remains identical to the matrix columns
coad_gene_symbols <- coad_deseq2_test$res_df$gene_name[match(rownames(coad_vst_counts), rownames(coad_deseq2_test$res_df))]
rownames(coad_vst_counts) <- coad_gene_symbols


# 2. Subset to the de genes (now that names match) and clean NAs
coad_vst_sig_de_genes <- coad_vst_counts[rownames(coad_vst_counts) %in% coad_deseq2_test$up_genes, ]
coad_vst_sig_de_genes <- coad_vst_sig_de_genes[!is.na(rownames(coad_vst_sig_de_genes)), ]
dim(coad_vst_sig_de_genes)
# [1] 45 16

# Align Labels and Run LOOCV______####

# 3. Setup labels (Engager = 1, Non-Engager = 0) 
# colData is already aligned to the matrix columns
y <- ifelse(colData(coad_deseq2_test$dds)$engager_group == "engager", 1, 0)
X <- t(coad_vst_sig_de_genes) # glmnet requires samples as rows
n_samples <- nrow(X)
coad_predicted_probs <- numeric(n_samples)

# LOOCV Loop for sig de genes____####
set.seed(7) # For reproducibility in cv.glmnet folds
for(i in 1:n_samples) {
  X_train <- X[-i, ]
  y_train <- y[-i]
  X_test  <- X[i, , drop = FALSE]
  
  # alpha=1 for LASSO; binomial for logistic
  # Using 5 folds for internal CV because training N is small (28)
  cv_fit <- cv.glmnet(X_train, y_train, family = "binomial", alpha = 1, nfolds = 5)
  
  # Predict probability for the held-out sample
  coad_predicted_probs[i] <- predict(cv_fit, X_test, s = "lambda.min", type = "response")
}

seeds <- 1:50
coad_seed_results <- map_dfr(seeds, function(s) {
  set.seed(s)
  probs <- numeric(n_samples)
  ok <- TRUE
  for (i in 1:n_samples) {
    fit <- tryCatch(
      cv.glmnet(X[-i, ], y[-i], family = "binomial", alpha = 1, nfolds = 5),
      error = function(e) NULL
    )
    if (is.null(fit)) { ok <- FALSE; break }
    probs[i] <- predict(fit, X[i, , drop = FALSE], s = "lambda.min", type = "response")
  }
  if (!ok) return(NULL)
  tibble(seed = s, auc = as.numeric(auc(roc(y, probs, quiet = TRUE))), pearson = cor(probs, y))
})

coad_seed_results %>% summarise(
  auc_mean = mean(auc), auc_sd = sd(auc), auc_min = min(auc), auc_max = max(auc),
  r_mean = mean(pearson), r_sd = sd(pearson)
)
# A tibble: 1 × 6
#   auc_mean  auc_sd auc_min auc_max r_mean   r_sd
#      <dbl>   <dbl>   <dbl>   <dbl>  <dbl>  <dbl>
# 1    0.985 0.00381   0.984       1  0.857 0.0120

# Evaluation and Plotting_____####

# Boxplot with Pearson Correlation ####
auc_val <- as.numeric(auc(roc(y, coad_predicted_probs, quiet = TRUE)))
cor_val <- cor(coad_predicted_probs, y)

results_df <- data.frame(
  true_label = factor(y, labels = c("Non-Engager", "Engager")),
  pred_prob = coad_predicted_probs
)

model_results_list$coad <- results_df
rm(results_df)

lasso_boxplot <- ggplot(model_results_list$coad, aes(x = true_label, y = pred_prob, fill = true_label)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.4, width = 0.5) +
  geom_jitter(aes(color = true_label), width = 0.2, size = 2.5) +
  theme_minimal() +
  scale_fill_manual(values = c("Non-Engager" = "#4575b4", "Engager" = "#d73027")) +
  scale_color_manual(values = c("Non-Engager" = "#4575b4", "Engager" = "#d73027")) +
  labs(
    title = paste0("COAD LOOCV Across ", length(coad_deseq2_test$up_genes), " DE Genes (AUC = ", round(auc_val, 3), ")"),
    subtitle = paste0("Pearson Correlation: ", round(cor_val, 3)),
    y = "Predicted Probability of Engager Status",
    x = NULL
  ) +
  theme(legend.position = "none", 
        plot.title = element_text(face = "bold", size = 10),
        plot.subtitle = element_text(size = 9)) +
  stat_compare_means(method = "wilcox.test", label = "p.format", label.x = 1.4)

ggsave(
  filename = "pkr/pkr_quartile_grouping/model_plots/coad_lasso_boxplot.png", 
  plot = lasso_boxplot, 
  width = 5.5,      # Width in inches
  height = 5,     # Height in inches
  units = "in",
  dpi = 300       # 300 DPI is the print standard
)



# ROC curve
# Calculate the ROC object
# Ensures "Engager" is treated as the positive class
roc_obj <- roc(model_results_list$coad$true_label, model_results_list$coad$pred_prob, levels = c("Non-Engager", "Engager"))

# Extract coordinates for ggplot
roc_data <- data.frame(
  FPR = 1 - roc_obj$specificities,
  TPR = roc_obj$sensitivities
)

# ROC plot ####
roc_plot <- ggplot(roc_data, aes(x = FPR, y = TPR)) +
  #geom_line(color = "red", size = 1) +
  geom_path(color = "red", size = 1) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey") +
  theme_minimal() +
  labs(
    title = paste0("ROC Curve: COAD LOOCV Across ", length(coad_deseq2_test$up_genes), " Genes"),
    subtitle = paste0("AUC = ", round(auc(roc_obj), 3)),
    x = "False Positive Rate",
    y = "True Positive Rate"
  ) +
  theme(plot.title = element_text(face = "bold"))

ggsave(
  filename = "pkr/pkr_quartile_grouping/model_plots/coad_lasso_roc_curve.png", 
  plot = roc_plot, 
  width = 6,      # Width in inches
  height = 5,     # Height in inches
  units = "in",
  dpi = 300       # 300 DPI is the print standard
)

# PR Curve ####
# Calculate the pr object
coad_pr_obj <- pr.curve(
  scores.class0 = model_results_list$coad %>% 
    filter(true_label == "Engager") %>% 
    pull(pred_prob),
  scores.class1 = model_results_list$coad %>% 
    filter(true_label == "Non-Engager") %>% 
    pull(pred_prob),
  curve = TRUE
)


pr_plot <- as.data.frame(coad_pr_obj$curve) %>% 
  setNames(c("recall", "precision", "threshold")) %>% 
  ggplot(aes(x = recall, y = precision)) +
  geom_path(color = "blue", linewidth = 1) + 
  theme_minimal() +
  labs(
    title = paste0("Precision-Recall Curve: COAD LOOCV Across ", length(coad_deseq2_test$up_genes), " DE Genes"),
    subtitle = paste0("PR AUC = ", round(coad_pr_obj$auc.integral, 3)),
    x = "Recall (True Positive Rate)",
    y = "Precision (Positive Predictive Value)"
  ) +
  theme(plot.title = element_text(face = "bold")) +
  geom_hline(
    yintercept = sum(model_results_list$coad$true_label == "Engager") / nrow(model_results_list$coad), 
    linetype = "dashed", color = "grey"
  )

ggsave(
  filename = "pkr/pkr_quartile_grouping/model_plots/coad_lasso_pr_curve.png", 
  plot = pr_plot, 
  width = 7,      # Width in inches
  height = 5,     # Height in inches
  units = "in",
  dpi = 300       # 300 DPI is the print standard
)


```

# 2) Gene signature to full TCGA RNA cohort
## BRCA
### Pull rnaseq data
```{r}
# Load full transcriptome and clinical data
brca_full_data <- prepare_and_cache_rnaseq("BRCA", matched_only = FALSE)

# Extract the metadata from the RNA-seq object
brca_clinical_check <- as.data.frame(colData(brca_full_data$se))
# Check if the vital_status and time columns are already there
target_cols <- c("vital_status", "days_to_death", "days_to_last_follow_up")
found_cols <- target_cols[target_cols %in% colnames(brca_clinical_check)]
print(found_cols)

brca_clinical <- brca_clinical_check
rm(brca_clinical_check)


```

### vst normalize 
To make sure gene expression values are the same scale as the training set
```{r}
# 1. Create the DESeq container for the full cohort
brca_dds_full <- DESeqDataSetFromMatrix(
    countData = brca_full_data$counts,
    colData = brca_full_data$meta,
    design = ~ 1 # Intercept only for population-wide normalization
)

# 2. Run vst
# This will take ~10-15 minutes on 1,111 samples
brca_vst_full <- vst(brca_dds_full, blind = TRUE)
brca_mat_full <- assay(brca_vst_full)

# 3. Map Gene IDs to Symbols
# Ensuring the rownames match 104-gene model (e.g., RAP2C)
rownames(brca_mat_full) <- rowData(brca_full_data$se)$gene_name[match(rownames(brca_mat_full), rowData(brca_full_data$se)$gene_id)]
```

### Get final model
train on all samples now that we know that the genes are useful for predicting engager/non-engager status from the training set
```{r}

# 1. Train the Master Model on all 30 Discovery Samples
y <- ifelse(colData(brca_deseq2_test$dds)$engager_group == "engager", 1, 0)
brca_x_train_sig_de_genes <- t(brca_vst_sig_de_genes) 
brca_y_labels_sig_de_genes <- y              

set.seed(42)
brca_model_sig_de_genes_final <- cv.glmnet(
    x = brca_x_train_sig_de_genes, 
    y = brca_y_labels_sig_de_genes, 
    family = "binomial", 
    alpha = 1, 
    nfolds = 5
)

# 2. Extract the non-zero active weights for reference
brca_final_weights <- as.matrix(coef(brca_model_sig_de_genes_final, s = "lambda.min"))
brca_active_genes <- brca_final_weights[brca_final_weights[, 1] != 0, , drop = FALSE]
print(brca_active_genes) # Every gene has a positive coefficient!

# Save as tsv
brca_active_genes %>% 
  as.data.frame() %>% 
  rename(coefficient = 1) %>% 
  mutate(gene_symbol = rownames(.)) %>% 
  select(gene_symbol, coefficient) %>% 
  arrange(desc(abs(coefficient))) %>% 
  write_tsv("~/pkr/pkr_quartile_grouping/model_plots/brca_lasso_gene_signature_quartile.tsv")

# 1. Get genes and weights ####
brca_gene_signature_table <- brca_active_genes %>% 
  as.data.frame() %>% 
  rename(coefficient = 1) %>% 
  mutate(gene_symbol = rownames(.)) %>% 
  select(gene_symbol, coefficient) %>% 
  arrange(desc(abs(coefficient))) %>% 
  # Make table
  gt() %>%
  tab_header(
    title = paste0("BRCA LASSO ", nrow(brca_active_genes)-1  ," Signature Coefficients"),
    subtitle = paste0("Selected from", length(brca_deseq2_test$up_genes) ," Candidate DE Genes")
  ) %>%
  cols_label(
    gene_symbol = "Gene Symbol",
    coefficient = "LASSO Coefficient (β)"
  ) %>%
  fmt_number(
    columns = coefficient,
    decimals = 4
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = gene_symbol == "(Intercept)")
  ) %>%
  tab_options(
    table.width = px(400),
    column_labels.background.color = "whitesmoke"
  ) 

brca_gene_signature_table
# Save as html manually via export

# 2. Create a zero-filled matrix for the missing genes (if any) ####
#brca_x_val_sig_de_genes <- t(brca_mat_full[rownames(brca_vst_sig_de_genes), ])

# error in line commented out above one of the up genes don't exist in the validation matrix (brca_mat_full). maybe an issue with gene annotations will check and fix
target_genes <- rownames(brca_vst_sig_de_genes)
available_genes <- rownames(brca_mat_full)

missing_genes <- setdiff(target_genes, available_genes)
print(missing_genes) 
# [1] "LOC100129540" "LOC112267871" "LOC100129203" "LOC100506444"

# LOC100506444 has largest weight and LOC112267871 is in the gene signature, see if i can find it in the full cohort by ensembl id
# get ensembl id to the gene symbols in training samples set
lost_ids <- brca_deseq2_test$res_df %>% 
  filter(gene_name %in% c("LOC112267871", "LOC100506444")) %>% 
  pull(ensembl_id)

lost_ids %in% (rowData(brca_full_data$se)$gene_id %>% str_remove("\\..*"))
# Both ensembl ids are found in the full cohort, will need to use ensembl ids to get the model signature genes for the full cohort matrix

# The columns are the 125 up genes the model was trained on
# Only the 15 with non-zero lasso coefficients (signature genes) affect the score,
# but predict() function requires all 125 up_gene columns present and in training order. That's why they are zero filled

# Get the ensembl id for each up gene, from the training data 
brca_upgene_symbols <- rownames(brca_vst_sig_de_genes)
brca_upgene_id_map <- brca_deseq2_test$res_df %>%
  filter(gene_name %in% brca_upgene_symbols) %>%
  distinct(gene_name, ensembl_id)

# Confirm each up_gene symbol maps to exactly one ensembl id (should return 0 rows)
brca_upgene_id_map %>% count(gene_name) %>% filter(n > 1)

# Copy the full-cohort matrix but label rows by Ensembl ID instead of symbol.
brca_mat_full_by_ensid <- assay(brca_vst_full)
rownames(brca_mat_full_by_ensid) <- rowData(brca_full_data$se)$gene_id %>% str_remove("\\..*")

# Pull each up gene from the full cohort by ensembl id, then relabel rows back to symbols
#    so the column names match the model's coefficient names.
brca_upgene_ids <- brca_upgene_id_map$ensembl_id[match(brca_upgene_symbols, brca_upgene_id_map$gene_name)]
brca_upgene_present <- brca_upgene_ids %in% rownames(brca_mat_full_by_ensid)
brca_x_val_upgenes <- t(brca_mat_full_by_ensid[brca_upgene_ids[brca_upgene_present], , drop = FALSE])
colnames(brca_x_val_upgenes) <- brca_upgene_symbols[brca_upgene_present]

# Zero-fill any up genes still not found by ensembl id. this is harmless for non-model signature up genes
#    (zero coeff), but confirm none of these are model signature genes.
brca_upgenes_missing <- brca_upgene_symbols[!brca_upgene_present]
if (length(brca_upgenes_missing) > 0) {
  brca_upgene_zerofill <- matrix(
    0,
    nrow = nrow(brca_x_val_upgenes), ncol = length(brca_upgenes_missing),
    dimnames = list(rownames(brca_x_val_upgenes), brca_upgenes_missing)
  )
  brca_x_val_upgenes <- cbind(brca_x_val_upgenes, brca_upgene_zerofill)
}

# Reorder columns to the exact training order (predict requires matching column order)
brca_x_val_sig_de_genes <- brca_x_val_upgenes[, brca_upgene_symbols]
dim(brca_x_val_sig_de_genes)
# [1] 1095  125


# 3. Generate the Engagement Probabilities ####
# score every patient in the full cohort with the trained signature, output each patient's sample probability of being an engager vs non_engager
brca_final_probs_sig_de_genes <- predict(
  brca_model_sig_de_genes_final, 
  newx = brca_x_val_sig_de_genes, 
  s = "lambda.min", 
  type = "response"
)

# 4. Consolidate into a tibble ####
brca_engagement_scores <- tibble(
  case_id = rownames(brca_final_probs_sig_de_genes),
  engagement_prob = as.numeric(brca_final_probs_sig_de_genes)
)

# 5. Sanity Check ####
# does  Model successfully identified extremes (near 0 and near 1)
summary(brca_engagement_scores$engagement_prob) 

brca_engagement_scores %>%
  mutate(bin = cut(engagement_prob, breaks = seq(0, 1, 0.1), include.lowest = TRUE)) %>%
  count(bin)

# # A tibble: 10 × 2
#    bin           n
#    <fct>     <int>
#  1 [0,0.1]     393
#  2 (0.1,0.2]   118
#  3 (0.2,0.3]    94
#  4 (0.3,0.4]    70
#  5 (0.4,0.5]    62
#  6 (0.5,0.6]    43
#  7 (0.6,0.7]    51
#  8 (0.7,0.8]    46
#  9 (0.8,0.9]    85
# 10 (0.9,1]     133

brca_engagement_scores %>%
  summarise(
    n_total = n(),
    n_above_0.5 = sum(engagement_prob > 0.5),
    n_above_0.25 = sum(engagement_prob > 0.25),
    n_above_0.1 = sum(engagement_prob > 0.1)
  )

# # A tibble: 1 × 4
#   n_total n_above_0.5 n_above_0.25 n_above_0.1
#     <int>       <int>        <int>       <int>
# 1    1095         358          537         702

# the probabilities are spread out. will probably take the upper and lower quartiles 

```

### Histogram and density plot of engager status probabilities to see if there are multiple populations
```{r}
ggplot(brca_engagement_scores, aes(x = engagement_prob)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "steelblue", color = "black", alpha = 0.7) +
  geom_density(color = "darkred", linewidth = 1) +
  theme_minimal() +
  labs(
    title = "Distribution of Engager Status Probabilities in BRCA",
    x = "Probability of Engager Status (0 to 1)",
    y = "Density"
  )

```

## COAD
### Pull rnaseq data
```{r}
# Load full transcriptome and clinical data
coad_full_data <- prepare_and_cache_rnaseq("COAD", matched_only = FALSE)

# Extract the metadata from the RNA-seq object
coad_clinical_check <- as.data.frame(colData(coad_full_data$se))
# Check if the vital_status and time columns are already there
target_cols <- c("vital_status", "days_to_death", "days_to_last_follow_up")
found_cols <- target_cols[target_cols %in% colnames(coad_clinical_check)]
print(found_cols)

coad_clinical <- coad_clinical_check
rm(coad_clinical_check)


```

### vst normalize 
To make sure gene expression values are the same scale as training set
```{r}
# 1. Create the DESeq container for the full cohort
coad_dds_full <- DESeqDataSetFromMatrix(
    countData = coad_full_data$counts,
    colData = coad_full_data$meta,
    design = ~ 1 # Intercept only for population-wide normalization
)

# 2. Run vst
coad_vst_full <- vst(coad_dds_full, blind = TRUE)
coad_mat_full <- assay(coad_vst_full)

# 3. Map Gene IDs to Symbols
# Ensuring the rownames match 57 sig de genes 
rownames(coad_mat_full) <- rowData(coad_full_data$se)$gene_name[match(rownames(coad_mat_full), rowData(coad_full_data$se)$gene_id)]
```

### Get final model
train on all 17 samples now that we know that the de genes are useful for predicting engager/non-engager status from the training set
```{r}
# 1. Train the Master Model on all 32 Discovery Samples
y <- ifelse(colData(coad_deseq2_test$dds)$engager_group == "engager", 1, 0)
length(y)
coad_x_train_sig_de_genes <- t(coad_vst_sig_de_genes) 
coad_y_labels <- y               

set.seed(42)
coad_model_sig_de_genes_final <- cv.glmnet(
    x = coad_x_train_sig_de_genes, 
    y = coad_y_labels, 
    family = "binomial", 
    alpha = 1, 
    nfolds = 5
)

# 2. Extract the non-zero active weights for reference
coad_final_weights <- as.matrix(coef(coad_model_sig_de_genes_final, s = "lambda.min"))
coad_active_genes <- coad_final_weights[coad_final_weights[, 1] != 0, , drop = FALSE]
print(coad_active_genes) 

#                lambda.min
# (Intercept)  -67.45528071
# JAG1           0.75766863
# PPARD          3.33776101
# CHIT1          0.28325655
# DSC3           0.22773432
# GPR137C        0.90397350
# CCK            0.09435162
# CYP2F1         0.64480312
# CHURC1         0.43098845
# LOC102723670   1.93224712

# Save as tsv
coad_active_genes %>% 
  as.data.frame() %>% 
  rename(coefficient = 1) %>% 
  mutate(gene_symbol = rownames(.)) %>% 
  select(gene_symbol, coefficient) %>% 
  arrange(desc(abs(coefficient))) %>% 
  write_tsv("pkr/pkr_quartile_grouping/model_plots/coad_lasso_gene_signature_quartile.tsv")

# 1. Get genes and weights ####
coad_gene_signature_table <- coad_active_genes %>% 
  as.data.frame() %>% 
  rename(coefficient = 1) %>% 
  mutate(gene_symbol = rownames(.)) %>% 
  select(gene_symbol, coefficient) %>% 
  arrange(desc(abs(coefficient))) %>% 
  # Make table
  gt() %>%
  tab_header(
    title = paste0("COAD ", nrow(coad_active_genes)-1,"-Gene LASSO Signature Coefficients"),
    subtitle = paste0("Selected from ",length(coad_deseq2_test$up_genes), " Candidate DE Genes (COAD Cohort)")
  ) %>%
  cols_label(
    gene_symbol = "Gene Symbol",
    coefficient = "LASSO Coefficient (β)"
  ) %>%
  fmt_number(
    columns = coefficient,
    decimals = 4
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = gene_symbol == "(Intercept)")
  ) %>%
  tab_options(
    table.width = px(400),
    column_labels.background.color = "whitesmoke"
  ) 

coad_gene_signature_table


# 2. Create a zero-filled matrix for the missing genes (if any) ####
target_genes <- rownames(coad_vst_sig_de_genes)
available_genes <- rownames(coad_mat_full)

missing_genes <- setdiff(target_genes, available_genes)
print(paste("Missing genes:", paste(missing_genes, collapse=", ")))
# [1] "Missing genes: POU2AF2, GPC1-AS1, LOC651337, LOC102723670"
intersect(missing_genes, rownames(coad_active_genes))
# [1] "LOC102723670" 

# Will try to recover this gene since it's coefficient is 1.93
lost_ids <- coad_deseq2_test$res_df %>%
  filter(gene_name == "LOC102723670") %>%
  pull(ensembl_id)
lost_ids %in% (rowData(coad_full_data$se)$gene_id %>% str_remove("\\..*"))

# The columns are the 45 up genes the model was trained on
# Only the 9 with non-zero lasso coefficients (signature genes) affect the score,
# but predict() function requires all 45 up_gene columns present and in training order. That's why they are zero filled
# Get the ensembl id for each up gene, from the training data 
coad_upgene_symbols <- rownames(coad_vst_sig_de_genes)
coad_upgene_id_map <- coad_deseq2_test$res_df %>%
  filter(gene_name %in% coad_upgene_symbols) %>%
  distinct(gene_name, ensembl_id)

# Confirm each up_gene symbol maps to exactly one ensembl id (should return 0 rows)
coad_upgene_id_map %>% count(gene_name) %>% filter(n > 1)

# Copy the full-cohort matrix but label rows by Ensembl ID instead of symbol.
coad_mat_full_by_ensid <- assay(coad_vst_full)
rownames(coad_mat_full_by_ensid) <- rowData(coad_full_data$se)$gene_id %>% str_remove("\\..*")

# Pull each up gene from the full cohort by ensembl id, then relabel rows back to symbols
#    so the column names match the model's coefficient names.
coad_upgene_ids <- coad_upgene_id_map$ensembl_id[match(coad_upgene_symbols, coad_upgene_id_map$gene_name)]
coad_upgene_present <- coad_upgene_ids %in% rownames(coad_mat_full_by_ensid)
coad_x_val_upgenes <- t(coad_mat_full_by_ensid[coad_upgene_ids[coad_upgene_present], , drop = FALSE])
colnames(coad_x_val_upgenes) <- coad_upgene_symbols[coad_upgene_present]

# Zero-fill any up genes still not found by ensembl id. this is harmless for non-model signature up genes
#    (zero coeff), but confirm none of these are model signature genes.
coad_upgenes_missing <- coad_upgene_symbols[!coad_upgene_present]
if (length(coad_upgenes_missing) > 0) {
  coad_upgene_zerofill <- matrix(
    0,
    nrow = nrow(coad_x_val_upgenes), ncol = length(coad_upgenes_missing),
    dimnames = list(rownames(coad_x_val_upgenes), coad_upgenes_missing)
  )
  coad_x_val_upgenes <- cbind(coad_x_val_upgenes, coad_upgene_zerofill)
}

# Reorder columns to the exact training order (predict requires matching column order)
coad_x_val_sig_de_genes <- coad_x_val_upgenes[, coad_upgene_symbols]
dim(coad_x_val_sig_de_genes)

intersect(coad_upgenes_missing, rownames(coad_active_genes))   # should be character(0)

# 4. Generate the Engagement Probabilities ####
coad_final_probs_sig_de_genes <- predict(
  coad_model_sig_de_genes_final, 
  newx = coad_x_val_sig_de_genes, 
  s = "lambda.min", 
  type = "response"
)

# 5. Consolidate into a tidy tibble ####
coad_engagement_scores <- tibble(
  case_id = rownames(coad_final_probs_sig_de_genes),
  engagement_prob = as.numeric(coad_final_probs_sig_de_genes)
)

# 6. Sanity Check ####
summary(coad_engagement_scores$engagement_prob)
#    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
# 0.00209 0.67503 0.92021 0.79484 0.98814 1.00000 

# Looks like it's skewed so that the majority of samples are assigned as engagers. might have to do with small training set (7 engagers) and strong model (e.g. high AUC).
# The opositive coefficients are also relatively large so modest expression may push patients to over 0.5

# To see full shape
coad_engagement_scores %>%
  mutate(bin = cut(engagement_prob, breaks = seq(0, 1, 0.1), include.lowest = TRUE)) %>%
  count(bin)
# A tibble: 10 × 2
#    bin           n
#    <fct>     <int>
#  1 [0,0.1]      11
#  2 (0.1,0.2]    14
#  3 (0.2,0.3]    10
#  4 (0.3,0.4]    18
#  5 (0.4,0.5]    18
#  6 (0.5,0.6]    21
#  7 (0.6,0.7]    29
#  8 (0.7,0.8]    35
#  9 (0.8,0.9]    57
# 10 (0.9,1]     245

coad_engagement_scores %>%
  summarise(n_total = n(),
            n_above_0.5 = sum(engagement_prob > 0.5),
            n_above_0.25 = sum(engagement_prob > 0.25))

# A tibble: 1 × 3
#   n_total n_above_0.5 n_above_0.25
#     <int>       <int>        <int>
# 1     458         387          430

```

### Histogram and density plot of engager status probabilities to see if there are multiple populations
looks like there are two populations at the extremes. will take the extreme populations and not cutoff by the median
```{r}
ggplot(coad_engagement_scores, aes(x = engagement_prob)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "steelblue", color = "black", alpha = 0.7) +
  geom_density(color = "darkred", linewidth = 1) +
  theme_minimal() +
  labs(
    title = "Distribution of Engager Status Probabilities in COAD",
    x = "Probability of Engager Status",
    y = "Density"
  )




```


# 3) Prognostic Analysis (survival curves, coxph)
## BRCA
### Survival curves
Get populations by using a threshold for KM curves but note that the coxph model will use the continuous probabilities (refer to notes)
```{r}

# 1. Define the 25% (quartile) cutoffs
lower_cutoff <- quantile(brca_engagement_scores$engagement_prob, 0.25)
upper_cutoff <- quantile(brca_engagement_scores$engagement_prob, 0.75)

# 2. Isolate extremes, strictly filter for primary tumors, and merge
brca_survival_data <- brca_engagement_scores %>%
  filter(engagement_prob <= lower_cutoff | engagement_prob >= upper_cutoff) %>%
  mutate(
    engagement_group = if_else(engagement_prob >= upper_cutoff, "Engager", "Non-Engager"),
    patient_id = substr(case_id, 1, 12),
    sample_type = substr(case_id, 14, 15) # Extract the 2-digit sample code
  ) %>%
  # Safeguard 1: keep only Primary Solid Tumors ("01")
  filter(sample_type == "01") %>%
  # Safeguard 2: drop replicate vials (A vs B)
  distinct(patient_id, .keep_all = TRUE) %>%
  
  # Join and build survival variables
  inner_join(
    brca_clinical %>% distinct(submitter_id, .keep_all = TRUE), 
    by = c("patient_id" = "submitter_id")
  ) %>%
  mutate(
    os_time = if_else(vital_status == "Dead", days_to_death, days_to_last_follow_up),
    os_status = if_else(vital_status == "Dead", 1, 0)
  ) %>%
  filter(!is.na(os_time) & os_time > 0 & !is.na(os_status))

# 3. Verify the final dataset
dim(brca_survival_data)
#  [1] 508 100 - 508 patients with survival data

table(brca_survival_data$engagement_group)

    # Engager Non-Engager 
    #     250         258



```

### Plot KM curves
cut at quartiles for the KM curves plot (coxph will be continuous and done separately)
```{r}

# See percentage of censored data
brca_survival_data %>%
  group_by(engagement_group) %>%
  summarize(
    total_n = n(),
    deaths = sum(os_status == 1),
    censored = sum(os_status == 0),
    percent_censored = round((censored / total_n) * 100, 1)
  )

# # A tibble: 2 × 5
#   engagement_group total_n deaths censored percent_censored
#   <fct>              <int>  <int>    <int>            <dbl>
# 1 Engager              250     36      214             85.6
# 2 Non-Engager          258     40      218             84.5


# 1. Fit the survival model using your existing variable name
brca_survival_data$engagement_group <- factor(
  brca_survival_data$engagement_group, 
  levels = c("Engager", "Non-Engager")
)
brca_km_fit <- survfit(Surv(os_time, os_status) ~ engagement_group, data = brca_survival_data)


# 2. Generate the Kaplan-Meier Plot
brca_surv_plot <- ggsurvplot(
  brca_km_fit,
  data = brca_survival_data,
  pval = TRUE,
  risk.table = TRUE,
  palette = c("#E41A1C", "#377EB8"), 
  title = "BRCA Overall Survival: Top vs Bottom Quartile",
  xlab = "Time (Days)",
  ylab = "Survival Probability",
  
  # legend position and labels
  legend = "top",
  legend.title = "Status",
  legend.labs = c("Engager", "Non-Engager"),
  
  # Granular time axis
  break.time.by = 1000,
  ggtheme = theme_minimal()
)

png("pkr/pkr_quartile_grouping/survival/brca_survplot_quartile.png", width = 7, height = 7, units = "in", res = 300)
print(brca_surv_plot)
dev.off()

```

### CoxPH
#### Continuous cox
```{r}
brca_cox_model <- coxph(
  Surv(os_time, os_status) ~ engagement_prob + ajcc_pathologic_stage + age_at_index,
  data = brca_survival_data
)
summary(brca_cox_model)
# Call:
# coxph(formula = Surv(os_time, os_status) ~ engagement_prob + 
#     ajcc_pathologic_stage + age_at_index, data = brca_survival_data)
# 
#   n= 483, number of events= 55 
#    (25 observations deleted due to missingness)
# 
#                                       coef  exp(coef)   se(coef)      z Pr(>|z|)    
# engagement_prob                 -1.965e-01  8.216e-01  3.439e-01 -0.571 0.567777    
# ajcc_pathologic_stageStage IA   -1.206e+00  2.994e-01  1.078e+00 -1.119 0.263284    
# ajcc_pathologic_stageStage IB   -1.411e+01  7.449e-07  3.159e+03 -0.004 0.996436    
# ajcc_pathologic_stageStage II    7.795e-01  2.180e+00  1.110e+00  0.702 0.482706    
# ajcc_pathologic_stageStage IIA   2.955e-01  1.344e+00  4.702e-01  0.628 0.529737    
# ajcc_pathologic_stageStage IIB   4.588e-01  1.582e+00  4.868e-01  0.942 0.345951    
# ajcc_pathologic_stageStage IIIA  9.205e-01  2.511e+00  5.139e-01  1.791 0.073260 .  
# ajcc_pathologic_stageStage IIIB  6.001e-01  1.822e+00  7.129e-01  0.842 0.399862    
# ajcc_pathologic_stageStage IIIC  1.214e+00  3.368e+00  8.202e-01  1.481 0.138735    
# ajcc_pathologic_stageStage IV    3.208e+00  2.472e+01  6.235e-01  5.145 2.68e-07 ***
# ajcc_pathologic_stageStage X     3.042e-01  1.355e+00  1.092e+00  0.278 0.780680    
# age_at_index                     4.444e-02  1.045e+00  1.168e-02  3.805 0.000142 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
#                                 exp(coef) exp(-coef) lower .95 upper .95
# engagement_prob                 8.216e-01  1.217e+00   0.41878     1.612
# ajcc_pathologic_stageStage IA   2.994e-01  3.340e+00   0.03619     2.477
# ajcc_pathologic_stageStage IB   7.449e-07  1.342e+06   0.00000       Inf
# ajcc_pathologic_stageStage II   2.180e+00  4.586e-01   0.24734    19.221
# ajcc_pathologic_stageStage IIA  1.344e+00  7.442e-01   0.53467     3.377
# ajcc_pathologic_stageStage IIB  1.582e+00  6.320e-01   0.60938     4.108
# ajcc_pathologic_stageStage IIIA 2.511e+00  3.983e-01   0.91693     6.874
# ajcc_pathologic_stageStage IIIB 1.822e+00  5.487e-01   0.45066     7.369
# ajcc_pathologic_stageStage IIIC 3.368e+00  2.969e-01   0.67485    16.811
# ajcc_pathologic_stageStage IV   2.472e+01  4.045e-02   7.28411    83.925
# ajcc_pathologic_stageStage X    1.355e+00  7.377e-01   0.15931    11.533
# age_at_index                    1.045e+00  9.565e-01   1.02178     1.070
# 
# Concordance= 0.772  (se = 0.033 )
# Likelihood ratio test= 45.92  on 12 df,   p=7e-06
# Wald test            = 61.11  on 12 df,   p=1e-08
# Score (logrank) test = 108.8  on 12 df,   p=<2e-16

brca_survival_data$engagement_group <- relevel(
  factor(brca_survival_data$engagement_group), ref = "Non-Engager"
)

brca_cox_grouped <- coxph(
  Surv(os_time, os_status) ~ engagement_group + ajcc_pathologic_stage + age_at_index,
  data = brca_survival_data
)
summary(brca_cox_grouped)
# Call:
# coxph(formula = Surv(os_time, os_status) ~ engagement_group + 
#     ajcc_pathologic_stage + age_at_index, data = brca_survival_data)
# 
#   n= 483, number of events= 55 
#    (25 observations deleted due to missingness)
# 
#                                       coef  exp(coef)   se(coef)      z Pr(>|z|)    
# engagement_groupEngager         -1.432e-01  8.666e-01  2.993e-01 -0.478 0.632386    
# ajcc_pathologic_stageStage IA   -1.198e+00  3.018e-01  1.078e+00 -1.111 0.266663    
# ajcc_pathologic_stageStage IB   -1.410e+01  7.523e-07  3.140e+03 -0.004 0.996417    
# ajcc_pathologic_stageStage II    7.869e-01  2.197e+00  1.112e+00  0.708 0.479213    
# ajcc_pathologic_stageStage IIA   2.921e-01  1.339e+00  4.704e-01  0.621 0.534532    
# ajcc_pathologic_stageStage IIB   4.600e-01  1.584e+00  4.872e-01  0.944 0.345051    
# ajcc_pathologic_stageStage IIIA  9.214e-01  2.513e+00  5.141e-01  1.792 0.073097 .  
# ajcc_pathologic_stageStage IIIB  6.086e-01  1.838e+00  7.138e-01  0.853 0.393909    
# ajcc_pathologic_stageStage IIIC  1.215e+00  3.369e+00  8.204e-01  1.481 0.138736    
# ajcc_pathologic_stageStage IV    3.202e+00  2.459e+01  6.249e-01  5.124 2.98e-07 ***
# ajcc_pathologic_stageStage X     3.085e-01  1.361e+00  1.094e+00  0.282 0.777982    
# age_at_index                     4.459e-02  1.046e+00  1.167e-02  3.819 0.000134 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
#                                 exp(coef) exp(-coef) lower .95 upper .95
# engagement_groupEngager         8.666e-01  1.154e+00   0.48199     1.558
# ajcc_pathologic_stageStage IA   3.018e-01  3.313e+00   0.03646     2.499
# ajcc_pathologic_stageStage IB   7.523e-07  1.329e+06   0.00000       Inf
# ajcc_pathologic_stageStage II   2.197e+00  4.553e-01   0.24838    19.426
# ajcc_pathologic_stageStage IIA  1.339e+00  7.467e-01   0.53272     3.367
# ajcc_pathologic_stageStage IIB  1.584e+00  6.313e-01   0.60964     4.116
# ajcc_pathologic_stageStage IIIA 2.513e+00  3.980e-01   0.91738     6.883
# ajcc_pathologic_stageStage IIIB 1.838e+00  5.441e-01   0.45361     7.446
# ajcc_pathologic_stageStage IIIC 3.369e+00  2.968e-01   0.67479    16.821
# ajcc_pathologic_stageStage IV   2.459e+01  4.067e-02   7.22404    83.672
# ajcc_pathologic_stageStage X    1.361e+00  7.345e-01   0.15942    11.627
# age_at_index                    1.046e+00  9.564e-01   1.02194     1.070
# 
# Concordance= 0.771  (se = 0.033 )
# Likelihood ratio test= 45.83  on 12 df,   p=7e-06
# Wald test            = 61.02  on 12 df,   p=1e-08
# Score (logrank) test = 108.7  on 12 df,   p=<2e-16

```


## COAD
### Survival curves

```{r}
#### CONTINUE HERE ####
# 1. Define the 25% (quartile) cutoffs
lower_cutoff <- quantile(coad_engagement_scores$engagement_prob, 0.25)
upper_cutoff <- quantile(coad_engagement_scores$engagement_prob, 0.75)

# 2. Isolate extremes, strictly filter for primary tumors, and merge
# coad_survival_data <- coad_engagement_scores %>%
#   filter(engagement_prob <= lower_cutoff | engagement_prob >= upper_cutoff) %>%
#   mutate(
#     engagement_group = if_else(engagement_prob >= upper_cutoff, "Engager", "Non-Engager"),
#     patient_id = substr(case_id, 1, 12),
#     sample_type = substr(case_id, 14, 15) # Extract the 2-digit sample code
#   ) %>%
#   # Safeguard 1: Strictly keep only Primary Solid Tumors ("01")
#   filter(sample_type == "01") %>%
#   # Safeguard 2: Now safely drop replicate vials (A vs B)
#   distinct(patient_id, .keep_all = TRUE) %>%
#   
#   # Join and build survival variables
#   inner_join(
#     coad_clinical %>% distinct(submitter_id, .keep_all = TRUE), 
#     by = c("patient_id" = "submitter_id")
#   ) %>%
#   mutate(
#     os_time = if_else(vital_status == "Dead", days_to_death, days_to_last_follow_up),
#     os_status = if_else(vital_status == "Dead", 1, 0)
#   ) %>%
#   filter(!is.na(os_time) & os_time > 0 & !is.na(os_status))

coad_survival_data <- coad_engagement_scores %>%
  mutate(
    patient_id  = substr(case_id, 1, 12),
    sample_type = substr(case_id, 14, 15)
  ) %>%
  filter(sample_type == "01") %>%
  distinct(patient_id, .keep_all = TRUE) %>%
  inner_join(
    coad_clinical %>% distinct(submitter_id, .keep_all = TRUE),
    by = c("patient_id" = "submitter_id")
  ) %>%
  mutate(
    os_time   = if_else(vital_status == "Dead", days_to_death, days_to_last_follow_up),
    os_status = if_else(vital_status == "Dead", 1, 0)
  ) %>%
  filter(!is.na(os_time) & os_time > 0 & !is.na(os_status))

# 3. Verify the final dataset
dim(coad_survival_data)
#  [1] 389 113

coad_survival_data %>% summarise(above_0.5 = sum(engagement_prob > 0.5))
# # A tibble: 1 × 1
#   above_0.5
#       <int>
# 1        33



```

### Plot KM curves
```{r}

coad_survival_data <- coad_survival_data %>%
  mutate(engagement_group = factor(
    if_else(engagement_prob > 0.5, "Engager", "Non-Engager"),
    levels = c("Engager", "Non-Engager")
  ))

# See percentage of censored data
coad_survival_data %>%
  group_by(engagement_group) %>%
  summarize(
    total_n = n(),
    deaths = sum(os_status == 1),
    censored = sum(os_status == 0),
    percent_censored = round((censored / total_n) * 100, 1)
  )

# A tibble: 2 × 5
#   engagement_group total_n deaths censored percent_censored
#   <fct>              <int>  <int>    <int>            <dbl>
# 1 Engager               33     12       21             63.6
# 2 Non-Engager          356     83      273             76.7


# 1. Fit the survival model using your existing variable name
coad_survival_data$engagement_group <- factor(
  coad_survival_data$engagement_group, 
  levels = c("Engager", "Non-Engager")
)
coad_km_fit <- survfit(Surv(os_time, os_status) ~ engagement_group, data = coad_survival_data)


# 2. Generate the Kaplan-Meier Plot
coad_surv_plot <- ggsurvplot(
  coad_km_fit,
  data = coad_survival_data,
  pval = TRUE,
  risk.table = TRUE,
  palette = c("#E41A1C", "#377EB8"), 
  title = "COAD Overall Survival: Engager (prob > 0.5) vs Non-Engager",
  xlab = "Time (Days)",
  ylab = "Survival Probability",
  
  # Standardize legend position and labels
  legend = "top",
  legend.title = "Status",
  legend.labs = c("Engager", "Non-Engager"),
  
  # Granular time axis
  break.time.by = 1000,
  ggtheme = theme_minimal()
)

png("pkr/pkr_quartile_grouping/survival/coad_survplot_quartile.png", width = 10, height = 7, units = "in", res = 300)
print(coad_surv_plot)
dev.off()
```

### CoxPH
with covariates
```{r}

# Force Non-Engager to be the reference group (baseline)
# coad_survival_data$engagement_group <- relevel(factor(coad_survival_data$engagement_group), ref = "Non-Engager")

# 1. Fit a Multivariate Cox Model
# include clinical factors
# coad_cox_model <- coxph(
#   Surv(os_time, os_status) ~ engagement_group + ajcc_pathologic_stage + age_at_index, 
#   data = coad_survival_data
# )

# Continuous Cox model
coad_cox_model <- coxph(
  Surv(os_time, os_status) ~ engagement_prob + ajcc_pathologic_stage + age_at_index,
  data = coad_survival_data
)
summary(coad_cox_model)

# Call:
# coxph(formula = Surv(os_time, os_status) ~ engagement_prob + 
#     ajcc_pathologic_stage + age_at_index, data = coad_survival_data)
# 
#   n= 349, number of events= 61 
#    (40 observations deleted due to missingness)
# 
#                                       coef  exp(coef)   se(coef)      z Pr(>|z|)    
# engagement_prob                  1.059e+00  2.883e+00  5.460e-01  1.939 0.052529 .  
# ajcc_pathologic_stageStage IA   -1.383e+01  9.825e-07  8.583e+03 -0.002 0.998714    
# ajcc_pathologic_stageStage II    1.134e+00  3.107e+00  7.162e-01  1.583 0.113443    
# ajcc_pathologic_stageStage IIA   4.008e-01  1.493e+00  6.632e-01  0.604 0.545562    
# ajcc_pathologic_stageStage IIB   1.550e+00  4.712e+00  9.149e-01  1.694 0.090187 .  
# ajcc_pathologic_stageStage III   1.704e+00  5.498e+00  7.673e-01  2.221 0.026338 *  
# ajcc_pathologic_stageStage IIIA -1.495e+01  3.230e-07  3.449e+03 -0.004 0.996542    
# ajcc_pathologic_stageStage IIIB  6.934e-01  2.000e+00  7.339e-01  0.945 0.344795    
# ajcc_pathologic_stageStage IIIC  1.602e+00  4.962e+00  6.792e-01  2.358 0.018364 *  
# ajcc_pathologic_stageStage IV    2.793e+00  1.633e+01  6.319e-01  4.420 9.86e-06 ***
# ajcc_pathologic_stageStage IVA   2.419e+00  1.123e+01  7.430e-01  3.255 0.001133 ** 
# ajcc_pathologic_stageStage IVB  -1.456e+01  4.749e-07  8.894e+03 -0.002 0.998694    
# age_at_index                     4.468e-02  1.046e+00  1.313e-02  3.402 0.000668 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
#                                 exp(coef) exp(-coef) lower .95 upper .95
# engagement_prob                 2.883e+00  3.469e-01    0.9885     8.406
# ajcc_pathologic_stageStage IA   9.825e-07  1.018e+06    0.0000       Inf
# ajcc_pathologic_stageStage II   3.107e+00  3.218e-01    0.7633    12.648
# ajcc_pathologic_stageStage IIA  1.493e+00  6.698e-01    0.4070     5.477
# ajcc_pathologic_stageStage IIB  4.712e+00  2.122e-01    0.7843    28.311
# ajcc_pathologic_stageStage III  5.498e+00  1.819e-01    1.2219    24.739
# ajcc_pathologic_stageStage IIIA 3.230e-07  3.096e+06    0.0000       Inf
# ajcc_pathologic_stageStage IIIB 2.000e+00  4.999e-01    0.4747     8.431
# ajcc_pathologic_stageStage IIIC 4.962e+00  2.015e-01    1.3106    18.784
# ajcc_pathologic_stageStage IV   1.633e+01  6.123e-02    4.7333    56.349
# ajcc_pathologic_stageStage IVA  1.123e+01  8.904e-02    2.6181    48.182
# ajcc_pathologic_stageStage IVB  4.749e-07  2.106e+06    0.0000       Inf
# age_at_index                    1.046e+00  9.563e-01    1.0191     1.073
# 
# Concordance= 0.802  (se = 0.031 )
# Likelihood ratio test= 70.55  on 13 df,   p=6e-10
# Wald test            = 70.74  on 13 df,   p=6e-10
# Score (logrank) test = 102.9  on 13 df,   p=5e-16
# 


# Now the coxph with the cutoff at 0.5
coad_survival_data$engagement_group <- relevel(
  factor(coad_survival_data$engagement_group), ref = "Non-Engager"
)
coad_cox_grouped_by_cutoff <- coxph(
  Surv(os_time, os_status) ~ engagement_group + ajcc_pathologic_stage + age_at_index,
  data = coad_survival_data
)
summary(coad_cox_grouped_by_cutoff)

# Call:
# coxph(formula = Surv(os_time, os_status) ~ engagement_group + 
#     ajcc_pathologic_stage + age_at_index, data = coad_survival_data)
# 
#   n= 349, number of events= 61 
#    (40 observations deleted due to missingness)
# 
#                                       coef  exp(coef)   se(coef)      z Pr(>|z|)    
# engagement_groupEngager          8.373e-01  2.310e+00  3.610e-01  2.320  0.02035 *  
# ajcc_pathologic_stageStage IA   -1.389e+01  9.237e-07  8.611e+03 -0.002  0.99871    
# ajcc_pathologic_stageStage II    1.197e+00  3.311e+00  7.161e-01  1.672  0.09458 .  
# ajcc_pathologic_stageStage IIA   4.020e-01  1.495e+00  6.631e-01  0.606  0.54437    
# ajcc_pathologic_stageStage IIB   1.513e+00  4.541e+00  9.151e-01  1.654  0.09822 .  
# ajcc_pathologic_stageStage III   1.787e+00  5.971e+00  7.678e-01  2.327  0.01995 *  
# ajcc_pathologic_stageStage IIIA -1.503e+01  2.973e-07  3.433e+03 -0.004  0.99651    
# ajcc_pathologic_stageStage IIIB  7.079e-01  2.030e+00  7.337e-01  0.965  0.33460    
# ajcc_pathologic_stageStage IIIC  1.608e+00  4.992e+00  6.793e-01  2.367  0.01793 *  
# ajcc_pathologic_stageStage IV    2.826e+00  1.688e+01  6.285e-01  4.496 6.91e-06 ***
# ajcc_pathologic_stageStage IVA   2.400e+00  1.102e+01  7.433e-01  3.228  0.00124 ** 
# ajcc_pathologic_stageStage IVB  -1.459e+01  4.602e-07  8.925e+03 -0.002  0.99870    
# age_at_index                     4.367e-02  1.045e+00  1.304e-02  3.349  0.00081 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
#                                 exp(coef) exp(-coef) lower .95 upper .95
# engagement_groupEngager         2.310e+00  4.329e-01    1.1387     4.687
# ajcc_pathologic_stageStage IA   9.237e-07  1.083e+06    0.0000       Inf
# ajcc_pathologic_stageStage II   3.311e+00  3.021e-01    0.8135    13.473
# ajcc_pathologic_stageStage IIA  1.495e+00  6.690e-01    0.4075     5.482
# ajcc_pathologic_stageStage IIB  4.541e+00  2.202e-01    0.7555    27.295
# ajcc_pathologic_stageStage III  5.971e+00  1.675e-01    1.3258    26.894
# ajcc_pathologic_stageStage IIIA 2.973e-07  3.364e+06    0.0000       Inf
# ajcc_pathologic_stageStage IIIB 2.030e+00  4.927e-01    0.4819     8.549
# ajcc_pathologic_stageStage IIIC 4.992e+00  2.003e-01    1.3186    18.902
# ajcc_pathologic_stageStage IV   1.688e+01  5.924e-02    4.9245    57.858
# ajcc_pathologic_stageStage IVA  1.102e+01  9.075e-02    2.5672    47.297
# ajcc_pathologic_stageStage IVB  4.602e-07  2.173e+06    0.0000       Inf
# age_at_index                    1.045e+00  9.573e-01    1.0183     1.072
# 
# Concordance= 0.804  (se = 0.031 )
# Likelihood ratio test= 71.75  on 13 df,   p=4e-10
# Wald test            = 72.32  on 13 df,   p=3e-10
# Score (logrank) test = 105.4  on 13 df,   p=<2e-16
```


