---
title: "de_analysis.Rmd"
output: html_document
date: "2026-04-09"
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(results = "hide")


```


```{r packages}
if(!require("BiocManager", quietly = T))
  install.packages("BiocManager")

# Install the necessary bioinformatics packages
# TCGAbiolinks: for data access
# GSVA: for the 94-gene signature scoring
# SummarizedExperiment: the standard data container for TCGA data
BiocManager::install(c("TCGAbiolinks", "GSVA", "SummarizedExperiment"))
BiocManager::install("IlluminaHumanMethylation450kanno.ilmn12.hg19")
BiocManager::install("sesameData")
BiocManager::install("sesame")
BiocManager::install("sesame")
BiocManager::install("edgeR")
BiocManager::install("org.Hs.eg.db")

install.packages("nloptr")
install.packages("ggpubr")
install.packages("cowplot")
install.packages("gt")
install.packages("minifi")

```


```{r load packages}
library(BiocManager)
library(TCGAbiolinks)
library(GSVA)
library(SummarizedExperiment)
library(tidyverse)
library(nloptr)
library(ggpubr)
library(cowplot)
library(gt)
library(webshot2)
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
library(sesameData)
library(sesame)
library(edgeR)
library(org.Hs.eg.db)
library(AnnotationDbi)
```

```{r read objects}

n_engager_loci_per_sample_candidates <- readRDS("n_engager_loci_per_sample_candidates.rds")
pol3_loci_long <- readRDS("pol3_loci_long.rds")
pol3_loci <- readRDS("pol3_loci.rds")
pol3_gr <- readRDS("pol3_gr.rds")

```

Note: HNSC has the strongest biological signal but the smallest ATAC-seq sample size in this dataset.
```{r}
# Extract raw counts for DEseq ####

candidates <- c("BRCA", "COAD", "KIRP", "STAD", "KIRC", "BLCA")

counts_list <- list()
meta_list <- list()

prepare_rnaseq <- function(cancer_type, matched_samples) {
  message("Processing ", cancer_type)
  
  query <- GDCquery(
    project = paste0("TCGA-", cancer_type),
    data.category = "Transcriptome Profiling",
    data.type = "Gene Expression Quantification",
    workflow.type = "STAR - Counts",
    sample.type = "Primary Tumor"
  )
  
  se <- GDCprepare(query)
  
  # Filter to ATAC-seq matched samples only
  matched_barcodes <- matched_samples %>%
    filter(cancer_type == .env$cancer_type) %>%
    pull(case_id)
  
  #se_matched <- se[, se$barcode %in% matched_barcodes]
  se_matched <- se[, substr(colData(se)$barcode, 1, 12) %in% substr(matched_barcodes, 1, 12)]
  se_matched <- se_matched[, !duplicated(substr(colData(se_matched)$barcode, 1, 12))]
  
  message("  Matched ", ncol(se_matched), " of ", length(matched_barcodes), " samples")
  
  list(
    counts = assay(se_matched, "unstranded"),
    se = se_matched,
    meta = colData(se_matched) %>%
      as.data.frame() %>%
      dplyr::select(barcode, sample_type) %>%
      dplyr::filter(barcode %in% colnames(assay(se_matched, "unstranded"))) %>%
      mutate(case_id = substr(barcode, 1, 12)) %>%
      left_join(
        matched_samples %>%
          filter(cancer_type == .env$cancer_type) %>%
          dplyr::select(case_id, group, n_engager_loci),
        by = "case_id"
      )
  )
}
```

# Try one cancertype fist BRCA
```{r}
brca_rnaseq_data <- prepare_rnaseq("BRCA", n_engager_loci_per_sample_candidates)

dim(brca_rnaseq_data$counts)

# does meta have group assignments?
head(brca_rnaseq_data$meta)

# check group distribution
table(brca_rnaseq_data$meta$group)


# Create DGEList object
dge <- DGEList(
  counts = brca_rnaseq_data$counts,
  group = brca_rnaseq_data$meta$group
)

# Filter lowly expressed genes
keep <- filterByExpr(dge)
dge <- dge[keep, , keep.lib.sizes = FALSE]

# Normalize
dge <- calcNormFactors(dge)

# Design matrix - engager vs baseline
design <- model.matrix(~ 0 + group, data = brca_rnaseq_data$meta)
colnames(design) <- levels(factor(brca_rnaseq_data$meta$group))

# Estimate dispersion
dge <- estimateDisp(dge, design)

# Fit model
fit <- glmQLFit(dge, design)

# Engager vs baseline contrast
con <- makeContrasts(engager - non_engager, levels = design)
qlf <- glmQLFTest(fit, contrast = con)

# Results
de_results_brca_engager_vs_nonengager <- topTags(qlf, n = Inf)$table
head(de_results_brca_engager_vs_nonengager)

# Strip version numbers from Ensembl IDs
de_results_brca_engager_vs_nonengager$ensembl_id <- rownames(de_results_brca_engager_vs_nonengager) %>% 
  str_remove("\\..*")

# Map to gene symbols
de_results_brca_engager_vs_nonengager$gene_name <- mapIds(
  org.Hs.eg.db,
  keys = de_results_brca_engager_vs_nonengager$ensembl_id,
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)

View(de_results_brca_engager_vs_nonengager)

de_results_brca_engager_vs_nonengager %>%
  filter(FDR < 0.01, !is.na(gene_name)) %>%
  nrow()
# 43 genes FDR < 0.01

de_results_brca_engager_vs_nonengager %>%
  filter(FDR < 0.01, !is.na(gene_name)) %>%
  count(direction = ifelse(logFC > 0, "up", "down"))
#   direction  n
# 1      down 22
# 2        up 21

# Sanity check -- are any IFN/innate immune genes in the upregulated set?
de_results_brca_engager_vs_nonengager %>%
  filter(FDR < 0.01, !is.na(gene_name), logFC > 0) %>%
  arrange(FDR) %>%
  pull(gene_name)
```

## Check in one gene in this upregulated list has any loci that are "engager loci" 5ppp==1 accessibility == 1 that's in rig-i recognized and pol II transcribed
```{r}
 
de_results_brca_engager_vs_nonengager %>%
  filter(gene_name == "ERVFRD-1") %>%
  dplyr::select(ensembl_id, logFC, FDR)

# Create GRanges for ERVFRD-1
ervfrd1_gr <- rowRanges(se_brca)[rowData(se_brca)$gene_name == "ERVFRD-1"]
ervfrd1_gr

# Create GRanges for engager loci in brca
# Unique engager loci accessible in at least one BRCA sample
brca_engager_loci <- pol3_loci_long %>%
  filter(cancer_type == "BRCA",
         is_5ppp == 1,
         accessibility == 1) %>%
  distinct(V1, V2, V3)

# Create GRanges
engager_gr_brca <- GRanges(
  seqnames = brca_engager_loci$V1,
  ranges = IRanges(
    start = as.integer(brca_engager_loci$V2),
    end = as.integer(brca_engager_loci$V3)
  )
)

# Creat GRanges for upregulated genes
up_genes <- de_results_brca_engager_vs_nonengager %>%
  filter(FDR < 0.01, !is.na(gene_name), logFC > 0) %>%
  pull(gene_name)

up_gr <- rowRanges(se_brca)[rowData(se_brca)$gene_name %in% up_genes]
up_gr

# Check overlap with upregulated DE genes
overlaps_up_engager <- findOverlaps(up_gr, engager_gr_brca)
names(up_gr)[queryHits(overlaps_up_engager)]
engager_gr_brca[subjectHits(overlaps_up_engager)]

# Check overlap with POL III loci
overlaps <- findOverlaps(ervfrd1_gr, pol3_gr)
overlaps

# # Something's wrong with the barcode format. it's returned 0matched samples
# # query and prepare for brca
# query_brca <- GDCquery(
#   project = "TCGA-BRCA",
#   data.category = "Transcriptome Profiling",
#   data.type = "Gene Expression Quantification",
#   workflow.type = "STAR - Counts",
#   sample.type = "Primary Tumor"
# )
# 
# se_brca <- GDCprepare(query_brca)
# head(colData(se_brca)$barcode)

# Check if any upregulated genes have "engager" loci
```

# Function for DE analysis
1) run edgeR engagers vs non_engagers
2) get up and down genes
3) Are any up genes in IFN signature genes?
4) Do up genes have engager loci within them?

Then we can test in BRCA 
```{r}

run_de_analysis <- function(cancer_type, matched_samples, pol3_loci_long, ifn_genes) {
  
  message("Running DE for ", cancer_type)
  
  # Prepare RNA-seq data
  data <- prepare_rnaseq(cancer_type, matched_samples)
  
  # 1. EdgeR DE engager vs non_engager ####
  dge <- DGEList(counts = data$counts, group = data$meta$group)
  keep <- filterByExpr(dge)
  dge <- dge[keep, , keep.lib.sizes = FALSE]
  dge <- calcNormFactors(dge)
  
  design <- model.matrix(~ 0 + group, data = data$meta)
  colnames(design) <- levels(factor(data$meta$group))
  dge <- estimateDisp(dge, design)
  fit <- glmQLFit(dge, design)
  
  con <- makeContrasts(engager - non_engager, levels = design)
  qlf <- glmQLFTest(fit, contrast = con)
  
  results <- topTags(qlf, n = Inf)$table
  results$ensembl_id <- rownames(results) %>% str_remove("\\..*")
  results$gene_name <- mapIds(org.Hs.eg.db,
                              keys = results$ensembl_id,
                              column = "SYMBOL",
                              keytype = "ENSEMBL",
                              multiVals = "first")
  
  # 2. Upregulated genes ####
  up_genes <- results %>%
    filter(FDR < 0.01, !is.na(gene_name), logFC > 0) %>%
    pull(gene_name)
  
  # 3. IFN overlap ####
  ifn_overlap <- intersect(up_genes, ifn_genes)
  
  # 4. Engager loci overlap ####
  engager_loci <- pol3_loci_long %>%
    filter(cancer_type == .env$cancer_type,
           is_5ppp == 1,
           accessibility == 1) %>%
    distinct(V1, V2, V3)
  
  engager_gr <- GRanges(
    seqnames = engager_loci$V1,
    ranges = IRanges(start = as.integer(engager_loci$V2),
                     end = as.integer(engager_loci$V3))
  )
  
  up_gr <- rowRanges(data$se)[rowData(data$se)$gene_name %in% up_genes]
  overlaps <- findOverlaps(up_gr, engager_gr)
  genes_with_engager_loci <- names(up_gr)[queryHits(overlaps)]
  
  list(
    de_results = results,
    up_genes = up_genes,
    ifn_overlap = ifn_overlap,
    genes_with_engager_loci = genes_with_engager_loci,
    n_samples = nrow(data$meta),
    n_per_group = table(data$meta$group)
  )
}

brca_de <- run_de_analysis("BRCA", n_engager_loci_per_sample_candidates, pol3_loci_long, ifn_genes)

# How many DE genes
length(brca_de$up_genes)

# Any IFN overlap
brca_de$ifn_overlap

# Any genes with engager loci
brca_de$genes_with_engager_loci

# Sample counts
brca_de$n_per_group
```

# For all cancer types
```{r}
candidates <- c("BRCA", "COAD", "KIRP", "STAD", "KIRC", "BLCA")

de_results_all <- lapply(candidates, run_de_analysis,
                         matched_samples = n_engager_loci_per_sample_candidates,
                         pol3_loci_long = pol3_loci_long,
                         ifn_genes = ifn_genes)
names(de_results_all) <- candidates

#### Debugging_____________________

# Run just the prepare step for COAD to inspect
coad_data <- prepare_rnaseq("COAD", n_engager_loci_per_sample_candidates)
nrow(coad_data$meta)
ncol(coad_data$counts)

# Check which barcodes are in counts but not meta
setdiff(colnames(coad_data$counts), coad_data$meta$barcode)

# Check which are in meta but not counts
setdiff(coad_data$meta$barcode, colnames(coad_data$counts))

table(coad_data$meta$group, useNA = "always")

coad_de <- run_de_analysis("COAD", n_engager_loci_per_sample_candidates, pol3_loci_long, ifn_genes)


rm(coad_de)
rm(coad_data)
```


