---
title: "de_analysis.Rmd"
output: html_document
date: "2026-04-09"
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(results = "hide")


```


```{r packages}
install.packages("BiocManager")

# Install the necessary bioinformatics packages
# TCGAbiolinks: for data access
# GSVA: for the 94-gene signature scoring
# SummarizedExperiment: the standard data container for TCGA data
BiocManager::install(c("TCGAbiolinks", "GSVA", "SummarizedExperiment"))
BiocManager::install("IlluminaHumanMethylation450kanno.ilmn12.hg19")
BiocManager::install("sesameData")
BiocManager::install("sesame")
BiocManager::install("sesame")
BiocManager::install("edgeR")
BiocManager::install("org.Hs.eg.db")
BiocManager::install("DESeq2")
BiocManager::install("apeglm", update = FALSE, ask = FALSE)

BiocManager::install(c("BiocGenerics", "S4Vectors", "IRanges"), force = TRUE, update = TRUE, ask = FALSE)
BiocManager::install("DESeq2", force=TRUE, update = FALSE, ask = FALSE)

install.packages("nloptr")
install.packages("ggpubr")
install.packages("cowplot")
install.packages("gt")
install.packages("minifi")

# 4/15/26 trying to get DESeq2
# 1. Remove the mismatched packages from your local Linux library
remove.packages(c("BiocGenerics", "S4Vectors", "IRanges"))

# 2. Force BiocManager to use the correct repository for R 4.3
BiocManager::install(version = "3.18", ask = FALSE)

# 3. Install DESeq2 cleanly from the 3.18 repository
BiocManager::install("DESeq2", update = FALSE, ask = FALSE)

BiocManager::install(c("BiocGenerics", "IRanges", "S4Vectors", "DESeq2"), update = FALSE, ask = FALSE)

# 1. Delete the corrupted phantom folder directly from the file system
unlink("/home/nvsharm/R/x86_64-pc-linux-gnu-library/4.3/S4Vectors", recursive = TRUE)

# 2. Install S4Vectors by pointing explicitly to the Bioconductor 3.18 repository
install.packages("S4Vectors", repos = "https://bioconductor.org/packages/3.18/bioc")

install.packages("IRanges", repos = "https://bioconductor.org/packages/3.18/bioc")

# 3. Test DESeq2
library(DESeq2)




# Install clusterProfiler and msigdbr
BiocManager::install("clusterProfiler")
install.packages("msigdbr")

```


```{r load packages}
library(BiocManager)
library(TCGAbiolinks)
library(GSVA)
library(SummarizedExperiment)
library(tidyverse)
library(nloptr)
library(ggpubr)
library(cowplot)
library(gt)
library(webshot2)
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
library(sesameData)
library(sesame)
library(edgeR)
library(org.Hs.eg.db)
library(AnnotationDbi)
library(S4Vectors)
library(IRanges)
library(DESeq2)
library(apeglm)
library(clusterProfiler)
library(msigdbr)
```

```{r read objects}

#candidate_sample_groups <- readRDS("candidate_sample_groups.rds")
#pol3_loci_long <- readRDS("pol3_loci_long.rds")
#pol3_gr <- readRDS("pol3_gr.rds")

```

# Extract raw counts for de analysis
Note: HNSC has the strongest biological signal but the smallest ATAC-seq sample size in this dataset.
```{r}


candidates <- c("BRCA", "COAD", "KIRP", "STAD", "KIRC", "BLCA")



prepare_rnaseq <- function(cancer_type, matched_samples) {
  message("Processing ", cancer_type)
  
  query <- GDCquery(
    project = paste0("TCGA-", cancer_type),
    data.category = "Transcriptome Profiling",
    data.type = "Gene Expression Quantification",
    workflow.type = "STAR - Counts",
    sample.type = "Primary Tumor"
  )
  
  se <- GDCprepare(query)
  
  # Filter to ATAC-seq matched samples only
  matched_barcodes <- matched_samples %>%
    filter(cancer_type == .env$cancer_type) %>%
    pull(case_id)
  
  #se_matched <- se[, se$barcode %in% matched_barcodes]
  se_matched <- se[, substr(colData(se)$barcode, 1, 12) %in% substr(matched_barcodes, 1, 12)]
  se_matched <- se_matched[, !duplicated(substr(colData(se_matched)$barcode, 1, 12))]
  
  message("  Matched ", ncol(se_matched), " of ", length(matched_barcodes), " samples")
  
  list(
    counts = assay(se_matched, "unstranded"),
    se = se_matched,
    meta = colData(se_matched) %>%
      as.data.frame() %>%
      mutate(case_id = substr(barcode, 1, 12)) %>%
      left_join(
        matched_samples %>%
          filter(cancer_type == .env$cancer_type) %>%
          dplyr::select(case_id, engager_group),
        by = "case_id"
      )
  )
}
```

```{r}

prepare_and_cache_rnaseq <- function(cancer_type, matched_samples, cache_dir = "tcga_cache") {
  
  # 1. Setup cache directory
  if(!dir.exists(cache_dir)) dir.create(cache_dir)
  cache_file <- file.path(cache_dir, paste0(cancer_type, "_rnaseq_matched.rds"))
  
  # 2. Check for cached data
  if(file.exists(cache_file)) {
    message("\n>>> Loading CACHED RNA-seq data for ", cancer_type)
    return(readRDS(cache_file))
  }
  
  # 3. If no cache, perform the download and processing
  message("\n>>> Downloading NEW RNA-seq data for ", cancer_type)
  
  query <- GDCquery(
    project = paste0("TCGA-", cancer_type),
    data.category = "Transcriptome Profiling",
    data.type = "Gene Expression Quantification",
    workflow.type = "STAR - Counts",
    sample.type = "Primary Tumor"
  )
  
  GDCdownload(query = query)
  se <- GDCprepare(query)
  
  # Filter to ATAC-seq matched samples
  matched_barcodes <- matched_samples %>%
    filter(cancer_type == .env$cancer_type) %>%
    pull(case_id)
  
  se_matched <- se[, substr(colData(se)$barcode, 1, 12) %in% substr(matched_barcodes, 1, 12)]
  se_matched <- se_matched[, !duplicated(substr(colData(se_matched)$barcode, 1, 12))]
  
  message("  Matched ", ncol(se_matched), " of ", length(matched_barcodes), " samples")
  
  # Compile the final object
  final_data <- list(
    counts = assay(se_matched, "unstranded"),
    se = se_matched,
    meta = colData(se_matched) %>%
      as.data.frame() %>%
      mutate(case_id = substr(barcode, 1, 12)) %>%
      left_join(
        matched_samples %>%
          filter(cancer_type == .env$cancer_type) %>%
          dplyr::select(case_id, engager_group),
        by = "case_id"
      )
  )
  
  # 4. Save to cache
  saveRDS(final_data, cache_file)
  message("  Saved to cache: ", cache_file)
  
  return(final_data)
}

```


# Functions for DE analysis
1) run edgeR engagers vs non_engagers
2) get up and down genes
3) Are any up genes in IFN signature genes?
4) Do up genes have engager loci within them?

Then we can test in BRCA 
## edgeR
```{r}
run_de_analysis <- function(cancer_type, matched_samples, pol3_loci_long, ifn_genes) {
  
  message("Running DE for ", cancer_type)
  
  # Prepare RNA-seq data
  data <- prepare_rnaseq(cancer_type, matched_samples)
  
  # 1. EdgeR DE engager vs non_engager ####
  dge <- DGEList(counts = data$counts, group = data$meta$engager_group)
  keep <- filterByExpr(dge)
  dge <- dge[keep, , keep.lib.sizes = FALSE]
  dge <- calcNormFactors(dge)
  
  design <- model.matrix(~ 0 + engager_group, data = data$meta)
  colnames(design) <- levels(factor(data$meta$engager_group))
  dge <- estimateDisp(dge, design)
  fit <- glmQLFit(dge, design)
  
  con <- makeContrasts(engager - non_engager, levels = design)
  qlf <- glmQLFTest(fit, contrast = con)
  
  results <- topTags(qlf, n = Inf)$table
  results$ensembl_id <- rownames(results) %>% str_remove("\\..*")
  results$gene_name <- mapIds(org.Hs.eg.db,
                              keys = results$ensembl_id,
                              column = "SYMBOL",
                              keytype = "ENSEMBL",
                              multiVals = "first")
  
  # 2. Upregulated and downregulated genes ####
  up_genes <- results %>%
    filter(FDR < 0.05, !is.na(gene_name), logFC > 0) %>%
    pull(gene_name)
  
  down_genes <- results %>%
    filter(FDR < 0.05, !is.na(gene_name), logFC < 0) %>%
    pull(gene_name)
  
  # 3. IFN overlap ####
  ifn_overlap <- intersect(up_genes, ifn_genes)
  
  # 4. Engager loci overlap ####
  engager_loci <- pol3_loci_long %>%
    filter(cancer_type == .env$cancer_type,
           is_5ppp == 1,
           accessibility == 1) %>%
    distinct(V1, V2, V3)
  
  engager_gr <- GRanges(
    seqnames = engager_loci$V1,
    ranges = IRanges(start = as.integer(engager_loci$V2),
                     end = as.integer(engager_loci$V3))
  )
  
  up_gr <- rowRanges(data$se)[rowData(data$se)$gene_name %in% up_genes]
  overlaps <- findOverlaps(up_gr, engager_gr)
  genes_with_engager_loci <- names(up_gr)[queryHits(overlaps)]
  
  list(
    de_results = results,
    up_genes = up_genes,
    down_genes = down_genes,
    ifn_overlap = ifn_overlap,
    genes_with_engager_loci = genes_with_engager_loci,
    n_samples = nrow(data$meta),
    n_per_group = table(data$meta$engager_group)
  )
}
```

### For one cancer type
```{r}
brca_de <- run_de_analysis("BRCA", candidate_sample_groups, pol3_loci_long, ifn_genes)

# How many DE genes
length(brca_de$up_genes) #11
length(brca_de$down_genes) #20

# Any IFN overlap
brca_de$ifn_overlap

# Any genes with engager loci
brca_de$genes_with_engager_loci

# Sample counts
brca_de$n_per_group

# Are there any IFN genes that are even in the expression matrix and were filtered out?
intersect(ifn_genes, rownames(brca_de$de_results %>% filter(!is.na(gene_name)))) # 0 in expression matrix


```

### For all cancer types
```{r}
candidates <- c("COAD", "KIRP", "STAD", "KIRC", "BLCA") #"BRCA", 

de_results_all <- lapply(candidates, run_de_analysis,
                         matched_samples = candidate_sample_groups,
                         pol3_loci_long = pol3_loci_long,
                         ifn_genes = ifn_genes)
names(de_results_all) <- candidates

de_results_all[["BRCA"]] <- brca_de

# Rename in active environment b/c doing DESeq2 also
#de_results_all_edgeR <- de_results_all

# Save object to disk
#saveRDS(de_results_all_edgeR, "de_results_all_edgeR.rds")

```

### Dig into DE results for all cancer types
```{r}
lapply(names(de_results_all), function(ct) {
  cat("\n", ct, "\n")
  cat("N per group:", de_results_all[[ct]]$n_per_group, "\n")
  cat("N up genes FDR<0.05:", length(de_results_all[[ct]]$up_genes), "\n")
  cat("N down genes FDR<0.05:", length(de_results_all[[ct]]$down_genes), "\n")
  cat("IFN overlap:", paste(de_results_all[[ct]]$ifn_overlap, collapse=", "), "\n")
  cat("Genes with engager loci:", paste(de_results_all[[ct]]$genes_with_engager_loci, collapse=", "), "\n")
})


```

## DESeq2
### Module 1 Core DESeq2 Function
```{r}

# Function handles matrix setup, variance stabilization, and DE testing 
# returns a list containing the raw dds object, the full results df, and intersection logic
run_deseq2_core <- function(cancer_type, matched_samples, pol3_loci_long, ifn_genes) {
  
  message("\n>>> Running DESeq2 Core for ", cancer_type)
  data <- prepare_and_cache_rnaseq(cancer_type, matched_samples)
  
  # Ensure reference level
  data$meta$engager_group <- factor(data$meta$engager_group, levels = c("non_engager", "engager"))
  
  dds <- DESeqDataSetFromMatrix(countData = data$counts,
                                colData = data$meta,
                                design = ~ engager_group)
  
  keep <- rowSums(counts(dds)) >= 10
  dds <- dds[keep,]
  dds <- DESeq(dds)
  res <- lfcShrink(dds, coef="engager_group_engager_vs_non_engager", type="apeglm")
  
  res_df <- as.data.frame(res)
  res_df$ensembl_id <- rownames(res_df) %>% str_remove("\\..*")
  res_df$gene_name <- mapIds(org.Hs.eg.db,
                             keys = res_df$ensembl_id,
                             column = "SYMBOL",
                             keytype = "ENSEMBL",
                             multiVals = "first")
  
  up_genes <- res_df %>% filter(padj < 0.05, !is.na(gene_name), log2FoldChange > 0.5) %>% pull(gene_name)
  down_genes <- res_df %>% filter(padj < 0.05, !is.na(gene_name), log2FoldChange < -0.5) %>% pull(gene_name)
  
  # Engager loci overlap logic
  engager_loci <- pol3_loci_long %>%
    filter(cancer_type == .env$cancer_type, is_5ppp == 1, accessibility == 1) %>%
    distinct(V1, V2, V3)
  
  engager_gr <- GRanges(seqnames = engager_loci$V1, ranges = IRanges(start = as.integer(engager_loci$V2), end = as.integer(engager_loci$V3)))
  up_gr <- rowRanges(data$se)[rowData(data$se)$gene_name %in% up_genes]
  overlaps <- findOverlaps(up_gr, engager_gr)
  genes_with_engager_loci <- names(up_gr)[queryHits(overlaps)]
  
  return(list(
    cancer_type = cancer_type,
    dds = dds,
    res_df = res_df,
    up_genes = up_genes,
    down_genes = down_genes,
    ifn_overlap = intersect(up_genes, ifn_genes),
    genes_with_engager_loci = genes_with_engager_loci,
    n_samples = nrow(data$meta),
    n_per_group = table(data$meta$engager_group)
  ))
}

```




### Module 2 QC Plotting
```{r}
generate_deseq_qc_plots <- function(deseq_output) {
  ct <- deseq_output$cancer_type
  dds <- deseq_output$dds
  
  dir.create("DESeq_QC_Plots", showWarnings = FALSE)
  
  # Dispersion Plot
  png(paste0("DESeq_QC_Plots/", ct, "_dispersion.png"), width=600, height=500)
  plotDispEsts(dds, main=paste(ct, "Dispersion Estimates"))
  dev.off()
  
  # PCA Plot
  vsd <- vst(dds, blind=FALSE)
  pca_data <- plotPCA(vsd, intgroup="engager_group", returnData=TRUE)
  percentVar <- round(100 * attr(pca_data, "percentVar"))
  
  pca_plot <- ggplot(pca_data, aes(PC1, PC2, color=engager_group)) +
    geom_point(size=3, alpha=0.8) +
    theme_minimal() +
    labs(title=paste(ct, "PCA"),
         x=paste0("PC1: ", percentVar[1], "%"),
         y=paste0("PC2: ", percentVar[2], "%")) +
    scale_color_manual(values=c("non_engager"="#00BFC4", "engager"="#F8766D"))
  
  ggsave(paste0("DESeq_QC_Plots/", ct, "_PCA.png"), plot=pca_plot, width=5, height=4)
  message("Saved QC plots for ", ct)
}
```


### Module 3 GSVA sample-by-sample scoring
```{r}
calculate_sample_scores <- function(deseq_output, custom_index = NULL) {
  
  # Default to the cancer-specific up_genes if no index is provided
  if(is.null(custom_index)) {
    custom_index <- deseq_output$up_genes
  }
  
  if(length(custom_index) == 0) {
    message("No index genes available for ", deseq_output$cancer_type)
    return(NULL)
  }
  
  # Extract VST normalized counts
  vsd <- vst(deseq_output$dds, blind=FALSE)
  vst_mat <- assay(vsd)
  
  # Map rownames from Ensembl to Symbol for GSVA matching
  rownames(vst_mat) <- deseq_output$res_df$gene_name[match(rownames(vst_mat), rownames(deseq_output$res_df))]
  vst_mat <- vst_mat[!is.na(rownames(vst_mat)), ] # Drop any genes that didn't map correctly to a symbol (NAs)
  
  # Put the up genes (for brca = 133 up genes) into a list format for GSVA package
  index_list <- list(Target_Index = custom_index)
  
  # run GSVA to get one score per column (per sample)
  gsva_res <- gsva(expr = vst_mat, gset.idx.list = index_list, method = "gsva", verbose = FALSE)
  
  # format into a table that can be used to make a boxplot
  sample_scores <- data.frame(
    case_id = colnames(vst_mat),
    custom_metric = as.numeric(gsva_res["Target_Index", ]),
    group = deseq_output$dds$engager_group
  )
  
  return(sample_scores)
}
```

### BRCA 
```{r}
#### Module 1_____ ####
brca_deseq2_test <- run_deseq2_core("BRCA", candidate_sample_groups, pol3_loci_long, ifn_genes)
brca_deseq2_test$n_per_group
# non_engager     engager 
#          15          14
length(brca_deseq2_test$up_genes)
# [1] 169
brca_deseq2_test$res_df %>% View


#### Module 2______ ####
# Generate the PCA and Dispersion plots
generate_deseq_qc_plots(brca_deseq2_test)


#### Module 3______ ####
# Generate the custom sample-by-sample metric
brca_gsva_scores <- calculate_sample_scores(brca_deseq2_test)

# Visualize the custom 133 gene score
group_colors <- c("non_engager" = "#4575b4", "engager" = "#d73027")

ggplot(brca_gsva_scores, aes(x = group, y = custom_metric, fill = group)) +
  # Create the boxplot with a bit of transparency
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.5, color = "black") +
  # Add the sample points (dots) and color them by group
  geom_jitter(aes(color = group), width = 0.2, size = 2.5, alpha = 0.8) +
  # Use a clean, professional theme
  theme_minimal() +
  # Map the colors we defined
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  # Refine labels and titles
  labs(
    title = "BRCA: Sample-Level Gene Set Scoring",
    subtitle = paste0("Based on ", length(brca_deseq2_test$up_genes), "-gene Signature"),
    y = "GSVA Enrichment Score",
    x = NULL # Removing 'group' label as the axis categories make it obvious
  ) +
  # Remove the legend since the x-axis already labels the groups
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 11, color = "black"),
    panel.grid.minor = element_blank()
  ) +
  stat_compare_means(method = "wilcox.test", label.x = 1.5, label.y = 0.6)




ggplot(brca_gsva_scores, aes(x = group, y = custom_metric, fill = group)) +
  geom_boxplot() +
  geom_jitter(width = 0.2) +
  theme_minimal() +
  labs(title = "BRCA: Sample-by-Sample Custom Metric (133-Gene Index)", y = "GSVA Score")


#### biological proof: enrichment of immunostimulatory pathways______ ####
# 2. Pull the Hallmark Interferon Alpha list
# ifn_alpha <- msigdbr(species = "human", category = "H") %>% 
#   filter(gs_name == "HALLMARK_INTERFERON_ALPHA_RESPONSE") %>% 
#   pull(gene_symbol)

# 2. See which of 33 genes overlap with ifn_genes
brca_deseq2_ifn_overlap <- intersect(brca_deseq2_test$up_genes, ifn_genes)

# 3. Check the results
print(paste("Number of overlapping IFN genes:", length(brca_deseq2_ifn_overlap)))
print(brca_deseq2_ifn_overlap)

# Check overlap with edgeR genes
intersect(brca_deseq2_test$up_genes, brca_de$up_genes)

```

### COAD 
```{r}
#### Module 1 ####
coad_deseq2_test <- run_deseq2_core("COAD", candidate_sample_groups, pol3_loci_long, ifn_genes)
brca_deseq2_test$n_per_group
# non_engager     engager 
#          15          14
length(coad_deseq2_test$up_genes)
# [1] 169
coad_deseq2_test$res_df %>% View


#### Module 2 ####
# Generate the PCA and Dispersion plots
generate_deseq_qc_plots(coad_deseq2_test)


#### Module 3 ####
# Generate the custom sample-by-sample metric
coad_gsva_scores <- calculate_sample_scores(coad_deseq2_test)

# Visualize the custom metric
ggplot(coad_gsva_scores, aes(x = group, y = custom_metric, fill = group)) +
  geom_boxplot() +
  geom_jitter(width = 0.2) +
  theme_minimal() +
  labs(title = "COAD: Sample-by-Sample Custom Metric (133-Gene Index)", y = "GSVA Score")

```

