---
title: "grouping_engager_status_pkr.Rmd"
output: html_document
date: "2026-06-03"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(results = "hide")


```

```{r packages}
install.packages("BiocManager")
BiocManager::install(c("TCGAbiolinks", "GSVA", "SummarizedExperiment"))
BiocManager::install("IlluminaHumanMethylation450kanno.ilmn12.hg19")
BiocManager::install("sesameData")
BiocManager::install("sesame")
BiocManager::install("sesame")
BiocManager::install("edgeR")
BiocManager::install("org.Hs.eg.db")
BiocManager::install("DESeq2")
BiocManager::install("apeglm", update = FALSE, ask = FALSE)
BiocManager::install("survminer")
BiocManager::install(c("BiocGenerics", "S4Vectors", "IRanges"), force = TRUE, update = TRUE, ask = FALSE)
BiocManager::install("DESeq2", force=TRUE, update = FALSE, ask = FALSE)

install.packages("nloptr")
install.packages("ggpubr")
install.packages("cowplot")
install.packages("gt")
install.packages("minifi")
install.packages("glmnet")
install.packages("ggrepel")
BiocManager::install("clusterProfiler")
install.packages("msigdbr")
install.packages("mclust")
install.packages("conflicted")

BiocManager::install("ComplexHeatmap")

```


```{r load packages}
library(BiocManager)
library(TCGAbiolinks)
library(GSVA)
library(SummarizedExperiment)
library(tidyverse)
library(nloptr)
library(ggpubr)
library(cowplot)
library(gt)
library(webshot2)
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
library(sesameData)
library(sesame)
library(edgeR)
library(org.Hs.eg.db)
library(AnnotationDbi)
library(S4Vectors)
library(IRanges)
library(DESeq2)
library(apeglm)
#library(clusterProfiler)
library(msigdbr)
library(glmnet)
library(survminer)
library(ggrepel)
library(mclust)
library(conflicted)
library(ComplexHeatmap)


conflict_prefer("select", "dplyr")
conflict_prefer("count", "dplyr")
conflict_prefer("filter", "dplyr")
```
1) filter the compilation file to the loci that are denoted pkr_enrichment == TRUE. these are the loci that are accessible (not necessarily transcribed by pol3 but could be pol2 anything, agnostic to pol3) and have pkr enrichment.
2) these loci also have tcga sample identifiers so i tie the pkr_enrichment column to the compilation dataframe only keeping pkr_enriched loci
3) filter by p-values with the assumption that if it is open, dsRNA is being transcribed and "engages" with pkr. 
4) cluster p-values into 3 groups just like previous analysis, and do the same downstream

# Read in coordinate file of pkr enriched loci and pol3 genomic ranges 
Look for overlap of 
```{r}
pkr_loci <-read_csv("~/real.pkr.csv") 
View(pkr_loci)

# All RNA central loci with pvalues of atacseq counts ("accessibility")
# Should have 1409 rows (loci) that overlap with identifiers in the complication file
pkr_accessibility_pvals <- read_csv("ATAC_TCGA_Compilation_pValue_Gene_Downgraded_FINAL.txt") %>% 
  filter(V1 %in% pkr_loci$identifier)
```

# transform p-vals, restrict to candidate cancers
```{r}
pkr_accessibility_pvals_long <- pkr_accessibility_pvals %>% 
  pivot_longer(
  cols = starts_with("TCGA"),
  names_to = "sample",
  values_to = "pval") %>% 
  mutate(
    cancer_type = str_extract(sample, "(?<=TCGA\\.)[A-Z]+"),
    pval = as.numeric(pval),
    sample = str_remove(sample,"^TCGA\\." )
  )

pkr_accessibility_pvals_wide <- pkr_accessibility_pvals_long %>% 
  pivot_wider(
    names_from = V1,
    values_from = pval
  )

dim(pkr_accessibility_pvals_wide)
# [1] 328 1411 -- 328 samples and 1411 columns (1409 locus columns plus sample and cancer_type) 

# Transform pvals for clustering
engager_loci_pvals <- pkr_accessibility_pvals_wide %>% 
  mutate(across(where(is.numeric), ~ -log10(. + 1e-300)))


# Get only candidate cancers
engager_loci_pvals_candidate_cancers <- engager_loci_pvals %>% 
  filter(cancer_type %in% c("BRCA", "COAD", "KIRP", "STAD", "KIRC", "BLCA"))

# seeing how many samples are in each cancer type
engager_loci_pvals_candidate_cancers %>% 
  count(cancer_type) %>% 
  arrange(n) 

# # A tibble: 6 × 2
#   cancer_type     n
#   <chr>       <int>
# 1 BLCA            8
# 2 KIRC           11
# 3 STAD           16
# 4 KIRP           33
# 5 COAD           36
# 6 BRCA           57


```

# GMM
```{r}
gmm_results_by_cancer <- engager_loci_pvals_candidate_cancers %>%
  group_by(cancer_type) %>%
  group_split() %>%
  map_dfr(function(df) {
    
    ct <- unique(df$cancer_type)
    samples <- df$sample
    
    fit <- df %>%
      dplyr::select(-sample, -cancer_type) %>%
      Mclust(G = 3)
    
    tibble(
      sample = samples,
      cancer_type = ct,
      cluster = fit$classification,
      prob_cluster1 = fit$z[, 1],
      prob_cluster2 = fit$z[, 2],
      prob_cluster3 = fit$z[, 3]
    )
  })

gmm_results_by_cancer %>%
  count(cancer_type, cluster)

# A tibble: 18 × 3
#    cancer_type cluster     n
#    <chr>         <dbl> <int>
#  1 BLCA              1     1
#  2 BLCA              2     6
#  3 BLCA              3     1
#  4 BRCA              1    53
#  5 BRCA              2     3
#  6 BRCA              3     1
#  7 COAD              1    26
#  8 COAD              2     9
#  9 COAD              3     1
# 10 KIRC              1     9
# 11 KIRC              2     1
# 12 KIRC              3     1
# 13 KIRP              1    10
# 14 KIRP              2    22
# 15 KIRP              3     1
# 16 STAD              1    14
# 17 STAD              2     1
# 18 STAD              3     1


# Looks weird so looking at distribution of transformed p-vals for each cancer type
engager_loci_pvals_candidate_cancers %>%
  pivot_longer(cols = where(is.numeric), names_to = "locus", values_to = "val") %>%
  ggplot(aes(x = cancer_type, y = val, fill = cancer_type)) +
  geom_violin() +
  theme_minimal() +
  labs(y = "-log10(p)", x = NULL) +
  theme(legend.position = "none")

engager_loci_pvals_candidate_cancers %>%
  pivot_longer(cols = where(is.numeric), names_to = "locus", values_to = "val") %>%
  group_by(cancer_type) %>%
  summarise(
    min = min(val),
    q25 = quantile(val, 0.25),
    median = median(val),
    q75 = quantile(val, 0.75),
    q95 = quantile(val, 0.95),
    q99 = quantile(val, 0.99),
    max = max(val),
    frac_300 = mean(val > 290),
    frac_0 = mean(val < 0.01)
  )


```

# Violin plot and table showing distribution of proportion of loci with p-val <0.05 in each cancer type
```{r}

#### Violin plot ####
pkr_accessibility_violin_plot <- pkr_accessibility_pvals_long %>%
  mutate(accessible = pval < 0.05) %>%
  group_by(sample, cancer_type) %>%
  summarise(frac_accessible = mean(accessible), .groups = "drop") %>%
  filter(cancer_type %in% c("BRCA", "COAD", "KIRP", "STAD", "KIRC", "BLCA")) %>%
  ggplot(aes(x = cancer_type, y = frac_accessible)) +
  geom_violin(aes(fill = cancer_type), alpha = 0.5, trim = FALSE,
              draw_quantiles = c(0.25, 0.5, 0.75)) +
  geom_jitter(width = 0.12, size = 1.2, alpha = 0.7) +
  theme_minimal() +
  labs(y = "Proportion of PKR loci accessible (p < 0.05)", x = NULL) +
  theme(legend.position = "none")

ggsave("pkr/pkr_accessible_proportions_violin_plot.png", 
       plot = pkr_accessibility_violin_plot,
       width = 7,
       height = 5,
       units = "in", 
       dpi = 300)

#### Engager grouops by quartile cut ####
pkr_engager_groups <- pkr_accessibility_pvals_long %>%
  mutate(accessible = pval < 0.05) %>%
  group_by(sample, cancer_type) %>%
  summarise(frac_accessible = mean(accessible), .groups = "drop") %>%
  filter(cancer_type %in% c("BRCA","COAD","KIRP","STAD","KIRC","BLCA")) %>%
  group_by(cancer_type) %>%
  mutate(
    lower_cut = quantile(frac_accessible, 0.25),
    upper_cut = quantile(frac_accessible, 0.75),
    group = case_when(
      frac_accessible <= lower_cut ~ "Non_Engager",
      frac_accessible >= upper_cut ~ "Engager",
      TRUE ~ "Middle"
    )
  ) %>%
  ungroup()

pkr_engager_group_counts <- pkr_engager_groups %>%
  filter(group %in% c("Engager","Non_Engager")) %>%
  count(cancer_type, group) %>%
  pivot_wider(names_from = group, values_from = n, values_fill = 0)

# A tibble: 6 × 3
#   cancer_type Engager Non_Engager
#   <chr>         <int>       <int>
# 1 BLCA              2           2
# 2 BRCA             16          15
# 3 COAD              9           9
# 4 KIRC              3           3
# 5 KIRP              9           9
# 6 STAD              4           4

pkr_engager_group_counts %>%
  gt() %>%
  tab_header(
    title = "PKR engager groups by cancer type",
    subtitle = "Upper and lower quartiles of proportion of PKR loci accessible (p < 0.05)"
  ) %>%
  cols_label(cancer_type = "Cancer Type") %>%
  tab_options(column_labels.background.color = "whitesmoke") %>%
  gtsave("pkr/pkr_quartile_grouping/cluster_definitions/pkr_engager_group_counts.html")

pkr_sample_groups <- pkr_engager_groups %>%
  filter(group %in% c("Engager", "Non_Engager")) %>%
  select(sample, cancer_type, frac_accessible, grouping_label = group)

write_tsv(pkr_sample_groups, "pkr/pkr_quartile_grouping/cluster_definitions/pkr_sample_groups_quartile_grouping.tsv")


```

# PASTED FROM GMM GROUPING SCRIPT JUST USE AS REFERENCE 
```{r}










# A tibble: 18 × 3
#    cancer_type cluster     n
#    <chr>         <dbl> <int>
#  1 BLCA              1     1
#  2 BLCA              2     3
#  3 BLCA              3     4
#  4 BRCA              1    23
#  5 BRCA              2    27
#  6 BRCA              3     7
#  7 COAD              1     8
#  8 COAD              2    27
#  9 COAD              3     1
# 10 KIRC              1     5
# 11 KIRC              2     5
# 12 KIRC              3     1
# 13 KIRP              1    16
# 14 KIRP              2    16
# 15 KIRP              3     1
# 16 STAD              1     7
# 17 STAD              2     8
# 18 STAD              3     1

sample_gmm_scores <- engager_loci_pvals_candidate_cancers %>% 
  left_join(gmm_results_by_cancer %>%  select(sample, cluster),
            by = "sample"
            ) %>% 
  rowwise() %>% 
  mutate(sample_mean_accessibility = mean(c_across(starts_with("chr")))) %>% 
  ungroup()

# Rank clusters within each cancer type by mean accessibility
cluster_definitions <- sample_gmm_scores %>% 
  group_by(cancer_type, cluster) %>% 
  summarise(
    n_samples = n(),
    avg_cluster_accessibility = mean(sample_mean_accessibility),
    .groups = "drop"
  ) %>% 
  group_by(cancer_type) %>% 
  arrange(cancer_type, desc(avg_cluster_accessibility)) %>% 
  mutate(
    grouping_label = case_when(
      row_number() == 1 ~ "Engager",
      row_number() == 2 ~ "Intermediate",
      row_number() == 3 ~ "Non_Engager"
    )
  ) %>% 
  ungroup()

#write_tsv(cluster_definitions, "~/gmm_grouping/cluster_definitions/gmm_cluster_definitions.tsv")

# Make the gt table figure
cluster_definitions %>%
  mutate(grouping_label = factor(grouping_label,
                                 levels = c("Engager", "Intermediate", "Non_Engager"))) %>%
  arrange(cancer_type, grouping_label) %>%
  gt(groupname_col = "cancer_type") %>%
  tab_header(
    title = "GMM Cluster Assignments by Cancer Type",
    subtitle = "Clusters ranked by mean accessibility (-log10 p-value) across 132 RIG-I engager loci"
  ) %>%
  cols_label(
    cluster = "Cluster",
    n_samples = "N Samples",
    avg_cluster_accessibility = "Mean Accessibility",
    grouping_label = "Group"
  ) %>%
  fmt_number(columns = avg_cluster_accessibility, decimals = 1) %>%
  tab_options(row_group.background.color = "whitesmoke") %>% 
  gtsave(.,"~/gmm_grouping/cluster_definitions/gmm_cluster_definitions_table.html")


final_cluster_groups <- sample_gmm_scores %>%
  left_join(
    cluster_definitions %>% dplyr::select(cancer_type, cluster, grouping_label),
    by = c("cancer_type", "cluster")
  )

#write_tsv(final_cluster_groups, "~/gmm_grouping/gmm_sample_groups.tsv")

```

