---
title: "methylation_analysis_v1.Rmd"
output: html_document
date: "2026-04-01"
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(results = "hide")


```


```{r packages}
if(!require("BiocManager", quietly = T))
  install.packages("BiocManager")

# Install the necessary bioinformatics packages
# TCGAbiolinks: for data access
# GSVA: for the 94-gene signature scoring
# SummarizedExperiment: the standard data container for TCGA data
BiocManager::install(c("TCGAbiolinks", "GSVA", "SummarizedExperiment"))
BiocManager::install("IlluminaHumanMethylation450kanno.ilmn12.hg19")
BiocManager::install("sesameData")
BiocManager::install("sesame")
BiocManager::install("sesame")
BiocManager::install("edgeR")
BiocManager::install("org.Hs.eg.db")

install.packages("nloptr")
install.packages("ggpubr")
install.packages("cowplot")
install.packages("gt")
install.packages("minifi")

```


```{r load packages}
library(BiocManager)
library(TCGAbiolinks)
library(GSVA)
library(SummarizedExperiment)
library(tidyverse)
library(nloptr)
library(ggpubr)
library(cowplot)
library(gt)
library(webshot2)
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
library(sesameData)
library(sesame)
library(edgeR)
library(org.Hs.eg.db)
library(AnnotationDbi)
```

# Check which methylation data are available for candidate cancer types
```{r}

candidates <- c("TCGA-HNSC", "TCGA-STAD", "TCGA-KIRC", "TCGA-BLCA", 
                "TCGA-KIRP", "TCGA-BRCA", "TCGA-COAD")

# Query methylation data availability for one cancer first
query_met <- GDCquery(
  project = "TCGA-HNSC",
  data.category = "DNA Methylation",
  data.type = "Methylation Beta Value",
  platform = "Illumina Human Methylation 450"
)

getResults(query_met) %>% 
  count(sample_type)

#           sample_type   n
# 1          Metastatic   2
# 2       Primary Tumor 528
# 3 Solid Tissue Normal  50


```

# Check clinical data availability
```{r}

# Get clnicial data
clinical_hnsc <- GDCquery_clinic("TCGA-HNSC", type = "clinical")

clinical_hnsc <- GDCquery_clinic("TCGA-HNSC", type = "BCR XML")

query_clin <- GDCquery(
  project = "TCGA-HNSC",
  data.category = "Clinical",
  data.type = "Clinical Supplement",
  data.format = "BCR XML"
)

GDCdownload(query_clin)
clinical_hnsc <- GDCprepare_clinic(query_clin, clinical.info = "patient")

# What outcome variables are available?
colnames(clinical_hnsc)

# Check for specific survival columns 
clinical_hnsc %>% 
  select(any_of(c("vital_status", 
                  "days_to_death",
                  "days_to_last_followup",
                  "tumor_stage",
                  "ajcc_pathologic_stage"))) %>% 
  View()

clinical_hnsc %>% 
  select(vital_status, 
         days_to_death,
         days_to_last_followup,
         stage_event_pathologic_stage) %>% 
  summarise(
    n_total = n(),
    n_vital_status = sum(!is.na(vital_status)),
    n_days_to_death = sum(!is.na(days_to_death)),
    n_days_to_last_followup = sum(!is.na(days_to_last_followup))
  )

  # n_total n_vital_status n_days_to_death n_days_to_last_followup
  #    528            160             170                     314

# The number of patients with vital_status is only 160 ~30%
# need to define who is dead vs who is censored

# Check how many patients have either days_to_death OR days_to_last_followup but are missing vital_status
clinical_hnsc %>%
  count(vital_status, is.na(days_to_death))


```

# Data Availability table across all candidate cancer types 
Checking for methylation, RNA-seq (GSEA, IFN signature score analsyes), DNA-seq (for immunotherapy-related biomarker correlations), and survival data
```{r}
check_data_availability <- function(project_id) {
  
  message("Checking: ", project_id)
  
  # Methylation
  met <- tryCatch({
    q <- GDCquery(project = project_id,
                  data.category = "DNA Methylation",
                  data.type = "Methylation Beta Value",
                  platform = "Illumina Human Methylation 450")
    res <- getResults(q)
    tibble(
      n_met_tumor = sum(res$sample_type == "Primary Tumor"),
      n_met_normal = sum(res$sample_type == "Solid Tissue Normal")
    )
  }, error = function(e) tibble(n_met_tumor = NA, n_met_normal = NA))
  
  # RNA-seq
  rna <- tryCatch({
    q <- GDCquery(project = project_id,
                  data.category = "Transcriptome Profiling",
                  data.type = "Gene Expression Quantification",
                  workflow.type = "STAR - Counts",
                  sample.type = c("Primary Tumor", "Solid Tissue Normal"))
    res <- getResults(q)
    tibble(
      n_rna_tumor = sum(res$sample_type == "Primary Tumor"),
      n_rna_normal = sum(res$sample_type == "Solid Tissue Normal")
    )
  }, error = function(e) tibble(n_rna_tumor = NA, n_rna_normal = NA))
  
  # DNA-seq (WXS for somatic mutations/TMB/KRAS)
  dna <- tryCatch({
  q <- GDCquery(project = project_id,
                data.category = "Simple Nucleotide Variation",
                data.type = "Masked Somatic Mutation")
  res <- getResults(q)
  tibble(n_dna_tumor = nrow(res))
}, error = function(e) tibble(n_dna_tumor = NA))
  # Clinical
  clin <- tryCatch({
    q <- GDCquery(project = project_id,
                  data.category = "Clinical",
                  data.type = "Clinical Supplement",
                  data.format = "BCR XML")
    res <- getResults(q)
    tibble(n_clinical = nrow(res))
  }, error = function(e) tibble(n_clinical = NA))
  
  bind_cols(tibble(project = project_id), met, rna, dna, clin)
}

# Run across candidates
candidates <- c("TCGA-HNSC", "TCGA-STAD", "TCGA-KIRC", 
                "TCGA-BLCA", "TCGA-KIRP", "TCGA-BRCA", "TCGA-COAD")

availability_table <- map_dfr(candidates, check_data_availability)

print(availability_table)
```

## make table
```{r}

availability_table %>%
  gt() %>%
  tab_header(
    title = "TCGA Data Availability for Candidate Cancer Types"
  ) %>%
  cols_label(
    project = "Cancer Type",
    n_met_tumor = "Tumor",
    n_met_normal = "Normal",
    n_rna_tumor = "Tumor",
    n_rna_normal = "Normal",
    n_dna_tumor = "Tumor",
    n_clinical = "Cases"
  ) %>%
  tab_spanner(label = "Methylation", columns = c(n_met_tumor, n_met_normal)) %>%
  tab_spanner(label = "RNA-seq", columns = c(n_rna_tumor, n_rna_normal)) %>%
  tab_spanner(label = "DNA-seq", columns = n_dna_tumor) %>%
  tab_spanner(label = "Clinical", columns = n_clinical) %>%
  tab_style(
    style = cell_fill(color = "#f0f0f0"),
    locations = cells_body(rows = seq(1, nrow(availability_table), 2))
  ) %>%
  opt_table_font(font = "Arial")
```


# Checking things
```{r}
q <- GDCquery(
  project = "TCGA-HNSC",
  data.category = "Simple Nucleotide Variation",
  data.type = "Masked Somatic Mutation"
)
```

# Can make new script here!!!Look for overlap of Pol III transcribed loci that are RIG-I "engagers" and "non-engagers"
## Read in coordinate bed file
```{r}
pol3_loci <- read.table("~/PolIII_ON_OFF_map_sample_TCGA_5ppp.bed",
                        header = TRUE,
                        sep = "\t",
                        stringsAsFactors = FALSE)

View(pol3_loci)
dim(pol3_loci)
# [1] 9663  340

# How many RIG-I recognized loci do we have? ####
table(pol3_loci$is_5ppp)
#    0    1 
# 9531  132 

# What RNA types are in the engager set ####
pol3_loci %>% 
  filter(is_5ppp ==1) %>% 
  count(V8) %>% 
  arrange(desc(n))

#             V8  n
# 1         tRNA 80
# 2   snaR A RNA 14
# 3       u6 RNA 14
# 4 snaR B-I RNA 10
# 5      5S rRNA  7
# 6        Y RNA  3
# 7          SRP  2
# 8        Vault  2


# What chromosomes are they on ####
pol3_loci %>%
  filter(is_5ppp == 1) %>%
  count(V1) %>%
  arrange(desc(n))
#       V1  n
# 1   chr6 35
# 2  chr19 25
# 3   chr1 15
# 4   chr7 10
# 5  chr16  6
# 6   chr2  5
# 7  chr10  4
# 8  chr14  4
# 9  chr17  4
# 10  chr5  4
# 11 chr12  3
# 12  chr9  3
# 13 chr11  2
# 14 chr13  2
# 15 chr15  2
# 16  chr4  2
# 17  chr8  2
# 18 chr18  1
# 19 chr20  1
# 20  chr3  1
# 21  chrX  1

# Count how many loci are RIG-I engagers by cancer type ####

# get necessary info and convert to long format
pol3_loci_long <- pol3_loci %>% 
  mutate(across(c(V2,V3,ACC.265:UCEC.96), as.character)) %>% 
  select(V1, V2, V3, gene_name, is_5ppp, ACC.265:UCEC.96) %>% 
  pivot_longer(-c(gene_name, is_5ppp, V1, V2, V3),         #Don't want to pivot these columns
               names_to = "sample",
               values_to = "accessibility") %>% #View()  
  mutate(cancer_type = str_extract(sample, "^[A-Z]+"),
         accessibility = as.integer(accessibility)) #%>%  View()
```

## Descriptive summary tables
```{r}
# Table 1: locus level count per cancer type and locus engager status ####
# For each locus, how many samples are accessible, engagers and non-engagers, by cancer type

pol3_loci_long_t1 <-pol3_loci_long %>%
  group_by(cancer_type, V1, V2, V3, is_5ppp) %>%
  summarise(
    n_samples_accessible_at_this_locus = sum(accessibility == 1),
    n_samples_engagers_at_this_locus = sum(accessibility == 1 & is_5ppp == 1),
    n_samples_non_engagers_at_this_locus = sum(accessibility == 1 & is_5ppp == 0),
    n_samples_discrepant_at_this_locus = sum(accessibility == 0 & is_5ppp == 1),
    .groups = "drop"
  )

View(pol3_loci_long_t1)

# Table 2 characterizes the triphosphatome landscape across cancer types ####
# using the lab's BED file (pol3_loci)
# This table characterizes the triphosphatome landscape across cancer types
# using the lab's BED file (pol3_loci)
# - Triphosphatome landscape: how many RIG-I engager (is_5ppp==1) vs 
#   non-engager (is_5ppp==0) POL III loci are accessible across cancer types
# - n_total_samples: number of ATAC-seq samples per cancer type in the lab's dataset
# NOTE: Table 2 filters to accessibility == 1 only
# Rationale: discrepant samples (is_5ppp == 1 but accessibility == 0) are excluded
# from downstream methylation analysis (5ppp discrepancy concept)
# These are loci where RIG-I engagement was determined in HEK293T cells
# but the locus is not accessible in the TCGA tumor sample
# Discrepant counts are tracked in table 1 (n_samples_discrepant_at_this_locus)
# for transparency but not used in downstream analyses

pol3_loci_long_t2 <- pol3_loci_long %>% 
  filter(accessibility == 1) %>% 
  group_by(cancer_type) %>% 
  summarise(
    n_engager_loci = sum(is_5ppp == 1),
    n_non_engager_loci = sum(is_5ppp == 0),
    n_total_samples = n_distinct(sample),
    .groups = "drop"
  ) %>% 
  mutate(pct_engager_loci = n_engager_loci / (n_engager_loci + n_non_engager_loci) * 100)

View(pol3_loci_long_t2)
```

# Check overlap with TCGA data. Are these loci covered in the TCGA methylation arrays?
```{r}

# Get TCGA coordinates from GDCportal pull ####
results_df <- getResults(query_met)
barcodes <- results_df$cases.submitter_id[results_df$sample_type == "Primary Tumor"][1:2]
barcodes

query_met_small <- GDCquery(
  project = "TCGA-HNSC",
  data.category = "DNA Methylation",
  data.type = "Methylation Beta Value",
  platform = "Illumina Human Methylation 450",
  barcode = barcodes
)

GDCdownload(query_met_small)
se_met_small <- GDCprepare(query_met_small)
probe_coords <- as.data.frame(rowRanges(se_met_small))
head(probe_coords)
dim(probe_coords)

# Check overlap with pol III loci ####

pol3_met_gr <- GRanges(
  seqnames = pol3_loci$V1,
  ranges = IRanges(start = as.integer(pol3_loci$V2), 
                   end = as.integer(pol3_loci$V3)),
  gene_name = pol3_loci$gene_name,
  is_5ppp = pol3_loci$is_5ppp
)

# Create GRanges from 450k probes
pol3_gr <- GRanges(
  seqnames = pol3_loci$V1,
  ranges = IRanges(start = as.integer(pol3_loci$V2), 
                   end = as.integer(pol3_loci$V3)),
  gene_name = pol3_loci$gene_name,
  is_5ppp = pol3_loci$is_5ppp
)

# Create GRanges from 450k probes
probe_gr <- GRanges(
  seqnames = probe_coords$seqnames,
  ranges = IRanges(start = probe_coords$start,
                   end = probe_coords$end),
  probe_id = rownames(probe_coords)
)

# Find overlaps 
# 320 overlaps out of 9663 polIII loci -- only 3.3% coverage ugh
overlaps <- findOverlaps(pol3_gr, probe_gr)
overlaps 

# How many unique POL III loci have coverage
n_covered <- length(unique(queryHits(overlaps)))
n_total <- length(pol3_gr)
cat("Covered loci:", n_covered, "out of", n_total, "(", round(n_covered/n_total*100, 1), "%)\n")

# How many are engagers (is_5ppp=1) vs non-engagers
covered_loci <- pol3_loci[unique(queryHits(overlaps)), ]
table(covered_loci$is_5ppp)

# Does TCGA have EPIC (850k) data?
query_epic <- GDCquery(
  project = "TCGA-HNSC",
  data.category = "DNA Methylation",
  data.type = "Methylation Beta Value",
  platform = "Illumina Human Methylation EPIC"
)
getResults(query_epic) %>% nrow()
```

# Pivoting - don't have good enough coverage for unsupervised clustering will assign groups based on trisphosphatome and ATACseq data

```{r}
# Count n_engager_loci per sample
# Engager = is_5ppp==1 AND accessible==1 (excluding discrepant)
n_engager_loci_per_sample <- pol3_loci_long %>%
  filter(accessibility == 1, is_5ppp == 1) %>%
  group_by(sample, cancer_type) %>%
  summarise(
    n_engager_loci = n_distinct(paste(V1, V2, V3)),
    .groups = "drop"
  )

# Plot distribution per cancer type
ggplot(n_engager_loci_per_sample, aes(x = n_engager_loci, fill = cancer_type)) +
  geom_histogram(bins = 30) +
  facet_wrap(~cancer_type, scales = "free_y") +
  theme_minimal() +
  labs(
    title = "Distribution of RIG-I Engager Loci per Sample",
    x = "N engager loci accessible",
    y = "N samples"
  ) +
  theme(legend.position = "none")
```

## Annotate samples in bed file with corresponding TCGA barcodes
```{r}
# Get metafile table to convert sample indices to project ids then get barcodes to retrieve RNAseq sample data

gdc_sample_metadata <- read_csv("~/gdc_sample_sheet_1.txt")
View(gdc_sample_metadata)

# Create matching key from project_id
gdc_sample_metadata <- gdc_sample_metadata %>% 
  mutate(sample = str_replace(project_id, "TCGA-", "") %>% 
           str_replace("-", "."))

# Join to n_engager_loci_per_sample
n_engager_loci_per_sample <- n_engager_loci_per_sample %>% 
  left_join(gdc_sample_metadata %>% select(sample, case_id, sample_id, sample_type),
            by = "sample")

View(n_engager_loci_per_sample)
```

## Classify samples into engagers, non-engagers, and baseline based on triphosphatome and atac-seq
```{r}

n_engager_loci_per_sample %>% 
  filter(cancer_type %in% c("HNSC", "STAD", "KIRC", "BLCA", "KIRP", "BRCA", "COAD")) %>%
  summary()

 #   sample          cancer_type        n_engager_loci    case_id           sample_id         sample_type       
 # Length:167         Length:167         Min.   :17.00   Length:167         Length:167         Length:167        
 # Class :character   Class :character   1st Qu.:33.00   Class :character   Class :character   Class :character  
 # Mode  :character   Mode  :character   Median :38.00   Mode  :character   Mode  :character   Mode  :character  
 #                                       Mean   :40.02                                                           
 #                                       3rd Qu.:46.50                                                           
 #                                       Max.   :69.00     


n_engager_loci_per_sample_candidates <- n_engager_loci_per_sample %>%
  filter(cancer_type %in% c("HNSC", "STAD", "KIRC", "BLCA", "KIRP", "BRCA", "COAD")) %>%
  group_by(cancer_type) %>%
  mutate(
    tertile = ntile(n_engager_loci, 3),
    group = case_when(
      tertile == 3 ~ "engager",
      tertile == 1 ~ "baseline",
      tertile == 2 ~ "non_engager"
    )
  ) %>%
  ungroup()

# Check group counts per cancer type
n_engager_loci_per_sample_candidates %>%
  count(cancer_type, group) %>% View()

# A tibble: 21 × 3
#    cancer_type group           n
#    <chr>       <chr>       <int>
#  1 BLCA        baseline        3
#  2 BLCA        engager         2
#  3 BLCA        non_engager     3
#  4 BRCA        baseline       19
#  5 BRCA        engager        19
#  6 BRCA        non_engager    19
#  7 COAD        baseline       12
#  8 COAD        engager        12
#  9 COAD        non_engager    12
# 10 HNSC        baseline        2
# 11 HNSC        engager         2
# 12 HNSC        non_engager     2
# 13 KIRC        baseline        4
# 14 KIRC        engager         3
# 15 KIRC        non_engager     4
# 16 KIRP        baseline       11
# 17 KIRP        engager        11
# 18 KIRP        non_engager    11
# 19 STAD        baseline        6
# 20 STAD        engager         5
# 21 STAD        non_engager     5


# Table of groups counts per sample for each cancer type ####
n_engager_loci_per_sample_candidates %>%
  count(cancer_type, group) %>%
  pivot_wider(names_from = group, values_from = n) %>%
  select(cancer_type, engager, non_engager, baseline) %>%
  mutate(total = engager + non_engager + baseline) %>%
  gt() %>%
  cols_label(
    cancer_type = "Cancer type",
    engager = "Engager",
    non_engager = "Non-engager",
    baseline = "Baseline",
    total = "Total"
  ) %>%
  tab_header(title = "RIG-I engager group assignments by cancer type") 

```

# DE analysis on candidate cancer types with >=3 samples per group
Note: HNSC has the strongest biological signal but the smallest ATAC-seq sample size in this dataset.
```{r}
# Extract raw counts for DEseq ####

candidates <- c("BRCA", "COAD", "KIRP", "STAD", "KIRC", "BLCA")

counts_list <- list()
meta_list <- list()

prepare_rnaseq <- function(cancer_type, matched_samples) {
  message("Processing ", cancer_type)
  
  query <- GDCquery(
    project = paste0("TCGA-", cancer_type),
    data.category = "Transcriptome Profiling",
    data.type = "Gene Expression Quantification",
    workflow.type = "STAR - Counts",
    sample.type = "Primary Tumor"
  )
  
  se <- GDCprepare(query)
  
  # Filter to ATAC-seq matched samples only
  matched_barcodes <- matched_samples %>%
    filter(cancer_type == .env$cancer_type) %>%
    pull(case_id)
  
  #se_matched <- se[, se$barcode %in% matched_barcodes]
  se_matched <- se[, substr(colData(se)$barcode, 1, 12) %in% substr(matched_barcodes, 1, 12)]
  se_matched <- se_matched[, !duplicated(substr(colData(se_matched)$barcode, 1, 12))]
  
  message("  Matched ", ncol(se_matched), " of ", length(matched_barcodes), " samples")
  
  list(
    counts = assay(se_matched, "unstranded"),
    meta = colData(se_matched) %>%
      as.data.frame() %>%
      select(barcode, sample_type) %>%
      mutate(case_id = substr(barcode, 1, 12)) %>%
      left_join(
        matched_samples %>%
          filter(cancer_type == .env$cancer_type) %>%
          select(case_id, group, n_engager_loci),
        by = "case_id"
      )
  )
}
```

### Try one cancertype fist BRCA
```{r}

brca_rnaseq_data <- prepare_rnaseq("BRCA", n_engager_loci_per_sample_candidates)

dim(brca_rnaseq_data$counts)

# does meta have group assignments?
head(brca_rnaseq_data$meta)

# check group distribution
table(brca_rnaseq_data$meta$group)

# DE analysis for brca #####
# Create DGEList object
dge <- DGEList(
  counts = brca_rnaseq_data$counts,
  group = brca_rnaseq_data$meta$group
)

# Filter lowly expressed genes
keep <- filterByExpr(dge)
dge <- dge[keep, , keep.lib.sizes = FALSE]

# Normalize
dge <- calcNormFactors(dge)

# Design matrix - engager vs baseline
design <- model.matrix(~ 0 + group, data = brca_rnaseq_data$meta)
colnames(design) <- levels(factor(brca_rnaseq_data$meta$group))

# Estimate dispersion
dge <- estimateDisp(dge, design)

# Fit model
fit <- glmQLFit(dge, design)

# Engager vs baseline contrast
con <- makeContrasts(engager - non_engager, levels = design)
qlf <- glmQLFTest(fit, contrast = con)

# Results
de_results_brca_engager_vs_nonengager <- topTags(qlf, n = Inf)$table
head(de_results_brca_engager_vs_nonengager)

# Strip version numbers from Ensembl IDs
de_results_brca_engager_vs_nonengager$ensembl_id <- rownames(de_results_brca_engager_vs_nonengager) %>% 
  str_remove("\\..*")

# Map to gene symbols
de_results_brca_engager_vs_nonengager$gene_name <- mapIds(
  org.Hs.eg.db,
  keys = de_results_brca_engager_vs_nonengager$ensembl_id,
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)

View(de_results_brca_engager_vs_nonengager)

de_results_brca_engager_vs_nonengager %>%
  filter(FDR < 0.01, !is.na(gene_name)) %>%
  nrow()
# 43 genes FDR < 0.01

de_results_brca_engager_vs_nonengager %>%
  filter(FDR < 0.01, !is.na(gene_name)) %>%
  count(direction = ifelse(logFC > 0, "up", "down"))
#   direction  n
# 1      down 22
# 2        up 21

# Sanity check -- are any IFN/innate immune genes in the upregulated set?
de_results_brca_engager_vs_nonengager %>%
  filter(FDR < 0.01, !is.na(gene_name), logFC > 0) %>%
  arrange(FDR) %>%
  pull(gene_name)

# Check in one gene in this upregulated list has any loci that are "engager loci" 5ppp==1 accessibility == 1 that's in rig-i recognized and pol II transcribed ####
de_results_brca_engager_vs_nonengager %>%
  filter(gene_name == "ERVFRD-1") %>%
  dplyr::select(ensembl_id, logFC, FDR)

# Create GRanges for ERVFRD-1
ervfrd1_gr <- rowRanges(se_brca)[rowData(se_brca)$gene_name == "ERVFRD-1"]
ervfrd1_gr

# Create GRanges for engager loci in brca
# Unique engager loci accessible in at least one BRCA sample
brca_engager_loci <- pol3_loci_long %>%
  filter(cancer_type == "BRCA",
         is_5ppp == 1,
         accessibility == 1) %>%
  distinct(V1, V2, V3)

# Create GRanges
engager_gr_brca <- GRanges(
  seqnames = brca_engager_loci$V1,
  ranges = IRanges(
    start = as.integer(brca_engager_loci$V2),
    end = as.integer(brca_engager_loci$V3)
  )
)

# Creat GRanges for upregulated genes
up_genes <- de_results_brca_engager_vs_nonengager %>%
  filter(FDR < 0.01, !is.na(gene_name), logFC > 0) %>%
  pull(gene_name)

up_gr <- rowRanges(se_brca)[rowData(se_brca)$gene_name %in% up_genes]
up_gr

# Check overlap with upregulated DE genes
overlaps_up_engager <- findOverlaps(up_gr, engager_gr_brca)
names(up_gr)[queryHits(overlaps_up_engager)]
engager_gr_brca[subjectHits(overlaps_up_engager)]

# Check overlap with POL III loci
overlaps <- findOverlaps(ervfrd1_gr, pol3_gr)
overlaps

# # Something's wrong with the barcode format. it's returned 0matched samples
# # query and prepare for brca
# query_brca <- GDCquery(
#   project = "TCGA-BRCA",
#   data.category = "Transcriptome Profiling",
#   data.type = "Gene Expression Quantification",
#   workflow.type = "STAR - Counts",
#   sample.type = "Primary Tumor"
# )
# 
# se_brca <- GDCprepare(query_brca)
# head(colData(se_brca)$barcode)

# Check if any upregulated genes have "engager" loci
```

# Need to clean up saving necessary objects to pull into another script where i do the de analysis
```{r}
# Save objects for DE analysis script ####
saveRDS(n_engager_loci_per_sample_candidates, "n_engager_loci_per_sample_candidates.rds")
saveRDS(pol3_loci_long, "pol3_loci_long.rds")
saveRDS(pol3_loci, "pol3_loci.rds")
saveRDS(pol3_gr, "pol3_gr.rds")
```


