---
title: "methylation_analysis_v1.Rmd"
output: html_document
date: "2026-04-01"
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(results = "hide")


```


```{r packages}
if(!require("BiocManager", quietly = T))
  install.packages("BiocManager")

# Install the necessary bioinformatics packages
# TCGAbiolinks: for data access
# GSVA: for the 94-gene signature scoring
# SummarizedExperiment: the standard data container for TCGA data
BiocManager::install(c("TCGAbiolinks", "GSVA", "SummarizedExperiment"))
BiocManager::install("IlluminaHumanMethylation450kanno.ilmn12.hg19")
BiocManager::install("sesameData")
BiocManager::install("sesame")
BiocManager::install("sesame")
BiocManager::install("edgeR")
BiocManager::install("org.Hs.eg.db")

install.packages("nloptr")
install.packages("ggpubr")
install.packages("cowplot")
install.packages("gt")
install.packages("minifi")

```


```{r load packages}
library(BiocManager)
library(TCGAbiolinks)
library(GSVA)
library(SummarizedExperiment)
library(tidyverse)
library(nloptr)
library(ggpubr)
library(cowplot)
library(gt)
library(webshot2)
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
library(sesameData)
library(sesame)
library(edgeR)
library(org.Hs.eg.db)
library(AnnotationDbi)
```

# Check which methylation data are available for candidate cancer types
```{r}

candidates <- c("TCGA-HNSC", "TCGA-STAD", "TCGA-KIRC", "TCGA-BLCA", 
                "TCGA-KIRP", "TCGA-BRCA", "TCGA-COAD")

# Query methylation data availability for one cancer first
query_met <- GDCquery(
  project = "TCGA-HNSC",
  data.category = "DNA Methylation",
  data.type = "Methylation Beta Value",
  platform = "Illumina Human Methylation 450"
)

getResults(query_met) %>% 
  count(sample_type)

#           sample_type   n
# 1          Metastatic   2
# 2       Primary Tumor 528
# 3 Solid Tissue Normal  50


```

# Check clinical data availability
```{r}

# Get clnicial data
clinical_hnsc <- GDCquery_clinic("TCGA-HNSC", type = "clinical")

clinical_hnsc <- GDCquery_clinic("TCGA-HNSC", type = "BCR XML")

query_clin <- GDCquery(
  project = "TCGA-HNSC",
  data.category = "Clinical",
  data.type = "Clinical Supplement",
  data.format = "BCR XML"
)

GDCdownload(query_clin)
clinical_hnsc <- GDCprepare_clinic(query_clin, clinical.info = "patient")

# What outcome variables are available?
colnames(clinical_hnsc)

# Check for specific survival columns 
clinical_hnsc %>% 
  select(any_of(c("vital_status", 
                  "days_to_death",
                  "days_to_last_followup",
                  "tumor_stage",
                  "ajcc_pathologic_stage"))) %>% 
  View()

clinical_hnsc %>% 
  select(vital_status, 
         days_to_death,
         days_to_last_followup,
         stage_event_pathologic_stage) %>% 
  summarise(
    n_total = n(),
    n_vital_status = sum(!is.na(vital_status)),
    n_days_to_death = sum(!is.na(days_to_death)),
    n_days_to_last_followup = sum(!is.na(days_to_last_followup))
  )

  # n_total n_vital_status n_days_to_death n_days_to_last_followup
  #    528            160             170                     314

# The number of patients with vital_status is only 160 ~30%
# need to define who is dead vs who is censored

# Check how many patients have either days_to_death OR days_to_last_followup but are missing vital_status
clinical_hnsc %>%
  count(vital_status, is.na(days_to_death))


```

# Data Availability table across all candidate cancer types 
Checking for methylation, RNA-seq (GSEA, IFN signature score analsyes), DNA-seq (for immunotherapy-related biomarker correlations), and survival data
```{r}
check_data_availability <- function(project_id) {
  
  message("Checking: ", project_id)
  
  # Methylation
  met <- tryCatch({
    q <- GDCquery(project = project_id,
                  data.category = "DNA Methylation",
                  data.type = "Methylation Beta Value",
                  platform = "Illumina Human Methylation 450")
    res <- getResults(q)
    tibble(
      n_met_tumor = sum(res$sample_type == "Primary Tumor"),
      n_met_normal = sum(res$sample_type == "Solid Tissue Normal")
    )
  }, error = function(e) tibble(n_met_tumor = NA, n_met_normal = NA))
  
  # RNA-seq
  rna <- tryCatch({
    q <- GDCquery(project = project_id,
                  data.category = "Transcriptome Profiling",
                  data.type = "Gene Expression Quantification",
                  workflow.type = "STAR - Counts",
                  sample.type = c("Primary Tumor", "Solid Tissue Normal"))
    res <- getResults(q)
    tibble(
      n_rna_tumor = sum(res$sample_type == "Primary Tumor"),
      n_rna_normal = sum(res$sample_type == "Solid Tissue Normal")
    )
  }, error = function(e) tibble(n_rna_tumor = NA, n_rna_normal = NA))
  
  # DNA-seq (WXS for somatic mutations/TMB/KRAS)
  dna <- tryCatch({
  q <- GDCquery(project = project_id,
                data.category = "Simple Nucleotide Variation",
                data.type = "Masked Somatic Mutation")
  res <- getResults(q)
  tibble(n_dna_tumor = nrow(res))
}, error = function(e) tibble(n_dna_tumor = NA))
  # Clinical
  clin <- tryCatch({
    q <- GDCquery(project = project_id,
                  data.category = "Clinical",
                  data.type = "Clinical Supplement",
                  data.format = "BCR XML")
    res <- getResults(q)
    tibble(n_clinical = nrow(res))
  }, error = function(e) tibble(n_clinical = NA))
  
  bind_cols(tibble(project = project_id), met, rna, dna, clin)
}

# Run across candidates
candidates <- c("TCGA-HNSC", "TCGA-STAD", "TCGA-KIRC", 
                "TCGA-BLCA", "TCGA-KIRP", "TCGA-BRCA", "TCGA-COAD")

availability_table <- map_dfr(candidates, check_data_availability)

print(availability_table)
```

## make table
```{r}

availability_table %>%
  gt() %>%
  tab_header(
    title = "TCGA Data Availability for Candidate Cancer Types"
  ) %>%
  cols_label(
    project = "Cancer Type",
    n_met_tumor = "Tumor",
    n_met_normal = "Normal",
    n_rna_tumor = "Tumor",
    n_rna_normal = "Normal",
    n_dna_tumor = "Tumor",
    n_clinical = "Cases"
  ) %>%
  tab_spanner(label = "Methylation", columns = c(n_met_tumor, n_met_normal)) %>%
  tab_spanner(label = "RNA-seq", columns = c(n_rna_tumor, n_rna_normal)) %>%
  tab_spanner(label = "DNA-seq", columns = n_dna_tumor) %>%
  tab_spanner(label = "Clinical", columns = n_clinical) %>%
  tab_style(
    style = cell_fill(color = "#f0f0f0"),
    locations = cells_body(rows = seq(1, nrow(availability_table), 2))
  ) %>%
  opt_table_font(font = "Arial")
```


 Checking things
```{r}
q <- GDCquery(
  project = "TCGA-HNSC",
  data.category = "Simple Nucleotide Variation",
  data.type = "Masked Somatic Mutation"
)
```


# Read in coordinate bed file of poliii loci
```{r}
pol3_loci <- read.table("~/PolIII_ON_OFF_map_sample_TCGA_5ppp.bed",
                        header = TRUE,
                        sep = "\t",
                        stringsAsFactors = FALSE)

View(pol3_loci)
dim(pol3_loci)
# [1] 9663  340

# How many RIG-I recognized loci do we have? ####
table(pol3_loci$is_5ppp)
#    0    1 
# 9531  132 

# What RNA types are in the engager set ####
pol3_loci %>% 
  filter(is_5ppp ==1) %>% 
  count(V8) %>% 
  arrange(desc(n))

#             V8  n
# 1         tRNA 80
# 2   snaR A RNA 14
# 3       u6 RNA 14
# 4 snaR B-I RNA 10
# 5      5S rRNA  7
# 6        Y RNA  3
# 7          SRP  2
# 8        Vault  2


# What chromosomes are they on ####
pol3_loci %>%
  filter(is_5ppp == 1) %>%
  count(V1) %>%
  arrange(desc(n))
#       V1  n
# 1   chr6 35
# 2  chr19 25
# 3   chr1 15
# 4   chr7 10
# 5  chr16  6
# 6   chr2  5
# 7  chr10  4
# 8  chr14  4
# 9  chr17  4
# 10  chr5  4
# 11 chr12  3
# 12  chr9  3
# 13 chr11  2
# 14 chr13  2
# 15 chr15  2
# 16  chr4  2
# 17  chr8  2
# 18 chr18  1
# 19 chr20  1
# 20  chr3  1
# 21  chrX  1

# Count how many loci are RIG-I engagers by cancer type ####

# get necessary info and convert to long format
pol3_loci_long <- pol3_loci %>% 
  mutate(across(c(V2,V3,ACC.265:UCEC.96), as.character)) %>% 
  select(V1, V2, V3, gene_name, is_5ppp, ACC.265:UCEC.96) %>% 
  pivot_longer(-c(gene_name, is_5ppp, V1, V2, V3),         #Don't want to pivot these columns
               names_to = "sample",
               values_to = "accessibility") %>% #View()  
  mutate(cancer_type = str_extract(sample, "^[A-Z]+"),
         accessibility = as.integer(accessibility)) #%>%  View()
```

## Descriptive summary tables
```{r}
# Tables to categorize the different groups of Pol3 loci ####
per_sample_loci_summary_df <- pol3_loci_long %>% 
  group_by(sample) %>% 
  summarise(
    n_engager_loci = sum(accessibility == 1 & is_5ppp==1),
    n_non_engager_loci = sum(accessibility == 1 & is_5ppp==0),
    n_methylated_loci = sum(accessibility == 0 & is_5ppp ==0),
    n_discrepant_loci = sum(accessibility==0 & is_5ppp == 1),
    .groups = "drop"
  ) %>% 
  mutate(pct_engager_loci = n_engager_loci/9663) %>%  # percent of pol3 loci that are both accessible and rigi recognized 
  mutate(pct_accessible_loci_that_are_engager_loci = n_engager_loci/(n_engager_loci + n_non_engager_loci)) #%>% View


# Histogram plot the distribution of pct_accessible_loci_that_are_engager_loci
per_sample_loci_summary_df %>%
  mutate(cancer_type = str_extract(sample, "^[A-Z]+")) %>%
  filter(cancer_type %in% c("BRCA", "COAD", "KIRP", "STAD", "KIRC", "BLCA")) %>%
  ggplot(aes(x = pct_accessible_loci_that_are_engager_loci, fill = cancer_type)) +
  geom_histogram(binwidth = 0.005, color = "black", linewidth = 0.3) +
  facet_wrap(~cancer_type, scales = "free_y") +
  theme_minimal() +
  labs(x = "% accessible POL III loci that are RIG-I engagers", y = "N samples") +
  theme(legend.position = "none")

# Violin plot
per_sample_loci_summary_df %>%
  mutate(cancer_type = str_extract(sample, "^[A-Z]+")) %>%
  filter(cancer_type %in% c("BRCA", "COAD", "KIRP", "STAD", "KIRC", "BLCA")) %>%
  ggplot(aes(x = cancer_type, y = pct_accessible_loci_that_are_engager_loci, fill = cancer_type)) +
  geom_violin() +
  geom_jitter(width = 0.1, size = 1, alpha = 0.5) +
  theme_minimal() +
  labs(y = "% accessible POL III loci that are RIG-I engagers", x = NULL) +
  theme(legend.position = "none")

# Boxplot
per_sample_loci_summary_df %>%
  mutate(cancer_type = str_extract(sample, "^[A-Z]+")) %>%
  filter(cancer_type %in% c("BRCA", "COAD", "KIRP", "STAD", "KIRC", "BLCA")) %>%
  ggplot(aes(x = cancer_type, y = pct_accessible_loci_that_are_engager_loci, fill = cancer_type)) +
  geom_boxplot() +
  geom_jitter(width = 0.1, size = 1, alpha = 0.5) +
  theme_minimal() +
  labs(y = "% accessible POL III loci that are RIG-I engagers", x = NULL) +
  theme(legend.position = "none")

# number of samples in each quartile
per_sample_loci_summary_df %>%
  mutate(cancer_type = str_extract(sample, "^[A-Z]+")) %>%
  filter(cancer_type %in% c("BRCA", "COAD", "KIRP", "STAD", "KIRC", "BLCA")) %>%
  group_by(cancer_type) %>%
  mutate(quartile = ntile(pct_accessible_loci_that_are_engager_loci, 4)) %>%
  count(cancer_type, quartile) %>% View

  ungroup

per_sample_loci_summary_df %>%
  filter(n_engager_loci == 0) %>%
  nrow()


```




# Check overlap with TCGA data. Are these loci covered in the TCGA methylation arrays?
```{r}

# Get TCGA coordinates from GDCportal pull ####
results_df <- getResults(query_met)
barcodes <- results_df$cases.submitter_id[results_df$sample_type == "Primary Tumor"][1:2]
barcodes

query_met_small <- GDCquery(
  project = "TCGA-HNSC",
  data.category = "DNA Methylation",
  data.type = "Methylation Beta Value",
  platform = "Illumina Human Methylation 450",
  barcode = barcodes
)

GDCdownload(query_met_small)
se_met_small <- GDCprepare(query_met_small)
probe_coords <- as.data.frame(rowRanges(se_met_small))
head(probe_coords)
dim(probe_coords)

# Check overlap with pol III loci ####

pol3_met_gr <- GRanges(
  seqnames = pol3_loci$V1,
  ranges = IRanges(start = as.integer(pol3_loci$V2), 
                   end = as.integer(pol3_loci$V3)),
  gene_name = pol3_loci$gene_name,
  is_5ppp = pol3_loci$is_5ppp
)

# Create GRanges from 450k probes
pol3_gr <- GRanges(
  seqnames = pol3_loci$V1,
  ranges = IRanges(start = as.integer(pol3_loci$V2), 
                   end = as.integer(pol3_loci$V3)),
  gene_name = pol3_loci$gene_name,
  is_5ppp = pol3_loci$is_5ppp
)

# Create GRanges from 450k probes
probe_gr <- GRanges(
  seqnames = probe_coords$seqnames,
  ranges = IRanges(start = probe_coords$start,
                   end = probe_coords$end),
  probe_id = rownames(probe_coords)
)

# Find overlaps 
# 320 overlaps out of 9663 polIII loci -- only 3.3% coverage ugh
overlaps <- findOverlaps(pol3_gr, probe_gr)
overlaps 

# How many unique POL III loci have coverage
n_covered <- length(unique(queryHits(overlaps)))
n_total <- length(pol3_gr)
cat("Covered loci:", n_covered, "out of", n_total, "(", round(n_covered/n_total*100, 1), "%)\n")

# How many are engagers (is_5ppp=1) vs non-engagers
covered_loci <- pol3_loci[unique(queryHits(overlaps)), ]
table(covered_loci$is_5ppp)

# Does TCGA have EPIC (850k) data?
query_epic <- GDCquery(
  project = "TCGA-HNSC",
  data.category = "DNA Methylation",
  data.type = "Methylation Beta Value",
  platform = "Illumina Human Methylation EPIC"
)
getResults(query_epic) %>% nrow()
```

# Pivoting - don't have good enough coverage for unsupervised clustering will assign groups based on trisphosphatome and ATACseq data

```{r}
# Count n_engager_loci per sample
# Engager = is_5ppp==1 AND accessible==1 (excluding discrepant)
n_engager_loci_per_sample <- pol3_loci_long %>%
  filter(accessibility == 1, is_5ppp == 1) %>%
  group_by(sample, cancer_type) %>%
  summarise(
    n_engager_loci = n_distinct(paste(V1, V2, V3)),
    .groups = "drop"
  )

# Plot distribution per cancer type
ggplot(n_engager_loci_per_sample, aes(x = n_engager_loci, fill = cancer_type)) +
  geom_histogram(binwidth = 1, color = "black", linewidth=0.3) +
  facet_wrap(~cancer_type, scales = "free_y") +
  theme_minimal() +
  labs(
    title = "Distribution of RIG-I Engager Loci per Sample",
    x = "N engager loci accessible",
    y = "N samples"
  ) +
  theme(legend.position = "none")
```

## Annotate samples in bed file with corresponding TCGA barcodes
```{r}
# Get metafile table to convert sample indices to project ids then get barcodes to retrieve RNAseq sample data

gdc_sample_metadata <- read_csv("~/gdc_sample_sheet_1.txt")
View(gdc_sample_metadata)

# Create matching key from project_id
gdc_sample_metadata <- gdc_sample_metadata %>% 
  mutate(sample = str_replace(project_id, "TCGA-", "") %>% 
           str_replace("-", "."))

# Some samples had more than one BAM file so need to de-dup (e.g. COAD.3, COAD.343, COAD.408 all map to TCGA-AA-A010 because in gdc_sample_metadata that case had 3 BAM files so 3 different atacseq samples mapped to the same tCGA case)


# Join to n_engager_loci_per_sample
n_engager_loci_per_sample_final <- n_engager_loci_per_sample %>% 
  left_join(gdc_sample_metadata %>% dplyr::select(sample, case_id, sample_id, sample_type),
            by = "sample") %>% 
  group_by(case_id) %>% 
  slice_max(n_engager_loci, n=1, with_ties = FALSE) %>% 
  ungroup

View(n_engager_loci_per_sample_final)
```

## Classify samples into engagers, non-engagers, and baseline based on triphosphatome and atac-seq
```{r}
# make DF for type of loci (engager loci, etc) for each sample ####
per_sample_loci_summary_df <- pol3_loci_long %>% 
  group_by(sample) %>% 
  summarise(
    n_engager_loci = sum(accessibility == 1 & is_5ppp==1),
    n_non_engager_loci = sum(accessibility == 1 & is_5ppp==0),
    n_methylated_loci = sum(accessibility == 0 & is_5ppp ==0),
    n_discrepant_loci = sum(accessibility==0 & is_5ppp == 1),
    .groups = "drop"
  ) %>% 
  mutate(pct_engager_loci = n_engager_loci/9663) %>%  # percent of pol3 loci that are both accessible and rigi recognized 
  mutate(pct_accessible_loci_that_are_engager_loci = n_engager_loci/(n_engager_loci + n_non_engager_loci)) #%>% View


# Histogram plot the distribution of pct_accessible_loci_that_are_engager_loci ####
per_sample_loci_summary_df %>%
  mutate(cancer_type = str_extract(sample, "^[A-Z]+")) %>%
  filter(cancer_type %in% c("BRCA", "COAD", "KIRP", "STAD", "KIRC", "BLCA")) %>%
  ggplot(aes(x = pct_accessible_loci_that_are_engager_loci, fill = cancer_type)) +
  geom_histogram(binwidth = 0.005, color = "black", linewidth = 0.3) +
  facet_wrap(~cancer_type, scales = "free_y") +
  theme_minimal() +
  labs(x = "% accessible POL III loci that are RIG-I engagers", y = "N samples") +
  theme(legend.position = "none")

# Violin plot ####
per_sample_loci_summary_df %>%
  mutate(cancer_type = str_extract(sample, "^[A-Z]+")) %>%
  filter(cancer_type %in% c("BRCA", "COAD", "KIRP", "STAD", "KIRC", "BLCA")) %>%
  ggplot(aes(x = cancer_type, y = pct_accessible_loci_that_are_engager_loci, fill = cancer_type)) +
  geom_violin() +
  geom_jitter(width = 0.1, size = 1, alpha = 0.5) +
  theme_minimal() +
  labs(y = "% accessible POL III loci that are RIG-I engagers", x = NULL) +
  theme(legend.position = "none")

# Boxplot ####
per_sample_loci_summary_df %>%
  mutate(cancer_type = str_extract(sample, "^[A-Z]+")) %>%
  filter(cancer_type %in% c("BRCA", "COAD", "KIRP", "STAD", "KIRC", "BLCA")) %>%
  ggplot(aes(x = cancer_type, y = pct_accessible_loci_that_are_engager_loci, fill = cancer_type)) +
  geom_boxplot() +
  geom_jitter(width = 0.1, size = 1, alpha = 0.5) +
  theme_minimal() +
  labs(y = "% accessible POL III loci that are RIG-I engagers", x = NULL) +
  theme(legend.position = "none")

# number of samples in each quartile ####
per_sample_loci_summary_df %>%
  mutate(cancer_type = str_extract(sample, "^[A-Z]+")) %>%
  filter(cancer_type %in% c("BRCA", "COAD", "KIRP", "STAD", "KIRC", "BLCA")) %>%
  group_by(cancer_type) %>%
  mutate(quartile = ntile(pct_accessible_loci_that_are_engager_loci, 4)) %>%
  count(cancer_type, quartile) %>% View


per_sample_loci_summary_df %>%
  filter(n_engager_loci == 0) %>%
  nrow()


# Joing to gdc_sample_metadata to get TCGA barcodes to retrieve RNA

candidate_sample_groups <- per_sample_loci_summary_df %>%
  mutate(cancer_type = str_extract(sample, "^[A-Z]+")) %>%
  filter(cancer_type %in% c("BRCA", "COAD", "KIRP", "STAD", "KIRC", "BLCA")) %>%
  left_join(gdc_sample_metadata %>% dplyr::select(sample, case_id, sample_id, sample_type),
            by = "sample") %>% 
  group_by(case_id) %>% 
  slice_sample(n=1) %>% 
  ungroup() %>% 
  group_by(cancer_type) %>%
  mutate(quartile = ntile(pct_accessible_loci_that_are_engager_loci, 4),
         engager_group = case_when(
           quartile == 4 ~ "engager",
           quartile == 1 ~ "non_engager",
           TRUE ~ NA_character_
         )) %>%
  ungroup() %>% 
  filter(!is.na(engager_group)) #%>% View()
  

# Check if there are more than one sample matching to the same case_id
candidate_sample_groups %>% 
  group_by(cancer_type) %>% 
  count(case_id) %>% 
  filter(n>1)

  



```

# Save objects for DE analysis
```{r}

#saveRDS(candidate_sample_groups, "candidate_sample_groups.rds")
#saveRDS(pol3_loci_long, "pol3_loci_long.rds")
#saveRDS(pol3_gr, "pol3_gr.rds")

```

