---
title: "type1ifn_enrichment.Rmd"
output: html_document
date: "2026-01-23"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(results = "hide")

print("hi")
```


```{r packages}
if(!require("BiocManager", quietly = T))
  install.packages("BiocManager")

# Install the necessary bioinformatics packages
# TCGAbiolinks: for data access
# GSVA: for the 94-gene signature scoring
# SummarizedExperiment: the standard data container for TCGA data
BiocManager::install(c("TCGAbiolinks", "GSVA", "SummarizedExperiment"))

install.packages("nloptr")
install.packages("ggpubr")
install.packages("cowplot")
```


```{r load packages}
library(BiocManager)
library(TCGAbiolinks)
library(GSVA)
library(SummarizedExperiment)
library(tidyverse)
library(nloptr)
library(ggpubr)
library(cowplot)
```

# Type 1 IFN signaling pathway genes identified using MSigDB 
specifically "Hallmark Interferon Alpha Response"
```{r}

ifn_genes <- c("ADAR", "B2M", "BATF2", "BST2", "C1S", "CASP1", "CASP8", 
              "CCRL2", "CD47", "CD74", "CMPK2", "CNP", "CSF1", "CXCL10", 
              "CXCL11", "DDX60", "DHX58", "EIF2AK2", "ELF1", "EPSTI1", 
              "FAM125A", "FAM46A", "FTSJD2", "GBP2", "GBP4", "GMPR", 
              "HERC6", "HLA-C", "IFI27", "IFI30", "IFI35", "IFI44", 
              "IFI44L", "IFIH1", "IFIT2", "IFIT3", "IFITM1", "IFITM2", 
              "IFITM3", "IL15", "IL4R", "IL7", "IRF1", "IRF2", "IRF7", 
              "IRF9", "ISG15", "ISG20", "LAMP3", "LAP3", "LGALS3BP", 
              "LPAR6", "LY6E", "MOV10", "MX1", "NCOA7", "NMI", "NUB1", 
              "OAS1", "OASL", "OGFR", "PARP12", "PARP14", "PARP9", 
              "PLSCR1", "PNPT1", "PRIC285", "PROCR", "PSMA3", "PSMB8", 
              "PSMB9", "PSME1", "PSME2", "RIPK2", "RNF31", "RSAD2", 
              "RTP4", "SAMD9", "SAMD9L", "SELL", "SLC25A28", "SP110", 
              "STAT2", "TAP1", "TDRD7", "TMEM140", "TRAFD1", "TRIM14", 
              "TRIM21", "TRIM25", "TRIM26", "TRIM5", "TXNIP", "UBA7", 
              "UBE2L6", "USP18", "WARS")

ifn_genes_list <- list(ifn_signature = ifn_genes)

```

# Analyze by cancer type 
```{r analyze_cancer_type function}
all_projects <- getGDCprojects() %>% 
  filter(str_detect(project_id, "^TCGA")) %>% 
  pull(project_id) 

### output dir ####
if(!dir.exists("inf_plots")) dir.create("ifn_plots")
res_df <- tibble() #Wilcox results here

plot_list <- list()

analyze_cancer_type <- function(project_id, gene_list){
  
  message("\n>>> Starting analysis for: ", project_id)
  
  # GDC query
  query <- GDCquery(
    #project = "TCGA-COAD",
    project = project_id,
    data.category = "Transcriptome Profiling",
    data.type = "Gene Expression Quantification",
    workflow.type = "STAR - Counts",
    sample.type = c("Primary Tumor", "Solid Tissue Normal")
  )
  
  GDCdownload(query = query)
  se <- GDCprepare(query)
  
  # Get metadata ####
  meta <- colData(se) %>% 
    as.data.frame() %>% 
    #as_tibble(rownames = "barcode") %>% 
    dplyr::select(barcode, sample_type) %>% 
    mutate(sample_type = if_else(sample_type == "Primary Tumor", "Tumor", "Normal"))
  
  # Subset to signature genes ####
  keep_rows <- as.data.frame(rowData(se))$gene_name %in% gene_list
  se_subset <- se[keep_rows,]
  
  # # Prepare expression data #
  # 
  # gene_meta <- as.data.frame(rowData(se))
  # keep_rows <- gene_meta$gene_name %in% ifn_genes
  # 
  # se_ifn_subset <- se[keep_rows,]
  
  # Extract tpm matrix, replace ensembl ids with hgnc symbols as rownames
  expr_mat <- assay(se_subset,"tpm_unstrand") 
  rownames(expr_mat) <- rowData(se_subset)$gene_name
  
  # log transform for z-scoring step
  expr_mat_log <- log2(expr_mat + 1)
  
  
  # 5 genes dropped out -- which ones?
  # print(setdiff(ifn_genes, rownames(expr_mat)))
  #  [1] "FAM125A" "FAM46A"  "FTSJD2"  "PRIC285" "WARS"
  
  # log 2 fold change ####
  # Calcd from raw TPM (before z-scoring)
  # mean log2(tumor/normal) across all signature genes
  # positive = signature genes up in tumor vs normal on average
  # comparabel across cancer types since it's in absolute tpm
  log2fc_result <- expr_mat %>%
    as.data.frame() %>%
    rownames_to_column("gene_name") %>%
    pivot_longer(-gene_name, names_to = "barcode", values_to = "tpm") %>%
    inner_join(meta, by = "barcode") %>%
    group_by(gene_name, sample_type) %>%
    summarise(median_tpm = median(tpm), .groups = "drop") %>%
    pivot_wider(names_from = sample_type, values_from = median_tpm) %>%
    # +1 to avoid log of zero
    mutate(log2fc = log2((Tumor + 1) / (Normal + 1))) %>%
    summarise(mean_log2fc = mean(log2fc, na.rm = TRUE))
  
  # Z-score signature ####
  # For each gene: z = (x - mean) / sd across all samples in this cohort
  # Puts every gene on the same scale regardless of baseline abundance
  # Signature score per sample = mean of all gene z-scores for that sample
  manual_score_df <- expr_mat_log %>%
    as.data.frame() %>%
    rownames_to_column("gene_name") %>%
    pivot_longer(-gene_name, names_to = "barcode", values_to = "log_tpm") %>%
    group_by(gene_name) %>%
    mutate(z_score = (log_tpm - mean(log_tpm)) / sd(log_tpm)) %>%
    ungroup() %>%
    group_by(barcode) %>%
    summarise(signature_score = mean(z_score, na.rm = TRUE)) %>%
    inner_join(meta, by = "barcode")
  
  # Statistics ####
  # Wilcoxon rank-sum: non-parametric, appropriate for skewed score distributions
  # Delta = median tumor - median normal (positive = enriched in tumor)
  stats_result <- manual_score_df %>%
    summarise(
      project       = project_id,
      p_val         = wilcox.test(signature_score ~ sample_type, data = .)$p.value,
      median_tumor  = median(signature_score[sample_type == "Tumor"]),
      median_normal = median(signature_score[sample_type == "Normal"]),
      delta         = median_tumor - median_normal,
      mean_log2fc   = log2fc_result$mean_log2fc,
      n_tumor       = sum(sample_type == "Tumor"),
      n_normal      = sum(sample_type == "Normal")
    ) %>%
    mutate(significance = case_when(
      p_val < 0.001 ~ "***",
      p_val < 0.01  ~ "**",
      p_val < 0.05  ~ "*",
      TRUE          ~ "NS"
    ))
  
  # Plot ####
  
  p <- ggplot(manual_score_df, aes(x = sample_type, y = signature_score, fill = sample_type)) +
    geom_violin(alpha = 0.4) +
    geom_boxplot(width = 0.2, outlier.shape = NA) +
    # runs wilcoxon internally and places p-value label automatically
    stat_compare_means(method = "wilcox.test", label = "p.format") +
    theme_minimal() +
    labs(
      title    = project_id,
      subtitle = "Mean Z-score of IFN gene signature",
      caption = paste0("log2FC = ", round(log2fc_result$mean_log2fc, 2)),
      y        = "Signature Score (Z-units)",
      x        = NULL
    ) +
    theme(legend.position = "none")
  
  # Save to global plot_list for violin grid plot later
  plot_list[[project_id]] <<- p
  
  dir.create("IFN_Plots", showWarnings = FALSE)
  ggsave(
    filename = paste0("IFN_Plots/", project_id, "_IFN_signature.png"),
    plot = p, width = 4, height = 5, dpi = 300
  )
  
  # Save scores
  assign(paste0(project_id, "_scores"), manual_score_df, envir = .GlobalEnv)
  
  # Memory cleanup ####
  # Without this, looping through many cancer types will accumulate large
  # SE objects in RAM and eventually crash the session
  rm(se, se_subset, expr_mat, expr_mat_log, manual_score_df)
  gc()
  
  # Save incrementally in case of crash
  write_csv(
    stats_result,
    "IFN_pan_cancer_results.csv",
    append = file.exists("IFN_pan_cancer_results.csv")
  )
  
  return(stats_result)
}

```

# Run the function across cancer types
```{r}
options(timeout = 600)
# Start with a small subset to verify end-to-end before running all 22
test_cancers <- paste0("TCGA-", c("ESCA", "ACC", "UCEC", "LUSC", "SKCM", "BLCA", 
                       "CHOL", "LIHC", "KIRC", "LGG", "KIRP", "PRAD", "PCPG", 
                       "CESC", "THCA", "MESO", "GBM", "STAD", "COAD", "HNSC", "LUAD", "BRCA"
                       ))



# map_dfr applies the function to each cancer type and automatically binds
# the returned stats rows into a single dataframe
master_table <- map_dfr(test_cancers, function(proj) {
  tryCatch(
    analyze_cancer_type(proj, ifn_genes),
    error = function(e) {
      message("!! FAILED: ", proj, " | ", e$message)
      return(NULL)
    }
  )
}) %>% arrange(desc(delta)) # Rank by delta — cancers where IFN signature is most enriched in tumor vs normal




#analyze_cancer_type("TCGA-BRCA", ifn_genes)
```

# QC the results
```{r}
# Check for dupes
master_table %>%  count(project) %>% filter(n>1) # No duplicates

# Check which cancer types are missing presumably due to few samples/metadata?
setdiff(test_cancers, master_table$project)
# [1] "TCGA-ACC"  "TCGA-LGG"  "TCGA-MESO"

# Check plot_lit
length(plot_list)
names(plot_list)

```

# Variance
```{r}
# Calculate variance ratio for each cancer type
variance_df <- map_dfr(names(plot_list), function(proj) {
  
  # Retrieve the saved score dataframe
  scores <- get(paste0(proj, "_scores"))
  
  scores %>%
    group_by(sample_type) %>%
    summarise(var = var(signature_score), .groups = "drop") %>%
    pivot_wider(names_from = sample_type, values_from = var) %>%
    mutate(
      project = proj,
      var_ratio = Tumor / Normal  # >1 means tumor more heterogeneous
    ) %>%
    select(project, var_tumor = Tumor, var_normal = Normal, var_ratio)
})

# Join with master table
master_table <- master_table %>%
  left_join(variance_df, by = "project")

# Get variance ratio t/n
master_table <- master_table %>%
  mutate(var_ratio = var_tumor / var_normal)

# write final master table with var ratios
master_table %>%
  arrange(desc(mean_log2fc)) %>%
  write_csv("IFN_pan_cancer_results_final.csv")

master_table %>%
  arrange(desc(mean_log2fc)) %>%
  select(project, p_val, delta, mean_log2fc, var_ratio, n_tumor, n_normal, significance) %>%
  print(n = 25)
```


# Violin grid plot ranked by log2fc for all cancer types run
```{r}
# Reorder everything based on log2fc ranking
ordered_log2fc <- master_table %>% 
  arrange(desc(mean_log2fc)) %>% 
  pull(project)

ordered_plots <- plot_list[ordered_log2fc]

# Try to make more visually easy to read
# Strip down plots for grid display
# add variance ratio

# Get var_ratio lookup from master table
var_lookup <- master_table %>% 
  select(project, var_ratio) %>% 
  deframe()  # converts to named vector for easy lookup

stripped_plots <- map(names(ordered_plots), function(proj) {
  p <- ordered_plots[[proj]]
  vr <- round(var_lookup[proj], 2)
  
  p + labs(
    subtitle = NULL,
    y        = NULL,
    x        = NULL,
    caption  = paste0("log2FC = ", round(master_table$mean_log2fc[master_table$project == proj], 2),
                      " | var ratio = ", vr)
  ) +
  theme(
    axis.text.y  = element_text(size = 12, face = "bold"),
    axis.text.x  = element_text(size = 12, face = "bold"),
    plot.title   = element_text(size = 16, face = "bold"),
    plot.caption = element_text(size = 14, face = "bold"),
    plot.margin = margin(t = 5, r = 5, b = 20, l = 5)  # increase bottom margin
  )
})

violin_grid_stripped <- plot_grid(
  plotlist = stripped_plots,
  #ncol = 4,
  ncol = 5,
  align = "hv"
  )



# Save
ggsave(
  filename = "IFN_violin_grid_clean.png",
  plot = violin_grid_stripped,
  width = 24,
  height = ceiling(length(ordered_plots) / 5) * 5,
  dpi = 300
)
```

## Filter to types of interest by satisfying criteria then plot filtered version of violin grid
```{r}
candidates <- master_table %>%
  filter(
    mean_log2fc > 0,
    delta > 0,
    median_tumor > median_normal,
    p_val < 0.05,
    n_normal >= 10,
    var_ratio > 1
  ) %>%
  pull(project)

# Also add flagged ones separately
flagged <- master_table %>%
  filter(
    mean_log2fc > 0,
    delta > 0,
    n_normal >= 10,
    p_val >= 0.05  # NS but biologically motivated
  ) %>%
  pull(project)

# Combine candidates and flagged for the grid
to_plot <- union(candidates, flagged)

# Filter ordered_plots maintaining log2FC order
candidate_plots <- ordered_plots[names(ordered_plots) %in% to_plot]

#### plot and save the filtered plots ####
stripped_plots_candidates <- map(names(candidate_plots), function(proj) {
  p <- candidate_plots[[proj]]
  vr <- round(var_lookup[proj], 2)
  
  p + labs(
    subtitle = NULL,
    y        = NULL,
    x        = NULL,
    caption  = paste0("log2FC = ", round(master_table$mean_log2fc[master_table$project == proj], 2),
                      " | var ratio = ", vr)
  ) +
  theme(
    axis.text.y  = element_text(size = 12, face = "bold"),
    axis.text.x  = element_text(size = 12, face = "bold"),
    plot.title   = element_text(size = 16, face = "bold"),
    plot.caption = element_text(size = 14, face = "bold"),
    plot.margin = margin(t = 5, r = 5, b = 20, l = 5)  # increase bottom margin
  )
})

violin_grid_stripped_candidates <- plot_grid(
  plotlist = stripped_plots_candidates,
  #ncol = 4,
  ncol = 5,
  align = "hv"
  )

# Save
ggsave(
  filename = "IFN_violin_grid_clean_candidates.png",
  plot = violin_grid_stripped_candidates,
  width = 24,
  height = ceiling(length(stripped_plots_candidates) / 5) * 5,
  dpi = 300
)

```

# snaR-A stuff - let's start with %snar-a ON vs IFN signaling delta and log2fc
```{r}
snara_pct <- tibble(
  project = paste0("TCGA-", c("TGCT", "COAD", "HNSC", "ESCA", "ACC", "UCEC", 
                               "LUSC", "STAD", "SKCM", "LUAD", "BLCA", "CHOL",
                               "BRCA", "LIHC", "KIRC", "LGG", "KIRP", "PRAD",
                               "PCPG", "CESC", "THCA", "MESO", "GBM")),
  snara_pct_on = c(98, 73, 69, 51, 45, 44, 44, 37, 35, 29, 26, 25, 
                   17, 17, 16, 15, 12, 12, 8, 7, 7, 6, 3)
)

#define status for each cancer type
candidate_status <- master_table %>% 
  mutate(status = case_when(
    project %in% candidates ~ "Candidate", 
    project %in% flagged ~ "Flagged",
    TRUE                 ~ "Excluded"
  ))

# Join with snar-a data
bubble_df <- candidate_status %>% 
  inner_join(snara_pct, by = "project") %>% 
  mutate(project_label = gsub("TCGA-", "", project))
```


## snarA bubble plot
```{r}

# Plot
bubble_plot <- ggplot(bubble_df, aes(x = snara_pct_on, y = mean_log2fc,
                     size = n_normal,
                     color = status,
                     label = project_label)) +
  geom_point(alpha = 0.7) +
  geom_text(size = 3, vjust = -1, color = "black") +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  scale_color_manual(values = c("Candidate" = "#d73027",
                                "Flagged"   = "#fc8d59",
                                "Excluded"  = "grey60")) +
  scale_size_continuous(range = c(3, 12), breaks = c(5,20,50,100)) +
  theme_minimal() +
  labs(
    x     = "snaR-A % ON (Zhou et al 2025)",
    y     = "Mean log2FC",
    title = "IFN Signature vs snaR-A Accessibility",
    subtitle = "Point size = n normal samples",
    color = "Status",
    size  = "n normal samples"
  ) +
  theme(
    plot.title    = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10),
    axis.title    = element_text(size = 12, face = "bold"),
    axis.text     = element_text(size = 10),
    legend.title  = element_text(size = 10, face = "bold")
  )

ggsave("IFN_snara_bubble.png", plot = bubble_plot, width = 10, height = 7, dpi = 300)


```

